### 2017-07-09 ms/ms.Rnw
put directory on github

% Notes from call with Rob
% van Tienderen - sensitivity approach better for evolution
% send list of species to rob

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury
% leaf size

% OPTIONAL
% Synthesis model
% total stomatal density versus growth form (as supplement???)
% used ordered multinomial rather than lm for lf-light corr?
% COMPADRE?
% Model adequacy?
% FIGURE FOR SEM?

\documentclass[12pt, oneside]{article}

\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{topcapt}
% \usepackage[labelformat=empty]{caption} % for empty caption labels
\usepackage{caption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{color}
\usepackage{array}
\usepackage{gensymb}
\usepackage{setspace}
\newcommand{\stretchy}{2}
\usepackage{lineno}
\usepackage{array}
\usepackage{lscape}
\usepackage{multirow}
% \usepackage[none]{hyphenat}

% Make helvetica the default sans-serif font
\renewcommand\sfdefault{phv}

% Package command for citing R packages
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

% Command for abbreviating Ellenberg light indicator value
\newcommand{\el}{L-value}

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Light and life form interact to shape stomatal ratio among British angiosperms$^1$}
\author{Christopher D. Muir$^2$}
\date{} % delete this line to display the current date

\maketitle

$^1$ Manuscript received \underline{     }; revision accepted \underline{     }. \\

$^2$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on British herbs\\
\\
Word count: \\
Summary:  \\
Introduction:  \\
Methods and Results:  \\
Discussion:  \\
\# Figures and \# Tables, \# references

%--------------------------------------------------
% Acknowledgements and Contributions
%--------------------------------------------------

\section*{Acknowledgements}

% \section*{Author contribution statement}
% CDM designed the study, collected data, analyzed the data, and wrote the manuscript.

\linenumbers
\setstretch{\stretchy}
% \pagenumbering{gobble}

<<Preliminaries, eval=TRUE, echo=FALSE, results=hide>>=

  # REMOVE BEFORE PUBLICATION
  system('cp ~/Google\\ Drive/CommonResources/refs.bib ~/Google\\ Drive/StomatalRatio/BEF/ms/refs.bib')

  # Directories
  pathMain <- "~/Google Drive/StomatalRatio/BEF"
  pathRawData <- paste0(pathMain, "/RawData")
  pathProcData <- paste0(pathMain, "/ProcData")
  pathR <- paste0(pathMain, "/R")
  pathMS <- paste0(pathMain, "/ms")
  pathFigures <- paste0(pathMS, "/Figures")

  # Libraries
  library(vioplot)
  library(ape)
  library(rncl)
  library(phytools)
  library(rstanarm)
  library(phylolm)
  library(Rphylopars)
  library(lavaan)
  library(stringr)
  library(magrittr)
  library(taxize)
  library(dplyr)

  # Source functions
  source(paste0(pathR, "/functions.R"))

@

<<Process data, eval=TRUE, echo=FALSE, results=hide>>=

  # Stomata
  ## BEF
  stomata <- read.csv(paste0(pathRawData, "/stomata.csv"), stringsAsFactors = F, 
                      strip.white = T)

  ## Salisbury 1927
  salisbury <- read.csv(paste0(pathRawData, "/salisbury_1927.csv"), 
                        stringsAsFactors = FALSE, strip.white = TRUE)

  # This is done as a loop to prevent timing out
  # tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
  # for (i in 2:nrow(salisbury)) {
  #  tnr %<>% rbind(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
  # }
  # write.csv(temp, file = paste0(pathRawData, "/taxized_salisbury_1927.csv"), row.names = F)
  tnr <- read.csv(paste0(pathRawData, "/taxized_salisbury_1927.csv"), stringsAsFactors = FALSE)

  # When TNRS does not give an accepted name, I am using these based on searching The Plant List
  # http://www.theplantlist.org/
  # Plant List corrections

  old_name <- c("Betula alba",            "Betula pubescens",         "Fagus sylvatica",
                "Ulmus montana",          "Corylus avellana",         "Crataegus oxyacanthoides",
                "Rubus corylifolius",     "Cephalanthera pallens",    "Conopodium denudatum",
                "Helleborus foetidus",    "Primula acaulis",          "Byronia dioica",
                "Geranium phaeum",        "Geranium striatum",        "Hieracium sylvaticum",
                "Hypericum pulchrum",     "Lathyrus macrorrhizus",    "Luzula forsteri",
                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
                "Tilia parvifolia",       "Tilia vulgaris",           "Vicia sylvatica",
                "Iris foetidissima",      "Arrhenatherum avenaceum",  "Hordeum sylvaticum")
  new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylvatica",
                "Ulmus glabra",           "Corylus avellana",         "Crataegus laevigata",
                "Rubus plicatus",         "Cephalanthera longifolia", "Conopodium majus",
                "Helleborus foetidus",    "Primula vulgaris",         "Byronia dioica",
                "Geranium phaeum",        "Geranium versicolor",      "Hieracium murorum",
                "Hypericum pulchrum",     "Lathyrus linifolius",      "Luzula forsteri",
                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
                "Tilia cordata",          "Tilia x europaea",         "Vicia sylvatica",
                "Xyridion foetidissimum", "Arrhenatherum elatius",    "Hordelymus europaeus")

  tnr$acceptedname[match(old_name, tnr$submittedname)] <- new_name
  tnr %<>% mutate(species = submittedname)

  salisbury %<>% inner_join(tnr, by = "species")
  rm(tnr)

  # Find which Salisbury data are already in BEF data
  already_in <- which(salisbury$submittedname %in% stomata$species |
                        salisbury$acceptedname %in% stomata$species)
  x1 <- salisbury$submittedname %in% stomata$species
  salisbury$species[x1] <- salisbury$submittedname[x1]
  x2 <- (salisbury$acceptedname %in% stomata$species) & !x1
  salisbury$species[x2] <- salisbury$acceptedname[x2]

  tmp <- inner_join(stomata, salisbury, by = "species")

  # Overlapping data from BEF and Salisbury are clearly taken from Salisbury, but with some errors. Hence, when they do not match, I have use the Salisbury data I entered myself.
  with(tmp, plot(ab_density.x, ab_density.y))
  with(tmp, plot(ad_density.x, ad_density.y))

  x <- tmp %>%
    use_series(ab_density.x) %>%
    equals(tmp %>% use_series(ab_density.y)) %>%
    not() %>%
    which()

  tmp$ab_density.x[x] <- tmp$ab_density.y[x]

  x <- tmp %>%
    use_series(ad_density.x) %>%
    equals(tmp %>% use_series(ad_density.y)) %>%
    not() %>%
    which()

  tmp$ad_density.x[x] <- tmp$ad_density.y[x]

  tmp %<>% 
    select(species, acceptedname, ab_density.x, ad_density.x) %>%
    set_colnames(c("species", "acceptedname", "ab_density", "ad_density"))

  stomata %<>% 
    left_join(tmp, by = c("species", "ab_density", "ad_density"))

  # Add data from Salisbury not in BEF
  stomata %<>% 
    rbind(salisbury[!(x1 | x2), colnames(.)])

  # Remove duplicates
  stomata <- stomata[!duplicated(stomata$species), ]

  # Calculate stomatal ratio
  get_sr_propAd <- function(X) {
    as.numeric(X["ad_density"]) / (as.numeric(X["ab_density"]) + as.numeric(X["ad_density"]))
  }
  get_sr_even <- function(X) {
    min(as.numeric(X[c("ab_density", "ad_density")])) / 
      max(as.numeric(X[c("ab_density", "ad_density")]))
  }
  stomata %<>% mutate(sr_propAd = apply(., 1, get_sr_propAd),
                      sr_even = apply(., 1, get_sr_even))

  stomata <- stomata[which(!is.na(stomata$sr_propAd)), ]

  # Photosynthetic pathway
  photo <- read.csv(paste0(pathRawData, "/photo.csv"), stringsAsFactors = F, 
                    strip.white = T)

  # All missing taxa are C3 
  # Not C4 based on Sage et al 2011 (JXB 62:9 3155-3169)
  # Not CAM based on photosynthetic pathway of congeneric species in the data
  missing_photo <- 
    c("Alchemilla vulgaris",  "Asarum europaeum", "Byronia dioica",
      "Hieracium sylvaticum", "Luzula maxima",    "Prunus cerasus",
      "Pteridium aquilinum",  "Pyrus communis",   "Pyrus torminalis",
      "Quercus sessiliflora", "Rhamnus frangula", "Ribes grossularia",
      "Rubus corylifolius",   "Tilia vulgaris")
  photo %<>% rbind(data.frame(species = missing_photo, photo = "c3"))
  all(stomata$species %in% photo$species |
        stomata$acceptedname %in% photo$species) # should be true

  # PLANTATT (Hill et al 2004) for lifeform and Ellenberg light indicator values
  plantatt <- read.csv(paste0(pathRawData, "/plantatt.csv"), stringsAsFactors = F, 
                       strip.white = T)
  # stomata$species[!(stomata$species %in% plantatt$Taxon.name |
  #                    stomata$acceptedname %in% plantatt$Taxon.name)] # Missing data for several

  # Combine bulbous and nonbulbous geophytes
  plantatt$LF1[plantatt$LF1 == "Gb"] <- "Gn"

  # Combine hydrophytes and annual hydrophytes
  plantatt$LF1[plantatt$LF1 == "Hz"] <- "Hy"

  # Combine phanerophytes and nanophanerophytes
  plantatt$LF1[plantatt$LF1 == "Pn"] <- "Ph"

  ##### MESSING AROUND --------------------------------------------
  # Leaf size
  leaf_size <- read.csv(paste0(pathRawData, "/leaf_size.csv"), stringsAsFactors = F)
  leaf_size$leaf_size %<>% ordered(levels = c("<0.1", "0.1-1", "1-10", "10-100", "100-1000",
  																						">1000"))
  leaf_size <- read.csv(paste0(pathRawData, "/leda_leaf_area.csv"), stringsAsFactors = F,
  											strip.white = TRUE)
  leaf_size %<>% rename(species = SBS.name, leaf_size = single.value..mm.2.)
  leaf_size <- data.frame(leaf_size = tapply(leaf_size$leaf_size, leaf_size$species, mean))
  leaf_size$species <- rownames(leaf_size)

  seed_number <- read.csv(paste0(pathRawData, "/leda_seed_number.csv"), stringsAsFactors = F,
  											strip.white = TRUE)
  seed_number %<>% rename(species = SBS.name, seed_number = single.value)
  seed_number <- data.frame(seed_number = tapply(seed_number$seed_number, seed_number$species, mean))
  seed_number$species <- rownames(seed_number)
  which(!(stomata$species %in% seed_number$species |
  			stomata$acceptedname %in% seed_number$species))

  #####-----------------------------

  # Many missing leaf size
  which(!(stomata$species %in% leaf_size$species |
  			stomata$acceptedname %in% leaf_size$species))

  # Combine datasets
  xp <- match(stomata$species, photo$species)
  xpa <- match(stomata$acceptedname, photo$species)
  xp[is.na(xp)] <- xpa[is.na(xp)]

  xa <- match(stomata$species, plantatt$Taxon.name)
  xaa <- match(stomata$acceptedname, plantatt$Taxon.name)
  xa[is.na(xa)] <- xaa[is.na(xa)]

  xl <- match(stomata$species, leaf_size$species)
  xla <- match(stomata$acceptedname, leaf_size$species)
  xl[is.na(xl)] <- xla[is.na(xl)]

  xs <- match(stomata$species, seed_number$species)
  xsa <- match(stomata$acceptedname, seed_number$species)
  xs[is.na(xs)] <- xsa[is.na(xs)]

  stomata$photo <- photo$photo[xp]
  stomata$lifeform <- plantatt$LF1[xa]
  stomata$ellenberg_light <- plantatt$L[xa]
  stomata$height <- plantatt$Hght[xa]
  stomata$wh <- plantatt$W[xa]
  stomata$leaf_size <- leaf_size$leaf_size[xl]
  stomata$seed_number <- seed_number$seed_number[xs]

  # Remove missing values
  stomata <- stomata[complete.cases(stomata[, c("sr_propAd", "lifeform", "ellenberg_light")]), ]

  rm("already_in", "missing_photo", "new_name", "old_name", "photo", 
     "plantatt", "salisbury", "tmp", "x", "x1", "x2", "xp", "xpa", "xa", "xaa")

  # Export  
  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))

@

<<Leaf size, eval=TRUE, echo=FALSE, results=hide>>=

  # boxplot(sr_even ~ leaf_size, data = stomata)
  # boxplot(ellenberg_light ~ leaf_size, data = stomata)

  plot(sr_even ~ leaf_size, data = stomata, log = "x")
  boxplot(log10(leaf_size) ~ ellenberg_light, data = stomata)
  tmp <- read.csv(paste0(pathRawData, "/try_leaf_area.csv"), stringsAsFactors = FALSE,
  								strip.white = TRUE)

  stomata$species %in% tmp$AccSpeciesName | stomata$acceptedname %in% tmp$AccSpeciesName
 sort(table(tmp[which(tmp$AccSpeciesName %in% stomata$species), c("Dataset")]))
@

<<Phylogeny, eval=TRUE, echo=FALSE, results=hide>>=

  # Import phylogeny of British Flora from Lim et al 2014
  tmp <- readLines(paste0(pathRawData, "/S15105.nex"))
  l1 <- grep("BEGIN CHARACTERS", tmp) - 1
  l2 <- grep("BEGIN TREES", tmp) - 1
  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathRawData, "/Lim_etal_2014.nex"))
  phy <- read_nexus_phylo(paste0(pathRawData, "/Lim_etal_2014.nex"))

  # Check for species missing from phylogeny
  stomata$species %<>% str_replace_all(" ", "_")
  stomata$acceptedname %<>% str_replace_all(" ", "_")
  notInPhy <- which(!(stomata$species %in% phy$tip.label |
                        stomata$acceptedname %in% phy$tip.label))
  stomata[notInPhy, ]; length(notInPhy)

  # Modify tip labels

  # Using Cerastium brachypetalum as proxy for C. cerastoides based on similar divergence in Scheen et al. 2004 (Am J Bot, 91(6): 943-952)
  tmp <- gsub("Cerastium_brachypetalum", "Cerastium_cerastoides", tmp)

  # Using Coincya monensis as proxy for C. wrightii 
  tmp <- gsub("Coincya_monensis", "Coincya_wrightii", tmp)

  # Using Geranium pusillum as proxy for G. versicolor based on similar divergence in Palazzesi et al. 2012 (Bio J of Linn Soc, 107(1): 67-85)
  # tmp <- gsub("Geranium_pusillum", "Geranium_versicolor", tmp)

  # Using Gnaphalium luteoalbum as proxy for G. norvegicum 
  tmp <- gsub("Gnaphalium_luteoalbum", "Gnaphalium_norvegicum", tmp)

  # Lithospermum purpurocaeruleum is mispelled
  tmp <- gsub("Lithospermum_purpureocaeruleum", "Lithospermum_purpurocaeruleum", tmp)

  # Using Lythrum portula as proxy for L. hyssopifolia 
  tmp <- gsub("Lythrum_portula", "Lythrum_hyssopifolia", tmp)

  # Using Peucedanum ostruthium as proxy for P. officinale 
  tmp <- gsub("Peucedanum_ostruthium", "Peucedanum_officinale", tmp)

  # Changing Populus_nigra_sens.lat. to Populus nigra officinale 
  tmp <- gsub("Populus_nigra_sens.lat.", "Populus_nigra", tmp)

  # Changing Rosa_canina_agg. to Rosa_canina officinale 
  tmp <- gsub("Rosa_canina_agg.", "Rosa_canina", tmp)

  # Using Stachys alpina as proxy for S. germanica based on similar divergence in Salmaki et al. 2013 (Mol Phylo Evol, 69(3): 535-551)
  tmp <- gsub("Stachys_alpina", "Stachys_germanica", tmp)

  # Using Tripleurospermum_maritimum as proxy for T. inodorum 
  tmp <- gsub("Tripleurospermum_maritimum", "Tripleurospermum_inodorum", tmp)

  # Using Viola_arvensis as proxy for V. kitaibeliana 
  tmp <- gsub("Viola_arvensis", "Viola_kitaibeliana", tmp)

  # Write first modified phylogeny
  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
  phy <- read_nexus_phylo(paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))

  # Bind additional tips
  tips <- phy$tip.label
  nodes <- sapply(tips, function(x,y) which(y == x), y = phy$tip.label)
  edge.lengths <- setNames(phy$edge.length[sapply(nodes, function(x, y) {
    which(y == x) }, y = phy$edge[,2])], names(nodes))

  # Make Agrostemma githago sister to all (Silene sp., Lychnis sp.) [based on Fior et al. 2006, Am J Bot 104(2):399-411]
  n <- getMRCA(phy, phy$tip.label[c(grep("Silene", phy$tip.label), grep("Lychnis", phy$tip.label))])

  # Branch length is average of distance from (Silene sp., Lychnis sp.) MRCA to (Silene sp., Lychnis sp.) sp.
  el1 <- sapply(nodes[c(grep("Silene_", phy$tip.label), grep("Lychnis_", phy$tip.label))], 
                nodeheight, tree = phy)
  el2 <- nodeheight(tree = phy, node = n)

  phy <- bind.tip(phy, "Agrostemma_githago", edge.length = mean(el1 - el2), where = n)

  # Make Heracleum mantegazzianum as sister to H. sphondylium
  phy <- bind.tip(phy, "Heracleum_mantegazzianum", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Heracleum_sphondylium")],
                  where = which(phy$tip.label == "Heracleum_sphondylium"))

  # Make Portulaca oleracea sister to all Claytonia (based on Angiosperm Phylogeny Website v 13 http://www.mobot.org/mobot/research/apweb/welcome.html]
  # n <- getMRCA(phy, phy$tip.label[grep("Claytonia", phy$tip.label)])

  # Branch length is average of distance from Claytonia MRCA to Claytonia sp.
  # el1 <- sapply(nodes[grep("Claytonia_", names(nodes))], nodeheight, tree = phy)
  # el2 <- nodeheight(tree = phy, node = n)

  # phy <- bind.tip(phy, "Portulaca_oleracea", edge.length = mean(el1 - el2), where = n)

  # Make Spartina alterniflora, Spartina maritima, and Spartina x townsendii as sister to S. anglica
  phy <- bind.tip(phy, "Spartina_alterniflora", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
                  where = which(phy$tip.label == "Spartina_anglica"))

  phy <- bind.tip(phy, "Spartina_maritima", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
                  where = which(phy$tip.label == "Spartina_anglica"))

  # phy <- bind.tip(phy, "Spartina_x_townsendii", 
  #                 edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
  #                 where = which(phy$tip.label == "Spartina_anglica"))

  # Make Tephroseris integrifolia sister to all Senecio sp.
  n <- getMRCA(phy, phy$tip.label[grep("Senecio", phy$tip.label)])

  # Branch length is average of distance from Senecio MRCA to Senecio sp.
  el1 <- sapply(nodes[grep("Senecio_", names(nodes))], nodeheight, tree = phy)
  el2 <- nodeheight(tree = phy, node = n)

  phy <- bind.tip(phy, "Tephroseris_integrifolia", edge.length = mean(el1 - el2), where = n)

  # Make Viola hirta as sister to V. reichenbachiana (both in Viola sect. Viola)
  phy <- bind.tip(phy, "Viola_hirta", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Viola_reichenbachiana")],
                  where = which(phy$tip.label == "Viola_reichenbachiana"))

  # Write second modified phylogeny
  write.nexus(phy, file = paste0(pathProcData, "/Lim_etal_2014_mod2.nex"))

  # Check for species missing from modified phylogeny
  notInPhy <- which(!(stomata$species %in% phy$tip.label |
                        stomata$acceptedname %in% phy$tip.label))
  stomata[notInPhy, ]; length(notInPhy)

  rm("edge.lengths", "el1", "el2", "l1", "l2", "n", "nodes", "notInPhy", "tips", "tmp")

  # Run PATHd8 (Britton et al. 2002)
  cat(sprintf("Sequence length = 4692;

              %s

              mrca: Azolla_filiculoides, Solanum_nigrum, maxage=454;
              mrca: Osmunda_regalis, Asplenium_onopteris, minage=354;
              mrca: Pinus_sylvestris, Solanum_nigrum, minage=310;
              mrca: Nymphaea_alba, Solanum_nigrum, minage=140;
              mrca: Nymphaea_alba, Solanum_nigrum, maxage=217;
              mrca: Nymphaea_alba, Nuphar_lutea, minage=115;
              mrca: Chelidonium_majus, Solanum_nigrum, fixage=121;
              mrca: Quercus_robur, Bryonia_dioica, minage=84;

              name of mrca: Azolla_filiculoides, Solanum_nigrum, name=tracheophytes;
              name of mrca: Osmunda_regalis, Asplenium_onopteris, name=monilophytes;
              name of mrca: Pinus_sylvestris, Solanum_nigrum, name=seedplants;
              name of mrca: Nymphaea_alba, Solanum_nigrum, name=angiosperms;
              name of mrca: Nymphaea_alba, Nuphar_lutea, name=nymphaceae;
              name of mrca: Chelidonium_majus, Solanum_nigrum, name=eudicots;
              name of mrca: Quercus_robur, Bryonia_dioica, name=curcibitales_fagales;

              ", write.tree(phy)), file = paste0(pathMain, "/PATHd8/pathd8_in"))

  # Run PATHd8
  system("cd ~/Google\\ Drive/StomatalRatio/BEF/PATHd8; ./PATHd8 pathd8_in pathd8_out")

  # Import ultrametric tree
  pathd8_out <- readLines(paste0(pathMain, "/PATHd8/pathd8_out"))
  phy <- read.tree(text = substr(pathd8_out[11], 14, nchar(pathd8_out[11])))

  rm("pathd8_out")

@

<<Violin plot of Raunkiaer life form versus stomatal ratio, eval=TRUE, echo=FALSE, results=hide>>=

  # chamaephyte = subshrub (woody)
  # geophyte = cryptophyte resting in dry ground
  # hemicryptophyte = perennial with overwintering buds at soil surface
  # hydrophyte = cryptophyte resting in marshy ground
  # phanerophyte = normally woody perennial
  # therophyte = annual

  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
  names(lf) = c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

  pdf(paste0(pathFigures, "/FigureS_violin.pdf"), 8, 5.25)
  par(mai = c(1.75, 1.5, 0.25, 0.25), cex.lab = 1.5, las = 1, cex = 1.5, cex.axis = 0.5)

  vioplot(subset(stomata$sr_propAd, stomata$lifeform == names(lf[1]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[2]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[3]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[4]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[5]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[6]), na.rm = TRUE),
          col = "grey", colMed = "black", names = rep("", 6))

  clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
       grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
  # Sample sizes
  text(1:6 + 0.25, grconvertY(1.5, "inches", "user"), labels = lf, srt = 30, pos = 2)
  mtext(table(stomata$lifeform)[names(lf)], 3, at = 1:6, line = 0)
  title(ylab = "Stomatal Ratio",  line = 3.5)
  title(xlab = "Raunkiaer Life Form", line = 4.5)
  title(ylab = expression(SR[propAd] == SD[ad] / SD[total]), line = 2, cex.lab = 1) 
  dev.off()

  rm("lf")

@

<<Remove hydrophytes + c4 + cam + nonangiosperm, eval=TRUE, echo=FALSE, results=hide>>=

  nHydrophyte <- length(which(stomata$lifeform == "Hy"))
  nC4 <- length(which(stomata$photo == "c4"))
  nCAM <- length(which(stomata$photo == "cam"))

  x1 <- c("Ph", "Ch", "hc", "Gn", "Th")
  stomata <- stomata[which(stomata$lifeform %in% x1 & stomata$photo == "c3"), ]

  angioPhy <- extract.clade(phy, 
                            node = length(phy$tip.label) + which(phy$node.label == "angiosperms"))

  nNonangio <- length(which(!(stomata$species %in% angioPhy$tip.label | 
                                stomata$acceptedname %in% angioPhy$tip.label)))

  stomata <- stomata[stomata$species %in% angioPhy$tip.label |
                       stomata$acceptedname %in% angioPhy$tip.label, ]

  # Export new dataset and tree
  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
  write.nexus(angioPhy, file = paste0(pathProcData, "/angioPhy.nex"))

  rm("x1", "phy")

@

\section*{Abstract}

% (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including key points of discussion?

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

% POSSIBLE CONCLUSION: THIS TELLS US THAT CO2 LIMITS PHOTOSYNTHESIS AND HIGH LIGHT, BUT ONLY STRONG SELECTION IN FAST LIFE HISTORY (HERBS)...OR SOMETHING

\section*{Keywords}

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, stomata, stomatal ratio

\section*{INTRODUCTION}

%%%%% PARAGRAPH 1-2: general stuff about adaptation, stomata, stomatal ratio %%%%%
Natural selection shapes leaf anatomy in order to optimize its photosynthetic function in a given environment \citep{Haberlandt_1914, Givnish_1987, Smith_etal_1997}. By understanding the adaptive significance of leaf anatomical variation we can learn about natural history, find targets for crop improvement, and identify anatomical proxies for paleoclimates preserved in the fossil record [CITE]. The size, density, and distribution of stomata on a leaf vary widely and impact functions like the maximum photosynthetic rate, water-use efficiency, photosynthetic nitrogen-use efficiency, and susceptibility to foliar pathogens that infect through stomata [CITATIONS]. Hence, stomata have been especially useful in understanding plastic and evolutionary response to climate change and domestication (Royer, Ward, Woodward, Beerling, Milla et al...).

While the density and size of stomata have been researched extensively [CITATIONS], the adaptive significance of stomatal distribution is less well understood. Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}. Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions. By creating a second parallel pathway for CO$_2$ diffusion within the mesophyll, amphistomy optimally supplies CO$_2$ \citep{Parkhurst_1978, Gutschick_1984b, Jones_1985}. Amphistomy is correlated with greater CO$_2$ diffusion \citep{Beerling_Kelly_1996} and higher photosynthetic rates \citep{Mckown_etal_2014}. These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}. However, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986}, though empirical studies suggest great water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}. Hence, amphistomy clearly should be favored when CO$_2$ limits photosynthetic rate, but the open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?

% NOTES FOR FIRST TWO PARAGRAPHS
% Stomatal ratio and photo/water use: Parkhurst papers, 
% from Peat and Fitter 1994b "Jones (1985) found that amphistomaty may be advantageous as it allows greater CO2 transfer through the mesophyll, especially in thicker leaves, and greater leaf conductance in high light conditions when CO2 is limiting photosynthetic rate. However, his models showed that hypostomaty would be advantageous when there are high boundary layer resistances (for example in large leaves or deep within a plant canopy), high humidity or a temperature gradient across the leaf."

%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitat (i.e. high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitat are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from ~90 familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates. \citep{Bazzaz_1979, Korner_etal_1989}. NEED MORE RECENT CITATIONS ON THIS.

% NOTES FOR PARAGRAPHS 3-4:
% From Mott et al 1982: ''Korner, Scheel \& Bauer (1979) have recently reviewed the literature on maximum stomatal conductance and photosynthetic rates for plants of specific habitats and growth forms. Low maximum photosynthetic rates are observed for succulents, evergreen conifers, deciduous woody plants, herbs from shady habitats, evergreen woody plants, and desert and steppe shrubs. Higher maximum values are obtained for deciduous fruit trees, wild grasses, cultivated C$_4$ and C$_3$ grasses, herbaceous crop plants, herbs from open habitats, and plants from aquatic habitats and swamps. Bazzaz (1979) cites evidence that the maximum photosynthetic rate declines through successional stages in a plant community, pointing to higher rates for open field plants and grasses and lower rates for canopy deciduous trees and shade habitat herbs. Early successional trees appear to be intermediate in photosynthetic rates.''

%%%% PARAGRAPH 5-6 %%%%%
Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form are probably not independent because open habitat by definition consists of more short-statured, herbaceous plants. Even attempts to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927} are problematic because high light habitats select for faster life history stratagies \citep{Galloway_Etterson_2007} [<-- OTHER REFS IN THAT PAPER?]. Open habitat and groth form may also interact with one another. For example, amphistomy may only favored when CO$_2$ strongly limits photosynthetic rate \textit{and} photosynthetic rate strongly limits fitness.

To better understand the adapative significance of stomatal ratio, I asked three main questions:

\begin{enumerate}

  \item{Are light habitat and growth form correlated?}
  \item{Do light habitat and growth form influence stomatal ratio additively, or do their effects interact?}
  \item{Is evolution of stomatal ratio mediated by changes in stomatal density on the adaxial (upper) surface, abaxial (lower) surface, or both?}

\end{enumerate}

The final question is important for telling whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be limited. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.

% from previous paper
% Here, we examine whether stomata on the abaxial (`lower') surface of the leaf evolve independently of adaxial (`upper') stomata. Stomata are microscopic pores on the leaf surface formed by a pair of guard cells. The density, size, and arrangement of stomata on a leaf set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it \citep{Parkhurst_1978, Sack_etal_2003, Franks_Farquhar_2001, Galmes_etal_2013}. Hence, stomatal traits like density, size, and ratio of upper to lower stomata have strong effects on carbon assimilation and water-use efficiency.

% The proportion of stomata found on the upper surface also tends to increase during domestication, even as the total stomatal density stays constant \citep{Milla_etal_2013}. Amphistomy increases CO$_2$ diffusion within the leaf by opening up a second parallel pathway in the intercellular airspace for diffusion from substomatal cavities to mesophyll cell walls. However, stomata on the upper surface in particular may be costly. For example, upper stomata increase the susceptibility to rust pathogens in \textit{Populus} \citep{Mckown_etal_2014}. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits \citep{Buckley_etal_2015}. \cite{Muir_2015} reviewed the literature on other possible fitness costs.

%--------------------------------------------------
% Methods
%--------------------------------------------------

\section*{METHODS}

\subsection*{Data on stomatal ratio, light habitat, growth form, and phylogenetic relationships}

I obtained data on ab- and adaxial stomatal density on 370 species from British Ecological Flora \citep{Fitter_Peat_1994a, BEF}. I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). Seventeen species (4.6\%) in the dataset were not present in the phylogeny. For nine species, I used the position a congeneric species as a proxy for the focal species. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species. For the remaining eight missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.5-64 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

I excluded data from hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants.. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species. The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.

Following \cite{Muir_2015}, I calculated stomatal ratio in two different ways depending on what was most appropriate for the question:

\begin{equation} \label{eq:SRpropAd} 
  \mathrm{SR_{propAd}} = \frac{\mathrm{SD_{ad}}}{\mathrm{SD_{total}}}
\end{equation}

\begin{equation} \label{eq:SReven1} 
  \mathrm{SR_{even}} = \frac{\mathrm{min}\{\mathrm{SD_{ab}}, \mathrm{SD_{ad}}\}}{\mathrm{max}\{\mathrm{SD_{ab}}, \mathrm{SD_{ad}}\}}
\end{equation}

$\mathrm{SD_{ab}}$ and $\mathrm{SD_{ad}}$ are the stomatal densities on abaxial or adaxial surface, respectively. $\mathrm{SD_{total}} = \mathrm{SD_{ab}} + \mathrm{SD_{ad}}$. $\mathrm{SR_{propAd}}$ is the proportion of stomata density on the adaxial surface, which is useful for discriminating among hypostomatous ($\mathrm{SR_{propAd}} = 0$), amphistomatous (0 < $\mathrm{SR_{propAd}} < 1$), and hyperstomatous species ($\mathrm{SR_\mathrm{propAd}} = 1$). $\mathrm{SR_\mathrm{even}}$ indicates how evenly stomatal densities are distributed across both leaf surfaces. This expression is useful because several hypotheses are based on the fact that a more even distribution should optimize leaf CO$_2$ diffusion.

% NOTES

%  papers on Ellenberg light value:
%  Bartelheimer and Poschlod 2015
%  Diekmann, 2003
%  Ellenberg, 1974
%  Ellenberg et al., 1991
%  Silvertown et al., 2006

\subsection*{Testing for an association between open habitat and growth form}

<<Raunkiaer life form versus Ellenberg light indicator values, eval=TRUE, echo=FALSE, results=hide>>=

  # Make dichotomous and slightly adjust edge lengths to make exactly ultrametric
  modAngioPhy <- multi2di(angioPhy)
  modAngioPhy$edge.length[modAngioPhy$edge.length == 0] <- 0.02
  x <- diag(vcv.phylo(modAngioPhy))
  is.ultrametric(modAngioPhy)
  n <- length(modAngioPhy$tip.label)
  te <- sapply(1:n, function(x, y) which(y == x), y = modAngioPhy$edge[, 2]) # terminal edges
  modAngioPhy$edge.length[te] <- modAngioPhy$edge.length[te] + (max(x) - x)
  is.ultrametric(modAngioPhy)

  x <- which(stomata$acceptedname %in% modAngioPhy$tip.label &
               !(stomata$species %in% modAngioPhy$tip.label))
  stomata$species[x] <- stomata$acceptedname[x]

  x <- stomata$species[duplicated(stomata$species)]
  stomata[stomata$species %in% x, ]
  modAngioPhy <- drop.tip(modAngioPhy, 
                          modAngioPhy$tip.label[!modAngioPhy$tip.label %in% stomata$species])

  # phylolm shows that there is no phylogenetic signal in this regression (very high alpha)
  rownames(stomata) <- stomata$species # need to change row names for phylolm
  fitLFvEL_pl <- phylolm(ellenberg_light ~ -1 + lifeform, model = "OUrandomRoot", data = stomata, 
                         phy = modAngioPhy, upper.bound = 1e10)
  summary(fitLFvEL_pl)

  # therefore using standard ANOVA
  fitLFvEL <- lm(ellenberg_light ~ lifeform, data = stomata)
  aovLFvEL <- anova(fitLFvEL)

  # for plotting confidence intervals
  fitLFvEL <- lm(ellenberg_light ~ -1 + lifeform, data = stomata)

  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
  names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

  pdf(paste0(pathFigures, "/Figure_lf-light.pdf"), 4, 7)
  par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
  plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(0, 5), xlab = "", ylab = "", axes = F)

  for (i in 0:4) abline(h = i + seq(0.2, 0.6, 0.2), col = "grey")
  hs <- tapply(stomata$ellenberg_light, stomata$lifeform, hist, plot = F, breaks = 0:9)  
  for (i in 1:5) {
    rect(0:8, rep(i - 1, 9), 1:9, i - 1 + hs[[names(lf)[i]]]$density, col = "black", 
         border = "white", lwd = 2)
  }

  # label life forms
  s <- sprintf("%s (%s)", lf, table(stomata$lifeform)[names(lf)])
  text(0 + 0.5 * strwidth(s), 1:5, labels = s, pos = 1)

  # axes
  axis(1, at = 0.5:8.5, labels = 1:9, lwd = 0, line = -1.5)
  for (i in 0:4) axis(2, at = i + seq(0.2, 0.6, 0.2), labels = seq(0.2, 0.6, 0.2), lwd = 0, 
                      lwd.ticks = 1)

  # axis labels
  mtext(c("Shade\n", "Partial\nShade", "Sun\n"), 1, at = c(1.5, 4.5, 7.5), line = 1.5)
  title(xlab = "Ellenberg light\nindicator", cex.lab = 1.5, line = 5)
  title(ylab = "Proportion", cex.lab = 1.5)

  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
  rect(grconvertX(0, from = "npc", "user"), 0:4, 
       grconvertX(1, from = "npc", "user"), 0.75:4.75)

  # Means + 95% confidence intervals
  ci <- confint(fitLFvEL)
  ci <- ci[paste0("lifeform", names(lf)), ] # reorder
  mu <- coef(fitLFvEL)[paste0("lifeform", names(lf))]
  for (i in 1:5) {
    lines(ci[i, ], rep(i - 1 + 0.675, 2), lwd = 2)
    points(mu[i], i - 1 + 0.675, pch = 21, col = "black", bg = "white")
  }

  dev.off()

  rm("ci", "fitLFvEL_pl", "hs", "i", "mu", "n", "s", "te", "x")

@

I tested whether Raunki\ae r life form was associated \el~values among British angiosperms using ANOVA with Type-2 sum of squares. I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. See the R code accompanying this paper for further detail \citep{Muir_dryad}. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \el~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.

\subsection*{Open habitat, growth form, and stomatal ratio}

<<Raunkiaer life form and Ellenberg light indicator values versus SR_even, eval=TRUE, echo=FALSE, results=hide>>=

  # Phylogenetic regression model comparison
  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRb <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRd <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  aicSR <- sapply(list(fitSRa, fitSRb, fitSRc, fitSRd, fitSRe), AIC)

  # Parametric bootstrap CIs of full model for p-values
  # set.seed(36502)
  # fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, model = "OUrandomRoot",
  #                    data = stomata, phy = modAngioPhy, boot = 1e4, upper.bound = 10)
  # saveRDS(fitSR_pl, file = paste0(pathR, "/fitSR_pl.rds"))
  fitSR_pl <- readRDS(file = paste0(pathR, "/fitSR_pl.rds"))

  # Figure

  pdf(paste0(pathFigures, "/Figure_SRmultReg.pdf"), 4, 7)
  par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))

  for (i in 5:1) {

    plot(0, 0, type = "n", xlim = c(1, 9), ylim = c(-0.05, 1.05), xlab = "", ylab = "", axes = F,
         frame.plot = F)
    if (i %in% seq(1, 5, 2)) axis(2, at = seq(0, 1, 0.5), lwd = 0, lwd.ticks = 1, las = 1)

    # polygon of confidence intervals
    x <- seq(min(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]),
             max(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]), length.out = 1e3)
    bs <- sapply(x, function(X) {
      fitSR_pl$bootstrap[, sprintf('lifeform%s', names(lf)[i])] + 
        fitSR_pl$bootstrap[, sprintf('lifeform%s:ellenberg_light', names(lf)[i])] * X
    } )
    ci <- apply(bs, 2, quantile, probs = c(0.025, 0.5, 0.975))
    polygon(c(x, rev(x)), c(ci[1, ], rev(ci[3, ])), col = "grey", border = NA)
    points(x, ci[2, ], lwd = 2, type = "l") # median
    points(x, ci[1, ], lwd = 2, lty = 2, type = "l") # lower 95%
    points(x, ci[3, ], lwd = 2, lty = 2, type = "l") # upper 95%

    # plot points
    with(subset(stomata, stomata$lifeform == names(lf)[i]), 
         points(jitter(ellenberg_light), sr_even, pch = 21, bg = rgb(0, 0, 0, 0.1), cex = 2))

    # label life forms, slope, and statistical significance
    s1 <- sprintf("%s (%s)", lf[i], table(stomata$lifeform)[names(lf)[i]])
    text(1 + 0.5 * strwidth(s1), 1.05, labels = s1, pos = 1, adj = 1)
    slopeCI <- confint(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", names(lf)[i]), ]
    pval <- 1 - length(which(fitSR_pl$bootstrap[, sprintf("lifeform%s:ellenberg_light", 
                                                          names(lf)[i])] > 0)) / fitSR_pl$boot 
    s2 <- bquote(slope == .(round(coef(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", 
                                                         names(lf)[i])], 3))^.(sigStar(pval)))
    text(1 + 0.5 * strwidth(s2), 1.05 - 1.5 * strheight(s1), labels = s2, pos = 1, adj = 1)

  }

  # frame plot
  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
  abline(h = grconvertY(0:5, "npc", "user"))
  lines(rep(grconvertX(0, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))
  lines(rep(grconvertX(1, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))

  # axis labels
  axis(1, at = 1:9, labels = 1:9, lwd = 0, line = -1)
  axis(1, at = c(2, 5, 8), lwd = 0, labels = c("Shade", "Partial Shade", "Sun"),
       line = 0)
  mtext(1, text = "Ellenberg light indicator", cex = 1.5, line = 3, outer = T)

  mtext(2, text = "Stomatal Ratio",  line = 4.5, outer = T, cex = 1.5)

  mtext(2, text = expression(
    SR[even] == plain(min)~group("(", list(SD[ab], SD[ad]), ")") /
      plain(max)~group("(", list(SD[ab], SD[ad]), ")")), outer = T, line = 2.5)

  dev.off()

  # Parametric bootstrap statistical signicance of deltaAIC between model 
  # with interaction (fitSRa) and model without interaction (fitSRb)
  # set.seed(444294)
  # bootModB <- predict(fitSRb) + rTrait(n = 1e3, modAngioPhy, "OU", 
  #                                     list(sigma2 = fitSRb$sigma2, alpha = fitSRb$optpar))
  # colnames(bootModB) <- paste0("b", 1:ncol(bootModB))

  # fit model of model A to data simulated from model B
  # fitA2B <- lapply(colnames(bootModB), function(col) {
  #  stomata$Y <- bootModB[, col]
  #  phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata, 
  #          phy = modAngioPhy, upper.bound = 10)
  #  })

  # fit model of model A to data simulated from model B
  # fitB2B <- lapply(colnames(bootModB), function(col) {
  #  stomata$Y <- bootModB[, col]
  #  phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata, 
  #          phy = modAngioPhy, upper.bound = 10)
  #  })

  # refit models at parameter bounds
  # for (i in which(round(sapply(fitA2B, function(X) X$optpar), 2) == 10)) {
  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
  #  fitA2B[[i]] <- phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata,
  #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200, 
  #                         starting.value = 10)
  # }

  # refit models at parameter bounds
  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 10)) {
  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
   #  fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
   #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200,
   #                        starting.value = 10)
   # }

  # for some reason, this fixes those with alpha estimated at upper bound
  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 200)) {
  #   stomata$Y <- bootModB[, colnames(bootModB)[i]]
  #   fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
  #                          phy = modAngioPhy, lower.bound = 0, upper.bound = 300, 
  #                          starting.value = 1)
  # }

  # saveRDS(fitA2B, paste0(pathR, "/fitA2B.rds"))
  # saveRDS(fitB2B, paste0(pathR, "/fitB2B.rds"))

  fitA2B <- readRDS(paste0(pathR, "/fitA2B.rds"))
  fitB2B <- readRDS(paste0(pathR, "/fitB2B.rds"))

  # Distribution of deltaAIC
  obsDeltaAIC <- fitSRa$aic - fitSRb$aic
  distDeltaAIC <- sapply(fitA2B, function(X) X$aic) - sapply(fitB2B, function(X) X$aic)
  pInteraction <- length(which(obsDeltaAIC > distDeltaAIC)) / 1e3

  rm("bs", "ci", "distDeltaAIC", "fitA2B", "fitB2B", "i", "lf", "obsDeltaAIC", "pval",
     "s1", "s2", "slopeCI", "x")

@

 <<Testing leaf size model, eval=TRUE, echo=FALSE, results=hide>>=

  stomata$leaf_size %<>% replace(. == 0, NA)
  stomata %<>% mutate(log_leaf_size = log10(leaf_size))
  tmp <- drop.tip(modAngioPhy, stomata$species[is.na(stomata$leaf_size)])

  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light * log_leaf_size, model = "OUrandomRoot",
                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
  fitSRb <- phylolm(sr_even ~ lifeform * ellenberg_light + log_leaf_size, model = "OUrandomRoot",
                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
  fitSRc <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)

@

I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10,000 parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

\subsection*{Does ab- or adaxial stomatal density contribute more to stomatal ratio evolution?}

I used two complementary phylogenetic methods to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution. The contribution of each can be formalized using standard variance decomposition methods as derived below. Because stomatal density is highly skewed, I log-transformed values for normality:

\begin{equation} \label{eq:SReven2} 
  \mathrm{SR_{even}} = \frac{\mathrm{SD_{ad}}}{\mathrm{SD_{ab}}}
\end{equation}

\begin{equation} \label{eq:logSReven} 
  \mathrm{log(SR_{even})} = \mathrm{log(SD_{ad})} - \mathrm{log(SD_{ad})}
\end{equation}

\begin{equation} \label{eq:SReven3} 
  \mathrm{sr_{even}} = \mathrm{sd_{ad}} - \mathrm{sd_{ad}}
\end{equation}

Lowercase variables ($\mathrm{sr}$, $\mathrm{sd}$) indicate log-transformed values. Because some species had zero adaxial stomata, I added one to all values prior to log-transformation. For simplicity, I have defined $\mathrm{SR_{even}}$ here as the ratio of ad- to abaxial stomatal density because in most cases adaxial stomatal density is lower than abaxial (see Eq.~\ref{eq:SReven1}). The variance in $\mathrm{sr_{even}}$ can be decomposed into contributions of $\mathrm{sd_{ad}}$, $\mathrm{sd_{ab}}$, and their covariance:

\begin{equation} \label{eq:varDecomp}
	\mathrm{Var(sr_{even})} = \mathrm{Var(sd_{ad})} + \mathrm{Var(sd_{ad})} - 2 \mathrm{Cov(sd_{ad}, sd_{ab})}
\end{equation}

I estimated the phylogenetic covariance matrix between \el, $\mathrm{sd_{ab}}$, and $\mathrm{sd_{ad}}$ using a multivariate Ornstein-Uhlenbeck model fit in \pkg{Rphylopars} version 0.2.9 \citep{Goolsby_etal_2016, Goolsby_etal_2017}. From the covariance matrix, I estimated the contribution of abaxial density, adaxial density, and their covariance as:

\begin{equation} \label{eq:contribution}
	\frac{\mathrm{Var(sd_{ad})}}{\mathrm{Var(sr_{even})}}, \frac{\mathrm{Var(sd_{ab})}}{\mathrm{Var(sr_{even})}},~\textrm{and}~\frac{\mathrm{Cov(sd_{ad}, sd_{ab})}}{\mathrm{Var(sr_{even})}},
\end{equation}

respectively. Note that when ab- and adaxial densities positively covary, the contribution will be negative because this reduces the variance in stomatal ratio.

I was interested in whether light-mediated evolution of stomatal ratio acted mostly by increasing adaxial stomatal density while maintaining abaxial density, or keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface. The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$. The second scenario predicts that \el~acts similarly on both and that there is a negative covariance $\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$. I tested these competing predictions by fitting a simple phylogenetic structural equation model (SEM). The model uses the phylogenetic covariance matrix to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e. estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$). To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}. I tested whether parameter estimates were significantly different than zero using $z$-scores.

<<Stomatal density + light + growth form, eval=TRUE, echo=FALSE, results=hide>>=

  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", 
          "phanerophyte", "therophyte")
  names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

  # density of stomatal density by side and L-value
  dnsAb <- tapply(log10(stomata$ab_density + 1), stomata$ellenberg_light, function(X) {
    bp <- boxplot(X, plot = F)
    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
    ret$stats <- bp$stats
    ret$out <- bp$out
    ret
  } )

  dnsAd <- tapply(log10(stomata$ad_density + 1), stomata$ellenberg_light, function(X) {
    bp <- boxplot(X, plot = F)
    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
    ret$stats <- bp$stats
    ret$out <- bp$out
    ret
  } )

  pdf(paste0(pathFigures, "/Figure_SD-light.pdf"), 8, 10)
  par(mar = rep(0, 4), oma = c(5, 6, 1, 1), las = 1, mfrow = c(2, 1))
  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)

  mtext(1, text = "Ellenberg light indicator value", outer = T, cex = 2, line = 3)
  mtext(2, text = expression(paste("Stomatal density (no. ", mu, m^-2, ") [log scale]")), 
        outer = T, cex = 2, las = 0, line = 3)
  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))

  # lines
  abline(v = 3:9, col = "grey")

  # adaxial
  for (i in 3:9) {

    adDns <- dnsAd[[sprintf("%s", i)]]

    # scale density for plotting  
    adDns$d <- adDns$y * 0.4 / max(adDns$y)

    # polygon
    polygon(c(i + adDns$d, rev(i - adDns$d)), c(adDns$x, rev(adDns$x)), col = "grey")

    # outliers
    points(rep(i, length(adDns$out)), adDns$out, pch = 19)

    # bars (0.25 and 0.75 quantiles?)
    lines(rep(i, 2), adDns$stats[c(2, 4), 1], lwd = 5, lend = 1)

    # median
    points(i, adDns$stats[3, 1], pch = 19, cex = 1.5)

    # labels
    s <- "A. adaxial (upper)"
    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)

  }

  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)

  axis(1, at = 3:9)
  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))

  # lines
  abline(v = 3:9, col = "grey")

  # abaxial
  for (i in 3:9) {

    abDns <- dnsAb[[sprintf("%s", i)]]

    # scale density for plotting  
    abDns$d <- abDns$y * 0.4 / max(abDns$y)

    # polygons
    polygon(c(i - abDns$d, rev(i + abDns$d)), c(abDns$x, rev(abDns$x)), col = "grey")

    # outliers
    points(rep(i, length(adDns$out)), adDns$out, pch = 19)

    # bars (0.25 and 0.75 quantiles?)
    lines(rep(i, 2), abDns$stats[c(2, 4), 1], lwd = 5, lend = 1)

    # median
    points(i, abDns$stats[3, 1], pch = 19, cex = 1.5)

    # labels
    s <- "B. abaxial (lower)"
    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)

  }

  dev.off()

  rm("abDns", "adDns", "dnsAb", "dnsAd", "i", "lf", "s")

@

<<Phylogenetic structural equation model, eval=TRUE, echo=FALSE, results=hide>>=

  # small data.frame for Rphylopars
  td <- data.frame(species = stomata$species)
  td$logSDab <- log(stomata$ab_density + 1)
  td$logSDad <- log(stomata$ad_density + 1)
  td$ellenberg_light <- stomata$ellenberg_light

  # ppOU <- phylopars(td, modAngioPhy, model = "mvOU")
  # saveRDS(ppOU, paste0(pathR, "/ppOU.rds"))
  ppOU <- readRDS(paste0(pathR, "/ppOU.rds"))
  phyMat <- ppOU$pars$phylocov

  # define (co)variances for use in text
  varSDab <- phyMat["logSDab", "logSDab"]
  varSDad <- phyMat["logSDad", "logSDad"]
  covSD <- phyMat["logSDab", "logSDad"]
  varSR <- varSDab + varSDad - 2 * covSD

  ppBM <- phylopars(td, modAngioPhy, model = "BM")

  # AIC comparison of OU and BM
  AIC(ppOU, ppBM)

  # Strucutural equation model using lavaan
  model <- 'logSDab ~ ellenberg_light
            logSDad ~ ellenberg_light'

  fitSEM <- sem(model, sample.cov = ppOU$pars$phylocov, sample.nobs = nrow(td))

  # Calculate p-values based on z-scores
  pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, log = TRUE)
  names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)

  rm("td")

@

<<COMPADRE, eval=TRUE, echo=FALSE, results=hide>>=

compadre <- read.csv(paste0(pathRawData, "/Demographics - Feb 17 2017.csv"),
                     stringsAsFactors = FALSE)
compadre$SpeciesAccepted %<>% str_replace_all(" ", "_")
compadre %<>% rename(species = SpeciesAccepted)
compadre %<>% inner_join(stomata, by = "species")

plot(compadre$T[compadre$T > 0], compadre$sr_even[compadre$T > 0])
plot(compadre$T[compadre$T > 0], compadre$sr_propAd[compadre$T > 0])
cor.test(compadre$Efec, compadre$sr_even)

@

%--------------------------------------------------
% Results
%--------------------------------------------------

\section*{RESULTS}

\subsection*{Light tolerance varies among Raunki\ae r life forms}

Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}). Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).

\subsection*{Interactions between light and Raunki\ae r life form determine stomatal ratio}

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = \Sexpr{pInteraction}$).

\subsection*{Adaxial stomatal density contributes most of the variation in stomatal ratio}

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are \Sexpr{round(varSDad / varSR, 2)}, \Sexpr{round(varSDab / varSR, 2)}, and \Sexpr{round(-2 * covSD / varSR, 2)}, respectively. Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ \Sexpr{pval2latex(exp(pvalSEM["logSDad ~ ellenberg_light"]))} and \Sexpr{pval2latex(exp(pvalSEM["logSDab ~ ellenberg_light"]))}, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was \Sexpr{round(fitSEM@ParTable$est[2] / fitSEM@ParTable$est[1], 1)}$\times$ that of \el~on $\mathrm{sd_{ab}}$ (\Sexpr{round(fitSEM@ParTable$est[2], 2)} versus \Sexpr{round(fitSEM@ParTable$est[1], 2)}). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a \Sexpr{round(exp(fitSEM@ParTable$est[2]), 2)}-fold change in adaxial stomatal density versus a \Sexpr{round(exp(fitSEM@ParTable$est[1]), 2)}-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $\Sexpr{pval2latex(exp(pvalSEM["logSDab ~~ logSDad"]))}). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.

% NOTES FOR CALCULATING fold-chage
% log(sd1) = b1 * 3
% SD1 = exp(b1) ^ 3
% SD1 = exp(b1) ^ 4
% change from 3 to 4 is exp(b1)
% for SD2, change from 3 to 4 is exp(b2)

%--------------------------------------------------
% Discussion
%--------------------------------------------------

\section*{DISCUSSION}

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are necessarily dominated by herbs -- or the fact that species are not independent because of shared evolutionaey hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

\subsection*{Adaptive significance of amphistomy}

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies \textit{and} photosynthesis strongly limits fitness. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g. containing both annual and perennial forms). I predict that amphistomy will be favored much more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}. Since evaporative demand increases under high insolation, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because greater leaf wetness and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

\subsection*{Amphistomy as a proxy for open habitat}

These observations from the British flora strongly support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015}, but also point out previously unknown subtleties. These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$). Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantiative variation. Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'. A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

The nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (most tall trees) are almost always hypostomatous (see also \cite{Muir_2015}). Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction.

% phylo half-life estimate from Muir 2015 is 22my

\subsection*{Why does adaxial stomatal density control stomatal ratio?}

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domesication \citep{Milla_etal_2013}; crops species tend to have higher adaxial stomatal density than their wild relatives. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses.

There are at least two hypotheses that could explain why adaxial stomatal density is the most responsive. The first I refer to as the `real estate' hypothesis. In hypostomatous plants, the lower surface is already crowded with stomata, and hence plants must increase the real estate available for stomata by develping them on the upper surface whenever there is selection for greater stomatal density. When stomata are packed too densely on one surface, stomatal interference limits their functioning and hence may create a strong selective pressure for amphistomy \citep{Parlange_Waggoner_1970, Dow_etal_2014a}.

I refer to the second hypothesis as the `coordination' hypothesis. In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another. For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll. Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations. Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio.

% alternatives are different benefits and costs
% leaf orientation \citep{Smith_etal_1998}, stomatal interference , leaf hydraulic conductance (Sack, Buckley, de Boer?), change in stomatal ratio may be important to CO2 in amphi species (Vaccinium myrttillus, \citep{Woodward_Bazzaz_1988})

\subsection*{Conclusions - finish when analysis is complete}

% From previous paper:
% Adaptive evolution may be constrained if traits cannot evolve independently. In particular, if traits share developmental pathways, then they may be unable to respond differentially to selection. In this study, we examined whether stomata on the abaxial (lower) and adaxial (upper) surfaces can evolve independently. We adduce two new lines of evidence which suggest that stomatal function on each surface can readily respond to selection. First, species possess heritable variation that allows partially independent evolution of stomatal densities in response to selection; every study reviewed found loci which alter stomatal density on one surface but not the other. Second, the anatomical trait most closely connected to stomatal conductance, stomatal pore index \citep{Sack_etal_2003}, evolves independently on ab- and adaxial surfaces among wild tomato species. Together, these new lines of evidence demonstrate that natural selection on stomatal arrangement is not strongly constrained by development, although we lacked statistical power to detect weak constraint. It is therefore likely that variation in how stomatal conductance is partitioned between leaf surfaces is due to adaptive rather than nonadaptive forces.

% Indeed, much recent evidence indicates that selection finely tunes the ratio of stomata on the upper and lower leaf surface, although the adaptive significance of variation in stomatal ratio is unresolved. Stomatal ratio affects leaf function, increasing CO$_2$ diffusion \citep{Parkhurst_1978, Parkhurst_Mott_1990, Gutschick_1984b, Parkhurst_1994} and hydraulic conductance outside the xylem \citep{Buckley_etal_2015}. As predicted, amphistomy seems to be more common in circumstances when efficient CO$_2$ supply is important, such as high irradiance \citep{Mott_etal_1982, Gibson_1996, Smith_etal_1997, Jordan_etal_2014}, thick leaves \citep{Parkhurst_1978, Muir_etal_2014b}, herbaceous growth form \citep{Salisbury_1927, Muir_2015}, and domestication \citep{Milla_etal_2013}. Despite potential benefits of amphistomy, most plant species are hypostomatous, implying a fitness cost of upper stomata, such as increased infection by foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. For example, `upside-down' (resupinate) leaves with the abaxial surface facing upward have re-evolved hypostomy \citep{Lyshede_2002}, strongly implying a cost of upward facing stomata.

% To optimally balance fitness costs and benefits, natural selection must be able to change stomatal traits on one surface independently of the other. The present study shows that this is likely true and strikingly consistent on micro- and macroevolutionary timescales. Among \textit{Populus trichocarpa} populations and \textit{Solanum} species, the ratio of adaxial to abaxial $\mathrm{SPI}$ ($\mathrm{SPI}$ ratio) evolves mostly by changes in stomatal density rather than guard cell size. Within \textit{Populus}, populations are more amphistomatous at Northern latitudes with shorter growing seasons that may select for faster carbon assimilation \citep{Mckown_etal_2014, Kaluthota_etal_2015}. Latitudinal variation \textit{Populus trichocarpa} is due mostly to adaptive variation in adaxial stomatal density \citep{Mckown_etal_2014, Porth_etal_2015}. Stomatal density rather than size may have responded more readily to selection because there is no genetic covariance between ab- and adaxial stomatal density, permitting independent evolution \citep{Porth_etal_2015}. In contrast ab- and adaxial guard cell length positively covary, likely constraining evolution. Similarly, we found that over macroevolutionary timescales most of the variation in $\mathrm{SPI}$ among wild tomato species is due to changes in adaxial stomatal density rather than size. Indeed, stomatal density on each surface evolved independently, whereas guard cell lengths positively covaried (Table~\ref{table:Table2}). Adaptive evolution will likely take advantage of traits that evolve independently because this minimizes antagonistic pleiotropy. In a previous study, we found that loci affecting adaxial stomatal density were likely fixed by selection, but we did not measure stomatal size \citep{Muir_etal_2014a}. Overall, patterns within and between species indicate that selection on $\mathrm{SPI}$ ratio leads to greater change in stomatal densities rather sizes on each surface. Based on the analysis here, we conclude that changing stomatal density on one surface incurs less cost than changing size because the former is less constrained by shared developmental pathways.

% However, simulations show that model identification (e.g. Brownian motion versus Ornstein-Uhlenbeck) is usually correct, even when sample sizes are moderate \citep{Cressler_etal_2015, Ho_Ane_2014}.

\clearpage

%--------------------------------------------------
% Literature Cited
%--------------------------------------------------

\bibliography{refs}
\bibliographystyle{evolution}

\clearpage

%--------------------------------------------------
% Tables
%--------------------------------------------------

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform shape stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & \Sexpr{round(fitSRa$optpar, 2)}
  & \Sexpr{round(fitSRa$sigma2, 3)}
  & \Sexpr{round(cor(fitSRa$fitted.values, fitSRa$fitted.values + fitSRa$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRa)) + 2}
  & \Sexpr{round(fitSRa$logLik, 1)}
  & \Sexpr{round(fitSRa$aic, 1)}
  & \Sexpr{round(fitSRa$aic - min(aicSR), 1)} \\
    \el~+ lifeform
  & \Sexpr{round(fitSRb$optpar, 2)}
  & \Sexpr{round(fitSRb$sigma2, 3)}
  & \Sexpr{round(cor(fitSRb$fitted.values, fitSRb$fitted.values + fitSRb$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRb)) + 2}
  & \Sexpr{round(fitSRb$logLik, 1)}
  & \Sexpr{round(fitSRb$aic, 1)}
  & \Sexpr{round(fitSRb$aic - min(aicSR), 1)} \\
    \el
  & \Sexpr{round(fitSRc$optpar, 2)}
  & \Sexpr{round(fitSRc$sigma2, 3)}
  & \Sexpr{round(cor(fitSRc$fitted.values, fitSRc$fitted.values + fitSRc$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRc)) + 2}
  & \Sexpr{round(fitSRc$logLik, 1)}
  & \Sexpr{round(fitSRc$aic, 1)}
  & \Sexpr{round(fitSRc$aic - min(aicSR), 1)} \\
    lifeform
  & \Sexpr{round(fitSRd$optpar, 2)}
  & \Sexpr{round(fitSRd$sigma2, 3)}
  & \Sexpr{round(cor(fitSRd$fitted.values, fitSRd$fitted.values + fitSRd$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRd)) + 2}
  & \Sexpr{round(fitSRd$logLik, 1)}
  & \Sexpr{round(fitSRd$aic, 1)}
  & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
    1
  & \Sexpr{round(fitSRe$optpar, 2)}
  & \Sexpr{round(fitSRe$sigma2, 3)}
  & \Sexpr{0}
  & \Sexpr{length(coef(fitSRe)) + 2}
  & \Sexpr{round(fitSRe$logLik, 1)}
  & \Sexpr{round(fitSRe$aic, 1)}
  & \Sexpr{round(fitSRe$aic - min(aicSR), 1)} \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

%--------------------------------------------------
% Figures
%--------------------------------------------------

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_lf-light.pdf}}
\caption{Lifeforms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:lf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SRmultReg.pdf}}
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visial clarity.} 
\label{fig:SRmultReg}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SD-light.pdf}}
\caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis). The violin plot shows stomatal density ($y$-axis, log-scale) as a function of \el. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Points outside the polygons are statistical outliers.} 
\label{fig:SD-light}
\end{figure}

\clearpage

%--------------------------------------------------
% Supporting Information
%--------------------------------------------------

\section*{Supporting Information}

% Modify and restart table/figure numbering for appendixes
\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\renewcommand\theequation{S\arabic{equation}}    
\setcounter{table}{0}    
\setcounter{equation}{0}
\setcounter{figure}{0}

%--------------------------------------------------
% Supporting Figures
%--------------------------------------------------

\begin{figure}[ht]
\centerline{\includegraphics{Figures/FigureS_violin.pdf}}
\caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$). The violin plot shows stomatal ratio as a function of Raunki\ae r lifeform. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Sample sizes per lifeform in the dataset are given above the upper plot margin. $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
\label{fig:violin}
\end{figure}

\end{document}

### 2017-07-09 ms/ms.tex
put directory on github

--- 2017-07-09 ms/ms.Rnw
@@
-% Notes from call with Rob
-% van Tienderen - sensitivity approach better for evolution
-% send list of species to rob
-
 % TASKS
@@
 % in methods - mention stuff about Salisbury
-% leaf size
 
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{ms-concordance}
 
@@
 
-<<Preliminaries, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # REMOVE BEFORE PUBLICATION
-  system('cp ~/Google\\ Drive/CommonResources/refs.bib ~/Google\\ Drive/StomatalRatio/BEF/ms/refs.bib')
-
-  # Directories
-  pathMain <- "~/Google Drive/StomatalRatio/BEF"
-  pathRawData <- paste0(pathMain, "/RawData")
-  pathProcData <- paste0(pathMain, "/ProcData")
-  pathR <- paste0(pathMain, "/R")
-  pathMS <- paste0(pathMain, "/ms")
-  pathFigures <- paste0(pathMS, "/Figures")
-
-  # Libraries
-  library(vioplot)
-  library(ape)
-  library(rncl)
-  library(phytools)
-  library(rstanarm)
-  library(phylolm)
-  library(Rphylopars)
-  library(lavaan)
-  library(stringr)
-  library(magrittr)
-  library(taxize)
-  library(dplyr)
-
-  # Source functions
-  source(paste0(pathR, "/functions.R"))
-
-@
-
-<<Process data, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Stomata
-  ## BEF
-  stomata <- read.csv(paste0(pathRawData, "/stomata.csv"), stringsAsFactors = F, 
-                      strip.white = T)
-
-  ## Salisbury 1927
-  salisbury <- read.csv(paste0(pathRawData, "/salisbury_1927.csv"), 
-                        stringsAsFactors = FALSE, strip.white = TRUE)
-
-  # This is done as a loop to prevent timing out
-  # tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
-  # for (i in 2:nrow(salisbury)) {
-  #  tnr %<>% rbind(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
-  # }
-  # write.csv(temp, file = paste0(pathRawData, "/taxized_salisbury_1927.csv"), row.names = F)
-  tnr <- read.csv(paste0(pathRawData, "/taxized_salisbury_1927.csv"), stringsAsFactors = FALSE)
-
-  # When TNRS does not give an accepted name, I am using these based on searching The Plant List
-  # http://www.theplantlist.org/
-  # Plant List corrections
-
-  old_name <- c("Betula alba",            "Betula pubescens",         "Fagus sylvatica",
-                "Ulmus montana",          "Corylus avellana",         "Crataegus oxyacanthoides",
-                "Rubus corylifolius",     "Cephalanthera pallens",    "Conopodium denudatum",
-                "Helleborus foetidus",    "Primula acaulis",          "Byronia dioica",
-                "Geranium phaeum",        "Geranium striatum",        "Hieracium sylvaticum",
-                "Hypericum pulchrum",     "Lathyrus macrorrhizus",    "Luzula forsteri",
-                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
-                "Tilia parvifolia",       "Tilia vulgaris",           "Vicia sylvatica",
-                "Iris foetidissima",      "Arrhenatherum avenaceum",  "Hordeum sylvaticum")
-  new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylvatica",
-                "Ulmus glabra",           "Corylus avellana",         "Crataegus laevigata",
-                "Rubus plicatus",         "Cephalanthera longifolia", "Conopodium majus",
-                "Helleborus foetidus",    "Primula vulgaris",         "Byronia dioica",
-                "Geranium phaeum",        "Geranium versicolor",      "Hieracium murorum",
-                "Hypericum pulchrum",     "Lathyrus linifolius",      "Luzula forsteri",
-                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
-                "Tilia cordata",          "Tilia x europaea",         "Vicia sylvatica",
-                "Xyridion foetidissimum", "Arrhenatherum elatius",    "Hordelymus europaeus")
-
-  tnr$acceptedname[match(old_name, tnr$submittedname)] <- new_name
-  tnr %<>% mutate(species = submittedname)
-
-  salisbury %<>% inner_join(tnr, by = "species")
-  rm(tnr)
-
-  # Find which Salisbury data are already in BEF data
-  already_in <- which(salisbury$submittedname %in% stomata$species |
-                        salisbury$acceptedname %in% stomata$species)
-  x1 <- salisbury$submittedname %in% stomata$species
-  salisbury$species[x1] <- salisbury$submittedname[x1]
-  x2 <- (salisbury$acceptedname %in% stomata$species) & !x1
-  salisbury$species[x2] <- salisbury$acceptedname[x2]
-
-  tmp <- inner_join(stomata, salisbury, by = "species")
-
-  # Overlapping data from BEF and Salisbury are clearly taken from Salisbury, but with some errors. Hence, when they do not match, I have use the Salisbury data I entered myself.
-  with(tmp, plot(ab_density.x, ab_density.y))
-  with(tmp, plot(ad_density.x, ad_density.y))
-
-  x <- tmp %>%
-    use_series(ab_density.x) %>%
-    equals(tmp %>% use_series(ab_density.y)) %>%
-    not() %>%
-    which()
-
-  tmp$ab_density.x[x] <- tmp$ab_density.y[x]
-
-  x <- tmp %>%
-    use_series(ad_density.x) %>%
-    equals(tmp %>% use_series(ad_density.y)) %>%
-    not() %>%
-    which()
-
-  tmp$ad_density.x[x] <- tmp$ad_density.y[x]
-
-  tmp %<>% 
-    select(species, acceptedname, ab_density.x, ad_density.x) %>%
-    set_colnames(c("species", "acceptedname", "ab_density", "ad_density"))
-
-  stomata %<>% 
-    left_join(tmp, by = c("species", "ab_density", "ad_density"))
-
-  # Add data from Salisbury not in BEF
-  stomata %<>% 
-    rbind(salisbury[!(x1 | x2), colnames(.)])
-
-  # Remove duplicates
-  stomata <- stomata[!duplicated(stomata$species), ]
-
-  # Calculate stomatal ratio
-  get_sr_propAd <- function(X) {
-    as.numeric(X["ad_density"]) / (as.numeric(X["ab_density"]) + as.numeric(X["ad_density"]))
-  }
-  get_sr_even <- function(X) {
-    min(as.numeric(X[c("ab_density", "ad_density")])) / 
-      max(as.numeric(X[c("ab_density", "ad_density")]))
-  }
-  stomata %<>% mutate(sr_propAd = apply(., 1, get_sr_propAd),
-                      sr_even = apply(., 1, get_sr_even))
-
-  stomata <- stomata[which(!is.na(stomata$sr_propAd)), ]
-
-  # Photosynthetic pathway
-  photo <- read.csv(paste0(pathRawData, "/photo.csv"), stringsAsFactors = F, 
-                    strip.white = T)
-
-  # All missing taxa are C3 
-  # Not C4 based on Sage et al 2011 (JXB 62:9 3155-3169)
-  # Not CAM based on photosynthetic pathway of congeneric species in the data
-  missing_photo <- 
-    c("Alchemilla vulgaris",  "Asarum europaeum", "Byronia dioica",
-      "Hieracium sylvaticum", "Luzula maxima",    "Prunus cerasus",
-      "Pteridium aquilinum",  "Pyrus communis",   "Pyrus torminalis",
-      "Quercus sessiliflora", "Rhamnus frangula", "Ribes grossularia",
-      "Rubus corylifolius",   "Tilia vulgaris")
-  photo %<>% rbind(data.frame(species = missing_photo, photo = "c3"))
-  all(stomata$species %in% photo$species |
-        stomata$acceptedname %in% photo$species) # should be true
-
-  # PLANTATT (Hill et al 2004) for lifeform and Ellenberg light indicator values
-  plantatt <- read.csv(paste0(pathRawData, "/plantatt.csv"), stringsAsFactors = F, 
-                       strip.white = T)
-  # stomata$species[!(stomata$species %in% plantatt$Taxon.name |
-  #                    stomata$acceptedname %in% plantatt$Taxon.name)] # Missing data for several
-
-  # Combine bulbous and nonbulbous geophytes
-  plantatt$LF1[plantatt$LF1 == "Gb"] <- "Gn"
-
-  # Combine hydrophytes and annual hydrophytes
-  plantatt$LF1[plantatt$LF1 == "Hz"] <- "Hy"
-
-  # Combine phanerophytes and nanophanerophytes
-  plantatt$LF1[plantatt$LF1 == "Pn"] <- "Ph"
-
-  ##### MESSING AROUND --------------------------------------------
-  # Leaf size
-  leaf_size <- read.csv(paste0(pathRawData, "/leaf_size.csv"), stringsAsFactors = F)
-  leaf_size$leaf_size %<>% ordered(levels = c("<0.1", "0.1-1", "1-10", "10-100", "100-1000",
-  																						">1000"))
-  leaf_size <- read.csv(paste0(pathRawData, "/leda_leaf_area.csv"), stringsAsFactors = F,
-  											strip.white = TRUE)
-  leaf_size %<>% rename(species = SBS.name, leaf_size = single.value..mm.2.)
-  leaf_size <- data.frame(leaf_size = tapply(leaf_size$leaf_size, leaf_size$species, mean))
-  leaf_size$species <- rownames(leaf_size)
-
-  seed_number <- read.csv(paste0(pathRawData, "/leda_seed_number.csv"), stringsAsFactors = F,
-  											strip.white = TRUE)
-  seed_number %<>% rename(species = SBS.name, seed_number = single.value)
-  seed_number <- data.frame(seed_number = tapply(seed_number$seed_number, seed_number$species, mean))
-  seed_number$species <- rownames(seed_number)
-  which(!(stomata$species %in% seed_number$species |
-  			stomata$acceptedname %in% seed_number$species))
-
-  #####-----------------------------
-
-  # Many missing leaf size
-  which(!(stomata$species %in% leaf_size$species |
-  			stomata$acceptedname %in% leaf_size$species))
-
-  # Combine datasets
-  xp <- match(stomata$species, photo$species)
-  xpa <- match(stomata$acceptedname, photo$species)
-  xp[is.na(xp)] <- xpa[is.na(xp)]
-
-  xa <- match(stomata$species, plantatt$Taxon.name)
-  xaa <- match(stomata$acceptedname, plantatt$Taxon.name)
-  xa[is.na(xa)] <- xaa[is.na(xa)]
-
-  xl <- match(stomata$species, leaf_size$species)
-  xla <- match(stomata$acceptedname, leaf_size$species)
-  xl[is.na(xl)] <- xla[is.na(xl)]
-
-  xs <- match(stomata$species, seed_number$species)
-  xsa <- match(stomata$acceptedname, seed_number$species)
-  xs[is.na(xs)] <- xsa[is.na(xs)]
-
-  stomata$photo <- photo$photo[xp]
-  stomata$lifeform <- plantatt$LF1[xa]
-  stomata$ellenberg_light <- plantatt$L[xa]
-  stomata$height <- plantatt$Hght[xa]
-  stomata$wh <- plantatt$W[xa]
-  stomata$leaf_size <- leaf_size$leaf_size[xl]
-  stomata$seed_number <- seed_number$seed_number[xs]
-
-  # Remove missing values
-  stomata <- stomata[complete.cases(stomata[, c("sr_propAd", "lifeform", "ellenberg_light")]), ]
-
-  rm("already_in", "missing_photo", "new_name", "old_name", "photo", 
-     "plantatt", "salisbury", "tmp", "x", "x1", "x2", "xp", "xpa", "xa", "xaa")
-
-  # Export  
-  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
-
-@
-
-<<Leaf size, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # boxplot(sr_even ~ leaf_size, data = stomata)
-  # boxplot(ellenberg_light ~ leaf_size, data = stomata)
-
-  plot(sr_even ~ leaf_size, data = stomata, log = "x")
-  boxplot(log10(leaf_size) ~ ellenberg_light, data = stomata)
-  tmp <- read.csv(paste0(pathRawData, "/try_leaf_area.csv"), stringsAsFactors = FALSE,
-  								strip.white = TRUE)
-
-  stomata$species %in% tmp$AccSpeciesName | stomata$acceptedname %in% tmp$AccSpeciesName
- sort(table(tmp[which(tmp$AccSpeciesName %in% stomata$species), c("Dataset")]))
-@
-
-<<Phylogeny, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Import phylogeny of British Flora from Lim et al 2014
-  tmp <- readLines(paste0(pathRawData, "/S15105.nex"))
-  l1 <- grep("BEGIN CHARACTERS", tmp) - 1
-  l2 <- grep("BEGIN TREES", tmp) - 1
-  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathRawData, "/Lim_etal_2014.nex"))
-  phy <- read_nexus_phylo(paste0(pathRawData, "/Lim_etal_2014.nex"))
-
-  # Check for species missing from phylogeny
-  stomata$species %<>% str_replace_all(" ", "_")
-  stomata$acceptedname %<>% str_replace_all(" ", "_")
-  notInPhy <- which(!(stomata$species %in% phy$tip.label |
-                        stomata$acceptedname %in% phy$tip.label))
-  stomata[notInPhy, ]; length(notInPhy)
-
-  # Modify tip labels
-
-  # Using Cerastium brachypetalum as proxy for C. cerastoides based on similar divergence in Scheen et al. 2004 (Am J Bot, 91(6): 943-952)
-  tmp <- gsub("Cerastium_brachypetalum", "Cerastium_cerastoides", tmp)
-
-  # Using Coincya monensis as proxy for C. wrightii 
-  tmp <- gsub("Coincya_monensis", "Coincya_wrightii", tmp)
-
-  # Using Geranium pusillum as proxy for G. versicolor based on similar divergence in Palazzesi et al. 2012 (Bio J of Linn Soc, 107(1): 67-85)
-  # tmp <- gsub("Geranium_pusillum", "Geranium_versicolor", tmp)
-
-  # Using Gnaphalium luteoalbum as proxy for G. norvegicum 
-  tmp <- gsub("Gnaphalium_luteoalbum", "Gnaphalium_norvegicum", tmp)
-
-  # Lithospermum purpurocaeruleum is mispelled
-  tmp <- gsub("Lithospermum_purpureocaeruleum", "Lithospermum_purpurocaeruleum", tmp)
-
-  # Using Lythrum portula as proxy for L. hyssopifolia 
-  tmp <- gsub("Lythrum_portula", "Lythrum_hyssopifolia", tmp)
-
-  # Using Peucedanum ostruthium as proxy for P. officinale 
-  tmp <- gsub("Peucedanum_ostruthium", "Peucedanum_officinale", tmp)
-
-  # Changing Populus_nigra_sens.lat. to Populus nigra officinale 
-  tmp <- gsub("Populus_nigra_sens.lat.", "Populus_nigra", tmp)
-
-  # Changing Rosa_canina_agg. to Rosa_canina officinale 
-  tmp <- gsub("Rosa_canina_agg.", "Rosa_canina", tmp)
-
-  # Using Stachys alpina as proxy for S. germanica based on similar divergence in Salmaki et al. 2013 (Mol Phylo Evol, 69(3): 535-551)
-  tmp <- gsub("Stachys_alpina", "Stachys_germanica", tmp)
-
-  # Using Tripleurospermum_maritimum as proxy for T. inodorum 
-  tmp <- gsub("Tripleurospermum_maritimum", "Tripleurospermum_inodorum", tmp)
-
-  # Using Viola_arvensis as proxy for V. kitaibeliana 
-  tmp <- gsub("Viola_arvensis", "Viola_kitaibeliana", tmp)
-
-  # Write first modified phylogeny
-  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
-  phy <- read_nexus_phylo(paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
-
-  # Bind additional tips
-  tips <- phy$tip.label
-  nodes <- sapply(tips, function(x,y) which(y == x), y = phy$tip.label)
-  edge.lengths <- setNames(phy$edge.length[sapply(nodes, function(x, y) {
-    which(y == x) }, y = phy$edge[,2])], names(nodes))
-
-  # Make Agrostemma githago sister to all (Silene sp., Lychnis sp.) [based on Fior et al. 2006, Am J Bot 104(2):399-411]
-  n <- getMRCA(phy, phy$tip.label[c(grep("Silene", phy$tip.label), grep("Lychnis", phy$tip.label))])
-
-  # Branch length is average of distance from (Silene sp., Lychnis sp.) MRCA to (Silene sp., Lychnis sp.) sp.
-  el1 <- sapply(nodes[c(grep("Silene_", phy$tip.label), grep("Lychnis_", phy$tip.label))], 
-                nodeheight, tree = phy)
-  el2 <- nodeheight(tree = phy, node = n)
-
-  phy <- bind.tip(phy, "Agrostemma_githago", edge.length = mean(el1 - el2), where = n)
-
-  # Make Heracleum mantegazzianum as sister to H. sphondylium
-  phy <- bind.tip(phy, "Heracleum_mantegazzianum", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Heracleum_sphondylium")],
-                  where = which(phy$tip.label == "Heracleum_sphondylium"))
-
-  # Make Portulaca oleracea sister to all Claytonia (based on Angiosperm Phylogeny Website v 13 http://www.mobot.org/mobot/research/apweb/welcome.html]
-  # n <- getMRCA(phy, phy$tip.label[grep("Claytonia", phy$tip.label)])
-
-  # Branch length is average of distance from Claytonia MRCA to Claytonia sp.
-  # el1 <- sapply(nodes[grep("Claytonia_", names(nodes))], nodeheight, tree = phy)
-  # el2 <- nodeheight(tree = phy, node = n)
-
-  # phy <- bind.tip(phy, "Portulaca_oleracea", edge.length = mean(el1 - el2), where = n)
-
-  # Make Spartina alterniflora, Spartina maritima, and Spartina x townsendii as sister to S. anglica
-  phy <- bind.tip(phy, "Spartina_alterniflora", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-                  where = which(phy$tip.label == "Spartina_anglica"))
-
-  phy <- bind.tip(phy, "Spartina_maritima", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-                  where = which(phy$tip.label == "Spartina_anglica"))
-
-  # phy <- bind.tip(phy, "Spartina_x_townsendii", 
-  #                 edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-  #                 where = which(phy$tip.label == "Spartina_anglica"))
-
-  # Make Tephroseris integrifolia sister to all Senecio sp.
-  n <- getMRCA(phy, phy$tip.label[grep("Senecio", phy$tip.label)])
-
-  # Branch length is average of distance from Senecio MRCA to Senecio sp.
-  el1 <- sapply(nodes[grep("Senecio_", names(nodes))], nodeheight, tree = phy)
-  el2 <- nodeheight(tree = phy, node = n)
-
-  phy <- bind.tip(phy, "Tephroseris_integrifolia", edge.length = mean(el1 - el2), where = n)
-
-  # Make Viola hirta as sister to V. reichenbachiana (both in Viola sect. Viola)
-  phy <- bind.tip(phy, "Viola_hirta", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Viola_reichenbachiana")],
-                  where = which(phy$tip.label == "Viola_reichenbachiana"))
-
-  # Write second modified phylogeny
-  write.nexus(phy, file = paste0(pathProcData, "/Lim_etal_2014_mod2.nex"))
-
-  # Check for species missing from modified phylogeny
-  notInPhy <- which(!(stomata$species %in% phy$tip.label |
-                        stomata$acceptedname %in% phy$tip.label))
-  stomata[notInPhy, ]; length(notInPhy)
-
-  rm("edge.lengths", "el1", "el2", "l1", "l2", "n", "nodes", "notInPhy", "tips", "tmp")
-
-  # Run PATHd8 (Britton et al. 2002)
-  cat(sprintf("Sequence length = 4692;
-
-              %s
-
-              mrca: Azolla_filiculoides, Solanum_nigrum, maxage=454;
-              mrca: Osmunda_regalis, Asplenium_onopteris, minage=354;
-              mrca: Pinus_sylvestris, Solanum_nigrum, minage=310;
-              mrca: Nymphaea_alba, Solanum_nigrum, minage=140;
-              mrca: Nymphaea_alba, Solanum_nigrum, maxage=217;
-              mrca: Nymphaea_alba, Nuphar_lutea, minage=115;
-              mrca: Chelidonium_majus, Solanum_nigrum, fixage=121;
-              mrca: Quercus_robur, Bryonia_dioica, minage=84;
-
-              name of mrca: Azolla_filiculoides, Solanum_nigrum, name=tracheophytes;
-              name of mrca: Osmunda_regalis, Asplenium_onopteris, name=monilophytes;
-              name of mrca: Pinus_sylvestris, Solanum_nigrum, name=seedplants;
-              name of mrca: Nymphaea_alba, Solanum_nigrum, name=angiosperms;
-              name of mrca: Nymphaea_alba, Nuphar_lutea, name=nymphaceae;
-              name of mrca: Chelidonium_majus, Solanum_nigrum, name=eudicots;
-              name of mrca: Quercus_robur, Bryonia_dioica, name=curcibitales_fagales;
-
-              ", write.tree(phy)), file = paste0(pathMain, "/PATHd8/pathd8_in"))
-
-  # Run PATHd8
-  system("cd ~/Google\\ Drive/StomatalRatio/BEF/PATHd8; ./PATHd8 pathd8_in pathd8_out")
-
-  # Import ultrametric tree
-  pathd8_out <- readLines(paste0(pathMain, "/PATHd8/pathd8_out"))
-  phy <- read.tree(text = substr(pathd8_out[11], 14, nchar(pathd8_out[11])))
-
-  rm("pathd8_out")
-
-@
-
-<<Violin plot of Raunkiaer life form versus stomatal ratio, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # chamaephyte = subshrub (woody)
-  # geophyte = cryptophyte resting in dry ground
-  # hemicryptophyte = perennial with overwintering buds at soil surface
-  # hydrophyte = cryptophyte resting in marshy ground
-  # phanerophyte = normally woody perennial
-  # therophyte = annual
-
-  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
-  names(lf) = c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
-  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-  pdf(paste0(pathFigures, "/FigureS_violin.pdf"), 8, 5.25)
-  par(mai = c(1.75, 1.5, 0.25, 0.25), cex.lab = 1.5, las = 1, cex = 1.5, cex.axis = 0.5)
-
-  vioplot(subset(stomata$sr_propAd, stomata$lifeform == names(lf[1]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[2]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[3]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[4]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[5]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[6]), na.rm = TRUE),
-          col = "grey", colMed = "black", names = rep("", 6))
-
-  clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
-       grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
-  # Sample sizes
-  text(1:6 + 0.25, grconvertY(1.5, "inches", "user"), labels = lf, srt = 30, pos = 2)
-  mtext(table(stomata$lifeform)[names(lf)], 3, at = 1:6, line = 0)
-  title(ylab = "Stomatal Ratio",  line = 3.5)
-  title(xlab = "Raunkiaer Life Form", line = 4.5)
-  title(ylab = expression(SR[propAd] == SD[ad] / SD[total]), line = 2, cex.lab = 1) 
-  dev.off()
-
-  rm("lf")
-
-@
-
-<<Remove hydrophytes + c4 + cam + nonangiosperm, eval=TRUE, echo=FALSE, results=hide>>=
-
-  nHydrophyte <- length(which(stomata$lifeform == "Hy"))
-  nC4 <- length(which(stomata$photo == "c4"))
-  nCAM <- length(which(stomata$photo == "cam"))
-
-  x1 <- c("Ph", "Ch", "hc", "Gn", "Th")
-  stomata <- stomata[which(stomata$lifeform %in% x1 & stomata$photo == "c3"), ]
-
-  angioPhy <- extract.clade(phy, 
-                            node = length(phy$tip.label) + which(phy$node.label == "angiosperms"))
-
-  nNonangio <- length(which(!(stomata$species %in% angioPhy$tip.label | 
-                                stomata$acceptedname %in% angioPhy$tip.label)))
-
-  stomata <- stomata[stomata$species %in% angioPhy$tip.label |
-                       stomata$acceptedname %in% angioPhy$tip.label, ]
-
-  # Export new dataset and tree
-  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
-  write.nexus(angioPhy, file = paste0(pathProcData, "/angioPhy.nex"))
-
-  rm("x1", "phy")
-
-@
-
 \section*{Abstract}
@@
 In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on \Sexpr{nrow(stomata)} species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 374 species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
@@
 
-I excluded data from hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig.
+I excluded data from hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig.
 \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
-I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants..
-I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data.
-The final dataset contained \Sexpr{nrow(stomata)} species.
+I also excluded C$_4$ (3 species) and CAM (2 species) plants..
+I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
+The final dataset contained 374 species.
 The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.
@@
 
-<<Raunkiaer life form versus Ellenberg light indicator values, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Make dichotomous and slightly adjust edge lengths to make exactly ultrametric
-  modAngioPhy <- multi2di(angioPhy)
-  modAngioPhy$edge.length[modAngioPhy$edge.length == 0] <- 0.02
-  x <- diag(vcv.phylo(modAngioPhy))
-  is.ultrametric(modAngioPhy)
-  n <- length(modAngioPhy$tip.label)
-  te <- sapply(1:n, function(x, y) which(y == x), y = modAngioPhy$edge[, 2]) # terminal edges
-  modAngioPhy$edge.length[te] <- modAngioPhy$edge.length[te] + (max(x) - x)
-  is.ultrametric(modAngioPhy)
-
-  x <- which(stomata$acceptedname %in% modAngioPhy$tip.label &
-               !(stomata$species %in% modAngioPhy$tip.label))
-  stomata$species[x] <- stomata$acceptedname[x]
-
-  x <- stomata$species[duplicated(stomata$species)]
-  stomata[stomata$species %in% x, ]
-  modAngioPhy <- drop.tip(modAngioPhy, 
-                          modAngioPhy$tip.label[!modAngioPhy$tip.label %in% stomata$species])
-
-  # phylolm shows that there is no phylogenetic signal in this regression (very high alpha)
-  rownames(stomata) <- stomata$species # need to change row names for phylolm
-  fitLFvEL_pl <- phylolm(ellenberg_light ~ -1 + lifeform, model = "OUrandomRoot", data = stomata, 
-                         phy = modAngioPhy, upper.bound = 1e10)
-  summary(fitLFvEL_pl)
-
-  # therefore using standard ANOVA
-  fitLFvEL <- lm(ellenberg_light ~ lifeform, data = stomata)
-  aovLFvEL <- anova(fitLFvEL)
-
-  # for plotting confidence intervals
-  fitLFvEL <- lm(ellenberg_light ~ -1 + lifeform, data = stomata)
-
-  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
-  names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
-  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-  pdf(paste0(pathFigures, "/Figure_lf-light.pdf"), 4, 7)
-  par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
-  plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(0, 5), xlab = "", ylab = "", axes = F)
-
-  for (i in 0:4) abline(h = i + seq(0.2, 0.6, 0.2), col = "grey")
-  hs <- tapply(stomata$ellenberg_light, stomata$lifeform, hist, plot = F, breaks = 0:9)  
-  for (i in 1:5) {
-    rect(0:8, rep(i - 1, 9), 1:9, i - 1 + hs[[names(lf)[i]]]$density, col = "black", 
-         border = "white", lwd = 2)
-  }
-
-  # label life forms
-  s <- sprintf("%s (%s)", lf, table(stomata$lifeform)[names(lf)])
-  text(0 + 0.5 * strwidth(s), 1:5, labels = s, pos = 1)
-
-  # axes
-  axis(1, at = 0.5:8.5, labels = 1:9, lwd = 0, line = -1.5)
-  for (i in 0:4) axis(2, at = i + seq(0.2, 0.6, 0.2), labels = seq(0.2, 0.6, 0.2), lwd = 0, 
-                      lwd.ticks = 1)
-
-  # axis labels
-  mtext(c("Shade\n", "Partial\nShade", "Sun\n"), 1, at = c(1.5, 4.5, 7.5), line = 1.5)
-  title(xlab = "Ellenberg light\nindicator", cex.lab = 1.5, line = 5)
-  title(ylab = "Proportion", cex.lab = 1.5)
-
-  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
-       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
-  rect(grconvertX(0, from = "npc", "user"), 0:4, 
-       grconvertX(1, from = "npc", "user"), 0.75:4.75)
-
-  # Means + 95% confidence intervals
-  ci <- confint(fitLFvEL)
-  ci <- ci[paste0("lifeform", names(lf)), ] # reorder
-  mu <- coef(fitLFvEL)[paste0("lifeform", names(lf))]
-  for (i in 1:5) {
-    lines(ci[i, ], rep(i - 1 + 0.675, 2), lwd = 2)
-    points(mu[i], i - 1 + 0.675, pch = 21, col = "black", bg = "white")
-  }
-
-  dev.off()
-
-  rm("ci", "fitLFvEL_pl", "hs", "i", "mu", "n", "s", "te", "x")
-
-@
-
 I tested whether Raunki\ae r life form was associated \el~values among British angiosperms using ANOVA with Type-2 sum of squares.
@@
 \subsection*{Open habitat, growth form, and stomatal ratio}
-
-<<Raunkiaer life form and Ellenberg light indicator values versus SR_even, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Phylogenetic regression model comparison
-  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRb <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRd <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  aicSR <- sapply(list(fitSRa, fitSRb, fitSRc, fitSRd, fitSRe), AIC)
-
-  # Parametric bootstrap CIs of full model for p-values
-  # set.seed(36502)
-  # fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, model = "OUrandomRoot",
-  #                    data = stomata, phy = modAngioPhy, boot = 1e4, upper.bound = 10)
-  # saveRDS(fitSR_pl, file = paste0(pathR, "/fitSR_pl.rds"))
-  fitSR_pl <- readRDS(file = paste0(pathR, "/fitSR_pl.rds"))
-
-  # Figure
-
-  pdf(paste0(pathFigures, "/Figure_SRmultReg.pdf"), 4, 7)
-  par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
-
-  for (i in 5:1) {
-
-    plot(0, 0, type = "n", xlim = c(1, 9), ylim = c(-0.05, 1.05), xlab = "", ylab = "", axes = F,
-         frame.plot = F)
-    if (i %in% seq(1, 5, 2)) axis(2, at = seq(0, 1, 0.5), lwd = 0, lwd.ticks = 1, las = 1)
-
-    # polygon of confidence intervals
-    x <- seq(min(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]),
-             max(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]), length.out = 1e3)
-    bs <- sapply(x, function(X) {
-      fitSR_pl$bootstrap[, sprintf('lifeform%s', names(lf)[i])] + 
-        fitSR_pl$bootstrap[, sprintf('lifeform%s:ellenberg_light', names(lf)[i])] * X
-    } )
-    ci <- apply(bs, 2, quantile, probs = c(0.025, 0.5, 0.975))
-    polygon(c(x, rev(x)), c(ci[1, ], rev(ci[3, ])), col = "grey", border = NA)
-    points(x, ci[2, ], lwd = 2, type = "l") # median
-    points(x, ci[1, ], lwd = 2, lty = 2, type = "l") # lower 95%
-    points(x, ci[3, ], lwd = 2, lty = 2, type = "l") # upper 95%
-
-    # plot points
-    with(subset(stomata, stomata$lifeform == names(lf)[i]), 
-         points(jitter(ellenberg_light), sr_even, pch = 21, bg = rgb(0, 0, 0, 0.1), cex = 2))
-
-    # label life forms, slope, and statistical significance
-    s1 <- sprintf("%s (%s)", lf[i], table(stomata$lifeform)[names(lf)[i]])
-    text(1 + 0.5 * strwidth(s1), 1.05, labels = s1, pos = 1, adj = 1)
-    slopeCI <- confint(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", names(lf)[i]), ]
-    pval <- 1 - length(which(fitSR_pl$bootstrap[, sprintf("lifeform%s:ellenberg_light", 
-                                                          names(lf)[i])] > 0)) / fitSR_pl$boot 
-    s2 <- bquote(slope == .(round(coef(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", 
-                                                         names(lf)[i])], 3))^.(sigStar(pval)))
-    text(1 + 0.5 * strwidth(s2), 1.05 - 1.5 * strheight(s1), labels = s2, pos = 1, adj = 1)
-
-  }
-
-  # frame plot
-  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
-       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
-  abline(h = grconvertY(0:5, "npc", "user"))
-  lines(rep(grconvertX(0, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))
-  lines(rep(grconvertX(1, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))
-
-  # axis labels
-  axis(1, at = 1:9, labels = 1:9, lwd = 0, line = -1)
-  axis(1, at = c(2, 5, 8), lwd = 0, labels = c("Shade", "Partial Shade", "Sun"),
-       line = 0)
-  mtext(1, text = "Ellenberg light indicator", cex = 1.5, line = 3, outer = T)
-
-  mtext(2, text = "Stomatal Ratio",  line = 4.5, outer = T, cex = 1.5)
-
-  mtext(2, text = expression(
-    SR[even] == plain(min)~group("(", list(SD[ab], SD[ad]), ")") /
-      plain(max)~group("(", list(SD[ab], SD[ad]), ")")), outer = T, line = 2.5)
-
-  dev.off()
-
-  # Parametric bootstrap statistical signicance of deltaAIC between model 
-  # with interaction (fitSRa) and model without interaction (fitSRb)
-  # set.seed(444294)
-  # bootModB <- predict(fitSRb) + rTrait(n = 1e3, modAngioPhy, "OU", 
-  #                                     list(sigma2 = fitSRb$sigma2, alpha = fitSRb$optpar))
-  # colnames(bootModB) <- paste0("b", 1:ncol(bootModB))
-
-  # fit model of model A to data simulated from model B
-  # fitA2B <- lapply(colnames(bootModB), function(col) {
-  #  stomata$Y <- bootModB[, col]
-  #  phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata, 
-  #          phy = modAngioPhy, upper.bound = 10)
-  #  })
-
-  # fit model of model A to data simulated from model B
-  # fitB2B <- lapply(colnames(bootModB), function(col) {
-  #  stomata$Y <- bootModB[, col]
-  #  phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata, 
-  #          phy = modAngioPhy, upper.bound = 10)
-  #  })
-
-  # refit models at parameter bounds
-  # for (i in which(round(sapply(fitA2B, function(X) X$optpar), 2) == 10)) {
-  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
-  #  fitA2B[[i]] <- phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata,
-  #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200, 
-  #                         starting.value = 10)
-  # }
-
-  # refit models at parameter bounds
-  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 10)) {
-  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
-   #  fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
-   #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200,
-   #                        starting.value = 10)
-   # }
-
-  # for some reason, this fixes those with alpha estimated at upper bound
-  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 200)) {
-  #   stomata$Y <- bootModB[, colnames(bootModB)[i]]
-  #   fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
-  #                          phy = modAngioPhy, lower.bound = 0, upper.bound = 300, 
-  #                          starting.value = 1)
-  # }
-
-  # saveRDS(fitA2B, paste0(pathR, "/fitA2B.rds"))
-  # saveRDS(fitB2B, paste0(pathR, "/fitB2B.rds"))
-
-  fitA2B <- readRDS(paste0(pathR, "/fitA2B.rds"))
-  fitB2B <- readRDS(paste0(pathR, "/fitB2B.rds"))
-
-  # Distribution of deltaAIC
-  obsDeltaAIC <- fitSRa$aic - fitSRb$aic
-  distDeltaAIC <- sapply(fitA2B, function(X) X$aic) - sapply(fitB2B, function(X) X$aic)
-  pInteraction <- length(which(obsDeltaAIC > distDeltaAIC)) / 1e3
-
-  rm("bs", "ci", "distDeltaAIC", "fitA2B", "fitB2B", "i", "lf", "obsDeltaAIC", "pval",
-     "s1", "s2", "slopeCI", "x")
-
-@
-
- <<Testing leaf size model, eval=TRUE, echo=FALSE, results=hide>>=
-
-  stomata$leaf_size %<>% replace(. == 0, NA)
-  stomata %<>% mutate(log_leaf_size = log10(leaf_size))
-  tmp <- drop.tip(modAngioPhy, stomata$species[is.na(stomata$leaf_size)])
-
-  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light * log_leaf_size, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
-  fitSRb <- phylolm(sr_even ~ lifeform * ellenberg_light + log_leaf_size, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
-  fitSRc <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
-
-@
 
@@
 
-<<Stomatal density + light + growth form, eval=TRUE, echo=FALSE, results=hide>>=
-
-  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", 
-          "phanerophyte", "therophyte")
-  names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
-  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-  # density of stomatal density by side and L-value
-  dnsAb <- tapply(log10(stomata$ab_density + 1), stomata$ellenberg_light, function(X) {
-    bp <- boxplot(X, plot = F)
-    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
-                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
-    ret$stats <- bp$stats
-    ret$out <- bp$out
-    ret
-  } )
-
-  dnsAd <- tapply(log10(stomata$ad_density + 1), stomata$ellenberg_light, function(X) {
-    bp <- boxplot(X, plot = F)
-    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
-                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
-    ret$stats <- bp$stats
-    ret$out <- bp$out
-    ret
-  } )
-
-  pdf(paste0(pathFigures, "/Figure_SD-light.pdf"), 8, 10)
-  par(mar = rep(0, 4), oma = c(5, 6, 1, 1), las = 1, mfrow = c(2, 1))
-  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
-
-  mtext(1, text = "Ellenberg light indicator value", outer = T, cex = 2, line = 3)
-  mtext(2, text = expression(paste("Stomatal density (no. ", mu, m^-2, ") [log scale]")), 
-        outer = T, cex = 2, las = 0, line = 3)
-  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))
-
-  # lines
-  abline(v = 3:9, col = "grey")
-
-  # adaxial
-  for (i in 3:9) {
-
-    adDns <- dnsAd[[sprintf("%s", i)]]
-
-    # scale density for plotting  
-    adDns$d <- adDns$y * 0.4 / max(adDns$y)
-
-    # polygon
-    polygon(c(i + adDns$d, rev(i - adDns$d)), c(adDns$x, rev(adDns$x)), col = "grey")
-
-    # outliers
-    points(rep(i, length(adDns$out)), adDns$out, pch = 19)
-
-    # bars (0.25 and 0.75 quantiles?)
-    lines(rep(i, 2), adDns$stats[c(2, 4), 1], lwd = 5, lend = 1)
-
-    # median
-    points(i, adDns$stats[3, 1], pch = 19, cex = 1.5)
-
-    # labels
-    s <- "A. adaxial (upper)"
-    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
-         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)
-
-  }
-
-  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
-
-  axis(1, at = 3:9)
-  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))
-
-  # lines
-  abline(v = 3:9, col = "grey")
-
-  # abaxial
-  for (i in 3:9) {
-
-    abDns <- dnsAb[[sprintf("%s", i)]]
-
-    # scale density for plotting  
-    abDns$d <- abDns$y * 0.4 / max(abDns$y)
-
-    # polygons
-    polygon(c(i - abDns$d, rev(i + abDns$d)), c(abDns$x, rev(abDns$x)), col = "grey")
-
-    # outliers
-    points(rep(i, length(adDns$out)), adDns$out, pch = 19)
-
-    # bars (0.25 and 0.75 quantiles?)
-    lines(rep(i, 2), abDns$stats[c(2, 4), 1], lwd = 5, lend = 1)
-
-    # median
-    points(i, abDns$stats[3, 1], pch = 19, cex = 1.5)
-
-    # labels
-    s <- "B. abaxial (lower)"
-    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
-         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)
-
-  }
-
-  dev.off()
-
-  rm("abDns", "adDns", "dnsAb", "dnsAd", "i", "lf", "s")
-
-@
-
-<<Phylogenetic structural equation model, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # small data.frame for Rphylopars
-  td <- data.frame(species = stomata$species)
-  td$logSDab <- log(stomata$ab_density + 1)
-  td$logSDad <- log(stomata$ad_density + 1)
-  td$ellenberg_light <- stomata$ellenberg_light
-
-  # ppOU <- phylopars(td, modAngioPhy, model = "mvOU")
-  # saveRDS(ppOU, paste0(pathR, "/ppOU.rds"))
-  ppOU <- readRDS(paste0(pathR, "/ppOU.rds"))
-  phyMat <- ppOU$pars$phylocov
-
-  # define (co)variances for use in text
-  varSDab <- phyMat["logSDab", "logSDab"]
-  varSDad <- phyMat["logSDad", "logSDad"]
-  covSD <- phyMat["logSDab", "logSDad"]
-  varSR <- varSDab + varSDad - 2 * covSD
-
-  ppBM <- phylopars(td, modAngioPhy, model = "BM")
-
-  # AIC comparison of OU and BM
-  AIC(ppOU, ppBM)
-
-  # Strucutural equation model using lavaan
-  model <- 'logSDab ~ ellenberg_light
-            logSDad ~ ellenberg_light'
-
-  fitSEM <- sem(model, sample.cov = ppOU$pars$phylocov, sample.nobs = nrow(td))
-
-  # Calculate p-values based on z-scores
-  pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, log = TRUE)
-  names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)
-
-  rm("td")
-
-@
-
-<<COMPADRE, eval=TRUE, echo=FALSE, results=hide>>=
-
-compadre <- read.csv(paste0(pathRawData, "/Demographics - Feb 17 2017.csv"),
-                     stringsAsFactors = FALSE)
-compadre$SpeciesAccepted %<>% str_replace_all(" ", "_")
-compadre %<>% rename(species = SpeciesAccepted)
-compadre %<>% inner_join(stomata, by = "species")
-
-plot(compadre$T[compadre$T > 0], compadre$sr_even[compadre$T > 0])
-plot(compadre$T[compadre$T > 0], compadre$sr_propAd[compadre$T > 0])
-cor.test(compadre$Efec, compadre$sr_even)
-
-@
-
 %--------------------------------------------------
@@
 
-Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}).
+Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 369}$ = 18.4, $P$ = 8.1 $\times10^{-14}$).
 Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = \Sexpr{pInteraction}$).
+Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = 0.002$).
 
@@
 Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio.
-The contributions of adaxial density, abaxial density, and their covariance are \Sexpr{round(varSDad / varSR, 2)}, \Sexpr{round(varSDab / varSR, 2)}, and \Sexpr{round(-2 * covSD / varSR, 2)}, respectively.
+The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.39, and -0.53, respectively.
 Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive.
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ \Sexpr{pval2latex(exp(pvalSEM["logSDad ~ ellenberg_light"]))} and \Sexpr{pval2latex(exp(pvalSEM["logSDab ~ ellenberg_light"]))}, respectively).
-However, the regression of \el~on $\mathrm{sd_{ad}}$ was \Sexpr{round(fitSEM@ParTable$est[2] / fitSEM@ParTable$est[1], 1)}$\times$ that of \el~on $\mathrm{sd_{ab}}$ (\Sexpr{round(fitSEM@ParTable$est[2], 2)} versus \Sexpr{round(fitSEM@ParTable$est[1], 2)}).
-Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a \Sexpr{round(exp(fitSEM@ParTable$est[2]), 2)}-fold change in adaxial stomatal density versus a \Sexpr{round(exp(fitSEM@ParTable$est[1]), 2)}-fold change in abaxial stomatal density.
-The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $\Sexpr{pval2latex(exp(pvalSEM["logSDab ~~ logSDad"]))}).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 8.8 $\times10^{-7}$ and 4.9 $\times10^{-5}$, respectively).
+However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.2 versus 0.1).
+Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.22-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
+The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.1 $\times10^{-11}$).
 These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.
@@
 Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio.
-The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$).
+The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = 1.5$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$).
 This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal.
@@
     \el~$\times$ lifeform
-  & \Sexpr{round(fitSRa$optpar, 2)}
-  & \Sexpr{round(fitSRa$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRa$fitted.values, fitSRa$fitted.values + fitSRa$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRa)) + 2}
-  & \Sexpr{round(fitSRa$logLik, 1)}
-  & \Sexpr{round(fitSRa$aic, 1)}
-  & \Sexpr{round(fitSRa$aic - min(aicSR), 1)} \\
+  & 0.46
+  & 0.068
+  & 0.34
+  & 12
+  & -32.6
+  & 89.2
+  & 0 \\
     \el~+ lifeform
-  & \Sexpr{round(fitSRb$optpar, 2)}
-  & \Sexpr{round(fitSRb$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRb$fitted.values, fitSRb$fitted.values + fitSRb$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRb)) + 2}
-  & \Sexpr{round(fitSRb$logLik, 1)}
-  & \Sexpr{round(fitSRb$aic, 1)}
-  & \Sexpr{round(fitSRb$aic - min(aicSR), 1)} \\
+  & 0.47
+  & 0.071
+  & 0.32
+  & 8
+  & -39.8
+  & 95.6
+  & 6.3 \\
     \el
-  & \Sexpr{round(fitSRc$optpar, 2)}
-  & \Sexpr{round(fitSRc$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRc$fitted.values, fitSRc$fitted.values + fitSRc$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRc)) + 2}
-  & \Sexpr{round(fitSRc$logLik, 1)}
-  & \Sexpr{round(fitSRc$aic, 1)}
-  & \Sexpr{round(fitSRc$aic - min(aicSR), 1)} \\
+  & 0.65
+  & 0.109
+  & 0.26
+  & 4
+  & -58.9
+  & 125.9
+  & 36.6 \\
     lifeform
-  & \Sexpr{round(fitSRd$optpar, 2)}
-  & \Sexpr{round(fitSRd$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRd$fitted.values, fitSRd$fitted.values + fitSRd$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRd)) + 2}
-  & \Sexpr{round(fitSRd$logLik, 1)}
-  & \Sexpr{round(fitSRd$aic, 1)}
-  & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
+  & 0.34
+  & 0.066
+  & 0.15
+  & 7
+  & -79.2
+  & 172.4
+  & 83.2 \\
     1
-  & \Sexpr{round(fitSRe$optpar, 2)}
-  & \Sexpr{round(fitSRe$sigma2, 3)}
-  & \Sexpr{0}
-  & \Sexpr{length(coef(fitSRe)) + 2}
-  & \Sexpr{round(fitSRe$logLik, 1)}
-  & \Sexpr{round(fitSRe$aic, 1)}
-  & \Sexpr{round(fitSRe$aic - min(aicSR), 1)} \\
+  & 0.29
+  & 0.068
+  & 0
+  & 3
+  & -107.9
+  & 221.7
+  & 132.5 \\
 	\bottomrule

### 2017-07-09 R/Analysis.R

@@ -1,0 +1,833 @@
+##### Contents #####
+# 1. Preliminaries
+# 2. Process trait data
+# 3. Phylogeny
+# 4. Rauniker life-form versus SR
+# 5. Remove hydrophytes, helophytes, c4, and cam plants
+# #. Plot light versus stomatal ratio
+# #. Plot growth form versus stomatal ratio
+#
+
+##### 1. Preliminaries #####
+
+  # Directories
+  pathMain <- "~/Google Drive/StomatalRatio/BES"
+  pathRawData <- paste0(pathMain, "/RawData")
+  pathProcData <- paste0(pathMain, "/ProcData")
+  pathR <- paste0(pathMain, "/R")
+  pathMS <- paste0(pathMain, "/ms")
+  pathFigures <- paste0(pathMS, "/Figures")
+  
+  setwd(pathMain)
+  
+  # Libraries
+  library(vioplot)
+  library(ape)
+  library(rncl)
+  library(phytools)
+  library(rstanarm)
+  library(phylolm)
+  library(Rphylopars)
+  
+##### 2. Process data #####
+  
+  # Stomata
+  stomata <- read.csv(paste0(pathRawData, "/stomata.csv"), stringsAsFactors = F, strip.white = T)
+  stomata$sr_propAd <- apply(stomata, 1, function(X) as.numeric(X["ad_density"]) / (as.numeric(X["ab_density"]) + as.numeric(X["ad_density"])))
+  stomata$sr_even <- apply(stomata, 1, function(X) min(as.numeric(X[c("ab_density", "ad_density")])) / max(as.numeric(X[c("ab_density", "ad_density")])))
+  stomata <- stomata[which(!is.na(stomata$sr_propAd)), ]
+  
+  # OLD - Ellenberg light index
+  # ellenberg_light <- read.csv(paste0(pathRawData, "/ellenberg_light.csv"), stringsAsFactors = F, 
+  #                            strip.white = T)
+  
+  # OLD - Life-form
+  # lifeform <- read.csv(paste0(pathRawData, "/life-form.csv"), stringsAsFactors = F, strip.white = T)
+  
+  # Photosynthetic pathway
+  photo <- read.csv(paste0(pathRawData, "/photo.csv"), stringsAsFactors = F, strip.white = T)
+  all(stomata$species %in% photo$species) # should be true
+  
+  # PLANTATT (Hill et al 2004) for lifeform and Ellenberg light indicator values
[…]

### 2017-07-09 R/Process PLANTATT (temp code).R

@@ -1,0 +1,21 @@
+library(stringr)
+
+text <- readLines(paste0(pathRawData, "/PLANTATT/test1.txt"))
+
+# Remove " I "
+
+text <- gsub(" I ", "", text)
+
+# Space between columns
+
+text <- gsub("(?<!,) ", "\t", text, perl = TRUE)
+
+# Export
+
+writeLines(text, con = paste0(pathProcData, "/PLANTATT/test1.txt"))
+gregexpr("(?<!,) ", text, perl = TRUE)
+
+str_match_all(text[1], "(?<!,) ")
+word(text[1], start = 1, end = -1)
+
+gregexpr("a", "abc abc")

### 2017-07-09 R/betaRegression.R

@@ -1,0 +1,162 @@
+setwd("~/Google Drive/StomatalRatio/BES")
+library(rstan)
+library(MASS)
+library(boot)
+library(rstanarm)
+
+# Simple linear regression (for understanding)
+dat <- list(N = 1e2, K = 2, x = array(mvrnorm(1e2, c(0, 0), diag(2)), dim = c(100, 2)))
+dat$y <- as.numeric(10 + dat$x %*% array(c(2, 3), dim = c(2, 1)) + rnorm(1e2))
+m1 <- lm(dat$y ~ dat$x[, 1] + dat$x[, 2])
+m1 <- stan(file = "Stan/linearRegression.stan", chains = 1, iter = 2e3, data = dat)
+
+# Beta regression
+# Two groups: group 1 ~ beta(phi = 0.25, lambda = 1); group 1 ~ beta(phi = 0.75, lambda = 1);
+dat <- list(N = 1e3, K = 1)
+dat$x <- array(c(rep(0:1, each = dat$N / 2)), dim = c(dat$N, dat$K))
+phi <- 0.25 + dat$x %*% array(0.5, dim = c(dat$K, 1))
+lambda <- 1
+alpha <- lambda * phi
+beta <- lambda * (1 - phi)
+dat$y <- rbeta(dat$N, alpha, beta)
+m1a <- lm(dat$y ~ dat$x)
+summary(m1a)
+m1b <- stan(file = "Stan/betaRegression.stan", chains = 1, iter = 2e4, data = dat)
+summary(m1b)
+plot(m1b, pars = c("phi[1]", "phi[501]", "lambda"))
+
+# Truncated Beta regression
+# One group: group 1 ~ beta(phi = 0.25, lambda = 1)
+simPar <- list(N = 1e3, L = 0.01, U = 0.99, phi = 0.25, lambda = 1)
+simPar$alpha <- simPar$lambda * simPar$phi
+simPar$beta <- simPar$lambda * (1 - simPar$phi)
+dat <- list(y = rbeta(simPar$N, simPar$alpha, simPar$beta))
+dat$y_obs <- dat$y[dat$y > simPar$L & dat$y < simPar$U]
+dat$N_obs <- length(dat$y_obs)
+dat$N_censL <- length(which(dat$y <= simPar$L))
+dat$N_censU <- length(which(dat$y >= simPar$U))
+dat$L <- simPar$L
[…]

### 2017-07-09 R/functions.R

@@ -1,0 +1,37 @@
+# R functions associated with Muir 2017
+
+# Make statistical significance asterisks
+sigStar <- function(pvalue)
+{
+  if (pvalue >= 0.05) return("n.s.")
+  if (pvalue < 0.05 & pvalue >= 0.01) return("*")
+  if (pvalue < 0.01 & pvalue >= 0.001) return("**")
+  if (pvalue < 0.001) return("***")
+}
+
+# Determine a number's order of magnitude
+oom <- function(x)
+{
+  # x should be a vector of numbers
+  floor(log10(abs(x)))
+}
+
+# Determine number of significant digits to use for rounding in tables
+sigDig <- function(x)
+{
+  # x should be a vector of numeric elements to be rounded to same significant digit
+  ret <- round(x, -min(oom(x)))
+  return(ret)
+}
+
+# Text string for p-value in tables
+pval2latex <- function(x)
+{
+  stem <- x * 10 ^ -oom(x)
+  ret <- ifelse(oom(x) > -3, 
+                sprintf("%.*f", 3, x), 
+                sprintf("%.*f $\\\\times10^{%s}$", 1, stem, oom(x)))
+  return(ret)
+}
+
+

### 2017-07-11 ms/ms.Rnw
updating ms

@@
 
-<<Preliminaries, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # REMOVE BEFORE PUBLICATION
-  system('cp ~/Google\\ Drive/CommonResources/refs.bib ~/Google\\ Drive/StomatalRatio/BEF/ms/refs.bib')
-
-  # Directories
-  pathMain <- "~/Google Drive/StomatalRatio/BEF"
-  pathRawData <- paste0(pathMain, "/RawData")
-  pathProcData <- paste0(pathMain, "/ProcData")
-  pathR <- paste0(pathMain, "/R")
-  pathMS <- paste0(pathMain, "/ms")
-  pathFigures <- paste0(pathMS, "/Figures")
-
-  # Libraries
-  library(vioplot)
-  library(ape)
-  library(rncl)
-  library(phytools)
-  library(rstanarm)
-  library(phylolm)
-  library(Rphylopars)
-  library(lavaan)
-  library(stringr)
-  library(magrittr)
-  library(taxize)
-  library(dplyr)
-
-  # Source functions
-  source(paste0(pathR, "/functions.R"))
-
-@
-
 <<Process data, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Stomata
-  ## BEF
-  stomata <- read.csv(paste0(pathRawData, "/stomata.csv"), stringsAsFactors = F, 
-                      strip.white = T)
-
-  ## Salisbury 1927
-  salisbury <- read.csv(paste0(pathRawData, "/salisbury_1927.csv"), 
-                        stringsAsFactors = FALSE, strip.white = TRUE)
-
-  # This is done as a loop to prevent timing out
-  # tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
-  # for (i in 2:nrow(salisbury)) {
-  #  tnr %<>% rbind(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
-  # }
-  # write.csv(temp, file = paste0(pathRawData, "/taxized_salisbury_1927.csv"), row.names = F)
-  tnr <- read.csv(paste0(pathRawData, "/taxized_salisbury_1927.csv"), stringsAsFactors = FALSE)
 

### 2017-07-11 R/Analysis.R

@@ -45,51 +45,7 @@
                  data = stomata[, c("sr2", "ellenberg_light", "lifeform")], phy = tmp)
   summary(fit)  
 
-##### X. path analysis - experimental #####
-  
-  library("phylopath")
-  candidates <- list(A = DAG(logAb ~ ellenberg_light, logAd ~ ellenberg_light),
-                     B = DAG(logAb ~ ellenberg_light, logAd ~ ellenberg_light, logAd ~ logAb, logAb ~ logAd))
-  plot(candidates$A)
-  
-  library(Rphylopars)
-  
-  stomata$species <- stomata$species1
-  all(stomata$species %in% phy$tip.label) # should be TRUE
-  phy <- chronos(phy1)
-  plot(phy, show.tip.label = F)
-  axis(1)
-  
-  phy1 <- drop.tip(phy, phy$tip.label[!phy$tip.label %in% stomata$species])
-  plot(phy1, show.tip.label = F)
-  ppBM <- phylopars(stomata[, c("species", "ab_density", "ad_density")], tree = phy1,
-                    model = "BM", pheno_error = FALSE, pheno_correlated = FALSE)
-  cov2cor(ppBM$par$phylocov)
-  ppOU <- phylopars(stomata[, c("species", "ab_density", "ad_density")], tree = multi2di(phy1),
-                    model = "mvOU", pheno_error = FALSE, pheno_correlated = FALSE, full_alpha = TRUE)
-  
-  library("mvMORPH")
-  rownames(stomata) <- stomata$species1
-  #phy$edge.length <- phy$edge.length * 100
-  stomata$logAb <- log(stomata$ab_density + 1)
-  stomata$logAd <- log(stomata$ad_density + 1)
-  
-  mmOU <- mvOU(multi2di(phy1), stomata[, c("ab_density", "ad_density")], scale.height = FALSE)
-  mmBM <- mvBM(multi2di(phy1), stomata[, c("ab_density", "ad_density")], scale.height = FALSE)
 
-  mmOU <- mvOU(multi2di(phy1), stomata[, c("logAb", "logAd")], scale.height = FALSE)
-  mmBM <- mvBM(multi2di(phy1), stomata[, c("logAb", "logAd")], scale.height = FALSE)
-  
-##### 3. Plot light versus stomatal ratio #####
-  
-  vioplot(subset(stomata$sr2, stomata$ellenberg_light == 3, na.rm = TRUE),
[…]

### 2017-07-11 R/Process PLANTATT (temp code).R

@@ -1,21 +1,0 @@
-library(stringr)
-
-text <- readLines(paste0(pathRawData, "/PLANTATT/test1.txt"))
-
-# Remove " I "
-
-text <- gsub(" I ", "", text)
-
-# Space between columns
-
-text <- gsub("(?<!,) ", "\t", text, perl = TRUE)
-
-# Export
-
-writeLines(text, con = paste0(pathProcData, "/PLANTATT/test1.txt"))
-gregexpr("(?<!,) ", text, perl = TRUE)
-
-str_match_all(text[1], "(?<!,) ")
-word(text[1], start = 1, end = -1)
-
-gregexpr("a", "abc abc")

### 2017-07-11 R/betaRegression.R

@@ -1,162 +1,0 @@
-setwd("~/Google Drive/StomatalRatio/BES")
-library(rstan)
-library(MASS)
-library(boot)
-library(rstanarm)
-
-# Simple linear regression (for understanding)
-dat <- list(N = 1e2, K = 2, x = array(mvrnorm(1e2, c(0, 0), diag(2)), dim = c(100, 2)))
-dat$y <- as.numeric(10 + dat$x %*% array(c(2, 3), dim = c(2, 1)) + rnorm(1e2))
-m1 <- lm(dat$y ~ dat$x[, 1] + dat$x[, 2])
-m1 <- stan(file = "Stan/linearRegression.stan", chains = 1, iter = 2e3, data = dat)
-
-# Beta regression
-# Two groups: group 1 ~ beta(phi = 0.25, lambda = 1); group 1 ~ beta(phi = 0.75, lambda = 1);
-dat <- list(N = 1e3, K = 1)
-dat$x <- array(c(rep(0:1, each = dat$N / 2)), dim = c(dat$N, dat$K))
-phi <- 0.25 + dat$x %*% array(0.5, dim = c(dat$K, 1))
-lambda <- 1
-alpha <- lambda * phi
-beta <- lambda * (1 - phi)
-dat$y <- rbeta(dat$N, alpha, beta)
-m1a <- lm(dat$y ~ dat$x)
-summary(m1a)
-m1b <- stan(file = "Stan/betaRegression.stan", chains = 1, iter = 2e4, data = dat)
-summary(m1b)
-plot(m1b, pars = c("phi[1]", "phi[501]", "lambda"))
-
-# Truncated Beta regression
-# One group: group 1 ~ beta(phi = 0.25, lambda = 1)
-simPar <- list(N = 1e3, L = 0.01, U = 0.99, phi = 0.25, lambda = 1)
-simPar$alpha <- simPar$lambda * simPar$phi
-simPar$beta <- simPar$lambda * (1 - simPar$phi)
-dat <- list(y = rbeta(simPar$N, simPar$alpha, simPar$beta))
-dat$y_obs <- dat$y[dat$y > simPar$L & dat$y < simPar$U]
-dat$N_obs <- length(dat$y_obs)
-dat$N_censL <- length(which(dat$y <= simPar$L))
-dat$N_censU <- length(which(dat$y >= simPar$U))
-dat$L <- simPar$L
[…]

### 2017-07-11 R/header.R

@@ -20,6 +20,7 @@ library(rncl)
 library(Rphylopars)
 library(rstanarm)
 library(stringr)
+library(taxize)
 library(vioplot)
 
 # Directories

### 2017-07-11 R/install-packages.R

@@ -9,4 +9,5 @@ install.packages("rncl")
 install.packages("Rphylopars")
 install.packages("rstanarm")
 install.packages("stringr")
+install.packages("taxize")
 install.packages("vioplot")

### 2017-07-11 R/process-data.R

@@ -1,8 +1,21 @@
 source("R/header.R")
 
-# Stomata
+# British Ecological Flora
 
 stomata <- read_csv(str_c(path_raw_data, "/stomata.csv"))
+
+# Salisbury 1927
+
+salisbury <- read_csv(str_c(path_raw_data, "/salisbury_1927.csv"))
+
+# Taxize Salisbury dataset - this is done as a loop to prevent timing out
+tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
+for (i in 2:nrow(salisbury)) {
+  tnr %<>% bind_rows(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
+}
+write_csv(tnr, path = str_c(path_raw_data, "/taxized_salisbury_1927.csv"))
+tnr <- read_csv(str_c(path_raw_data, "/taxized_salisbury_1927.csv"))
+############## WORKING TO ADD IN CODE FROM MS.RNW WHERE ###################
 stomata %<>% mutate(sr_propAd = ad_density / (ab_density + ad_density),
                     sr_even = calc_sr_even(ab_density, ad_density))
 stomata %<>% filter(!is.na(sr_propAd))

### 2017-07-11 R/run-all.R

@@ -13,3 +13,6 @@ source("R/modify-phylogeny.R")
 # Analyze data
 source("R/analyse-light-vs-lifeform.R")
 source("R/analyse-stomata.R")
+
+# NEED TO GET PATH ANALYSIS STUFF
+# NEED TO KNIT REPORT

### 2017-07-11 ms/ms.Rnw
incorporated Salisbury data

@@
 % \pagenumbering{gobble}
-
-<<Process data, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # When TNRS does not give an accepted name, I am using these based on searching The Plant List
-  # http://www.theplantlist.org/
-  # Plant List corrections
-
-  old_name <- c("Betula alba",            "Betula pubescens",         "Fagus sylvatica",
-                "Ulmus montana",          "Corylus avellana",         "Crataegus oxyacanthoides",
-                "Rubus corylifolius",     "Cephalanthera pallens",    "Conopodium denudatum",
-                "Helleborus foetidus",    "Primula acaulis",          "Byronia dioica",
-                "Geranium phaeum",        "Geranium striatum",        "Hieracium sylvaticum",
-                "Hypericum pulchrum",     "Lathyrus macrorrhizus",    "Luzula forsteri",
-                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
-                "Tilia parvifolia",       "Tilia vulgaris",           "Vicia sylvatica",
-                "Iris foetidissima",      "Arrhenatherum avenaceum",  "Hordeum sylvaticum")
-  new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylvatica",
-                "Ulmus glabra",           "Corylus avellana",         "Crataegus laevigata",
-                "Rubus plicatus",         "Cephalanthera longifolia", "Conopodium majus",
-                "Helleborus foetidus",    "Primula vulgaris",         "Byronia dioica",
-                "Geranium phaeum",        "Geranium versicolor",      "Hieracium murorum",
-                "Hypericum pulchrum",     "Lathyrus linifolius",      "Luzula forsteri",
-                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
-                "Tilia cordata",          "Tilia x europaea",         "Vicia sylvatica",
-                "Xyridion foetidissimum", "Arrhenatherum elatius",    "Hordelymus europaeus")
-
-  tnr$acceptedname[match(old_name, tnr$submittedname)] <- new_name
-  tnr %<>% mutate(species = submittedname)
-
-  salisbury %<>% inner_join(tnr, by = "species")
-  rm(tnr)
-
-  # Find which Salisbury data are already in BEF data
-  already_in <- which(salisbury$submittedname %in% stomata$species |
-                        salisbury$acceptedname %in% stomata$species)
-  x1 <- salisbury$submittedname %in% stomata$species
-  salisbury$species[x1] <- salisbury$submittedname[x1]
-  x2 <- (salisbury$acceptedname %in% stomata$species) & !x1
-  salisbury$species[x2] <- salisbury$acceptedname[x2]
-
-  tmp <- inner_join(stomata, salisbury, by = "species")
-
-  # Overlapping data from BEF and Salisbury are clearly taken from Salisbury, but with some errors. Hence, when they do not match, I have use the Salisbury data I entered myself.
-  with(tmp, plot(ab_density.x, ab_density.y))
-  with(tmp, plot(ad_density.x, ad_density.y))
-
-  x <- tmp %>%
-    use_series(ab_density.x) %>%
-    equals(tmp %>% use_series(ab_density.y)) %>%
-    not() %>%
-    which()
-
-  tmp$ab_density.x[x] <- tmp$ab_density.y[x]
-
-  x <- tmp %>%
-    use_series(ad_density.x) %>%
-    equals(tmp %>% use_series(ad_density.y)) %>%
-    not() %>%
-    which()
-
-  tmp$ad_density.x[x] <- tmp$ad_density.y[x]
-
-  tmp %<>% 
-    select(species, acceptedname, ab_density.x, ad_density.x) %>%
-    set_colnames(c("species", "acceptedname", "ab_density", "ad_density"))
-
-  stomata %<>% 
-    left_join(tmp, by = c("species", "ab_density", "ad_density"))
-
-  # Add data from Salisbury not in BEF
-  stomata %<>% 
-    rbind(salisbury[!(x1 | x2), colnames(.)])
-
-  # Remove duplicates
-  stomata <- stomata[!duplicated(stomata$species), ]
-
-  # Calculate stomatal ratio
-  get_sr_propAd <- function(X) {
-    as.numeric(X["ad_density"]) / (as.numeric(X["ab_density"]) + as.numeric(X["ad_density"]))
-  }
-  get_sr_even <- function(X) {
-    min(as.numeric(X[c("ab_density", "ad_density")])) / 
-      max(as.numeric(X[c("ab_density", "ad_density")]))
-  }
-  stomata %<>% mutate(sr_propAd = apply(., 1, get_sr_propAd),
-                      sr_even = apply(., 1, get_sr_even))
-
-  stomata <- stomata[which(!is.na(stomata$sr_propAd)), ]
-
-  # Photosynthetic pathway
-  photo <- read.csv(paste0(pathRawData, "/photo.csv"), stringsAsFactors = F, 
-                    strip.white = T)
-
-  # All missing taxa are C3 
-  # Not C4 based on Sage et al 2011 (JXB 62:9 3155-3169)
-  # Not CAM based on photosynthetic pathway of congeneric species in the data
-  missing_photo <- 
-    c("Alchemilla vulgaris",  "Asarum europaeum", "Byronia dioica",
-      "Hieracium sylvaticum", "Luzula maxima",    "Prunus cerasus",
-      "Pteridium aquilinum",  "Pyrus communis",   "Pyrus torminalis",
-      "Quercus sessiliflora", "Rhamnus frangula", "Ribes grossularia",
-      "Rubus corylifolius",   "Tilia vulgaris")
-  photo %<>% rbind(data.frame(species = missing_photo, photo = "c3"))
-  all(stomata$species %in% photo$species |
-        stomata$acceptedname %in% photo$species) # should be true
-
-  # PLANTATT (Hill et al 2004) for lifeform and Ellenberg light indicator values
-  plantatt <- read.csv(paste0(pathRawData, "/plantatt.csv"), stringsAsFactors = F, 
-                       strip.white = T)
-  # stomata$species[!(stomata$species %in% plantatt$Taxon.name |
-  #                    stomata$acceptedname %in% plantatt$Taxon.name)] # Missing data for several
-
-  # Combine bulbous and nonbulbous geophytes
-  plantatt$LF1[plantatt$LF1 == "Gb"] <- "Gn"
-
-  # Combine hydrophytes and annual hydrophytes
-  plantatt$LF1[plantatt$LF1 == "Hz"] <- "Hy"
-
-  # Combine phanerophytes and nanophanerophytes
-  plantatt$LF1[plantatt$LF1 == "Pn"] <- "Ph"
-
-  ##### MESSING AROUND --------------------------------------------
-  # Leaf size
-  leaf_size <- read.csv(paste0(pathRawData, "/leaf_size.csv"), stringsAsFactors = F)
-  leaf_size$leaf_size %<>% ordered(levels = c("<0.1", "0.1-1", "1-10", "10-100", "100-1000",
-  																						">1000"))
-  leaf_size <- read.csv(paste0(pathRawData, "/leda_leaf_area.csv"), stringsAsFactors = F,
-  											strip.white = TRUE)
-  leaf_size %<>% rename(species = SBS.name, leaf_size = single.value..mm.2.)
-  leaf_size <- data.frame(leaf_size = tapply(leaf_size$leaf_size, leaf_size$species, mean))
-  leaf_size$species <- rownames(leaf_size)
-
-  seed_number <- read.csv(paste0(pathRawData, "/leda_seed_number.csv"), stringsAsFactors = F,
-  											strip.white = TRUE)
-  seed_number %<>% rename(species = SBS.name, seed_number = single.value)
-  seed_number <- data.frame(seed_number = tapply(seed_number$seed_number, seed_number$species, mean))
-  seed_number$species <- rownames(seed_number)
-  which(!(stomata$species %in% seed_number$species |
-  			stomata$acceptedname %in% seed_number$species))
-
-  #####-----------------------------
-
-  # Many missing leaf size
-  which(!(stomata$species %in% leaf_size$species |
-  			stomata$acceptedname %in% leaf_size$species))
-
-  # Combine datasets
-  xp <- match(stomata$species, photo$species)
-  xpa <- match(stomata$acceptedname, photo$species)
-  xp[is.na(xp)] <- xpa[is.na(xp)]
-
-  xa <- match(stomata$species, plantatt$Taxon.name)
-  xaa <- match(stomata$acceptedname, plantatt$Taxon.name)
-  xa[is.na(xa)] <- xaa[is.na(xa)]
-
-  xl <- match(stomata$species, leaf_size$species)
-  xla <- match(stomata$acceptedname, leaf_size$species)
-  xl[is.na(xl)] <- xla[is.na(xl)]
-
-  xs <- match(stomata$species, seed_number$species)
-  xsa <- match(stomata$acceptedname, seed_number$species)
-  xs[is.na(xs)] <- xsa[is.na(xs)]
-
-  stomata$photo <- photo$photo[xp]
-  stomata$lifeform <- plantatt$LF1[xa]
-  stomata$ellenberg_light <- plantatt$L[xa]
-  stomata$height <- plantatt$Hght[xa]
-  stomata$wh <- plantatt$W[xa]
-  stomata$leaf_size <- leaf_size$leaf_size[xl]
-  stomata$seed_number <- seed_number$seed_number[xs]
-
-  # Remove missing values
-  stomata <- stomata[complete.cases(stomata[, c("sr_propAd", "lifeform", "ellenberg_light")]), ]
-
-  rm("already_in", "missing_photo", "new_name", "old_name", "photo", 
-     "plantatt", "salisbury", "tmp", "x", "x1", "x2", "xp", "xpa", "xa", "xaa")
-
-  # Export  
-  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
-
-@
-
-<<Leaf size, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # boxplot(sr_even ~ leaf_size, data = stomata)
-  # boxplot(ellenberg_light ~ leaf_size, data = stomata)
-
-  plot(sr_even ~ leaf_size, data = stomata, log = "x")
-  boxplot(log10(leaf_size) ~ ellenberg_light, data = stomata)
-  tmp <- read.csv(paste0(pathRawData, "/try_leaf_area.csv"), stringsAsFactors = FALSE,
-  								strip.white = TRUE)
-
-  stomata$species %in% tmp$AccSpeciesName | stomata$acceptedname %in% tmp$AccSpeciesName
- sort(table(tmp[which(tmp$AccSpeciesName %in% stomata$species), c("Dataset")]))
-@
-
-<<Phylogeny, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Import phylogeny of British Flora from Lim et al 2014
-  tmp <- readLines(paste0(pathRawData, "/S15105.nex"))
-  l1 <- grep("BEGIN CHARACTERS", tmp) - 1
-  l2 <- grep("BEGIN TREES", tmp) - 1
-  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathRawData, "/Lim_etal_2014.nex"))
-  phy <- read_nexus_phylo(paste0(pathRawData, "/Lim_etal_2014.nex"))
-
-  # Check for species missing from phylogeny
-  stomata$species %<>% str_replace_all(" ", "_")
-  stomata$acceptedname %<>% str_replace_all(" ", "_")
-  notInPhy <- which(!(stomata$species %in% phy$tip.label |
-                        stomata$acceptedname %in% phy$tip.label))
-  stomata[notInPhy, ]; length(notInPhy)
-
-  # Modify tip labels
-
-  # Using Cerastium brachypetalum as proxy for C. cerastoides based on similar divergence in Scheen et al. 2004 (Am J Bot, 91(6): 943-952)
-  tmp <- gsub("Cerastium_brachypetalum", "Cerastium_cerastoides", tmp)
-
-  # Using Coincya monensis as proxy for C. wrightii 
-  tmp <- gsub("Coincya_monensis", "Coincya_wrightii", tmp)
-
-  # Using Geranium pusillum as proxy for G. versicolor based on similar divergence in Palazzesi et al. 2012 (Bio J of Linn Soc, 107(1): 67-85)
-  # tmp <- gsub("Geranium_pusillum", "Geranium_versicolor", tmp)
-
-  # Using Gnaphalium luteoalbum as proxy for G. norvegicum 
-  tmp <- gsub("Gnaphalium_luteoalbum", "Gnaphalium_norvegicum", tmp)
-
-  # Lithospermum purpurocaeruleum is mispelled
-  tmp <- gsub("Lithospermum_purpureocaeruleum", "Lithospermum_purpurocaeruleum", tmp)
-
-  # Using Lythrum portula as proxy for L. hyssopifolia 
-  tmp <- gsub("Lythrum_portula", "Lythrum_hyssopifolia", tmp)
-
-  # Using Peucedanum ostruthium as proxy for P. officinale 
-  tmp <- gsub("Peucedanum_ostruthium", "Peucedanum_officinale", tmp)
-
-  # Changing Populus_nigra_sens.lat. to Populus nigra officinale 
-  tmp <- gsub("Populus_nigra_sens.lat.", "Populus_nigra", tmp)
-
-  # Changing Rosa_canina_agg. to Rosa_canina officinale 
-  tmp <- gsub("Rosa_canina_agg.", "Rosa_canina", tmp)
-
-  # Using Stachys alpina as proxy for S. germanica based on similar divergence in Salmaki et al. 2013 (Mol Phylo Evol, 69(3): 535-551)
-  tmp <- gsub("Stachys_alpina", "Stachys_germanica", tmp)
-
-  # Using Tripleurospermum_maritimum as proxy for T. inodorum 
-  tmp <- gsub("Tripleurospermum_maritimum", "Tripleurospermum_inodorum", tmp)
-
-  # Using Viola_arvensis as proxy for V. kitaibeliana 
-  tmp <- gsub("Viola_arvensis", "Viola_kitaibeliana", tmp)
-
-  # Write first modified phylogeny
-  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
-  phy <- read_nexus_phylo(paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
-
-  # Bind additional tips
-  tips <- phy$tip.label
-  nodes <- sapply(tips, function(x,y) which(y == x), y = phy$tip.label)
-  edge.lengths <- setNames(phy$edge.length[sapply(nodes, function(x, y) {
-    which(y == x) }, y = phy$edge[,2])], names(nodes))
-
-  # Make Agrostemma githago sister to all (Silene sp., Lychnis sp.) [based on Fior et al. 2006, Am J Bot 104(2):399-411]
-  n <- getMRCA(phy, phy$tip.label[c(grep("Silene", phy$tip.label), grep("Lychnis", phy$tip.label))])
-
-  # Branch length is average of distance from (Silene sp., Lychnis sp.) MRCA to (Silene sp., Lychnis sp.) sp.
-  el1 <- sapply(nodes[c(grep("Silene_", phy$tip.label), grep("Lychnis_", phy$tip.label))], 
-                nodeheight, tree = phy)
-  el2 <- nodeheight(tree = phy, node = n)
-
-  phy <- bind.tip(phy, "Agrostemma_githago", edge.length = mean(el1 - el2), where = n)
-
-  # Make Heracleum mantegazzianum as sister to H. sphondylium
-  phy <- bind.tip(phy, "Heracleum_mantegazzianum", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Heracleum_sphondylium")],
-                  where = which(phy$tip.label == "Heracleum_sphondylium"))
-
-  # Make Portulaca oleracea sister to all Claytonia (based on Angiosperm Phylogeny Website v 13 http://www.mobot.org/mobot/research/apweb/welcome.html]
-  # n <- getMRCA(phy, phy$tip.label[grep("Claytonia", phy$tip.label)])
-
-  # Branch length is average of distance from Claytonia MRCA to Claytonia sp.
-  # el1 <- sapply(nodes[grep("Claytonia_", names(nodes))], nodeheight, tree = phy)
-  # el2 <- nodeheight(tree = phy, node = n)
-
-  # phy <- bind.tip(phy, "Portulaca_oleracea", edge.length = mean(el1 - el2), where = n)
-
-  # Make Spartina alterniflora, Spartina maritima, and Spartina x townsendii as sister to S. anglica
-  phy <- bind.tip(phy, "Spartina_alterniflora", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-                  where = which(phy$tip.label == "Spartina_anglica"))
-
-  phy <- bind.tip(phy, "Spartina_maritima", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-                  where = which(phy$tip.label == "Spartina_anglica"))
-
-  # phy <- bind.tip(phy, "Spartina_x_townsendii", 
-  #                 edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
-  #                 where = which(phy$tip.label == "Spartina_anglica"))
-
-  # Make Tephroseris integrifolia sister to all Senecio sp.
-  n <- getMRCA(phy, phy$tip.label[grep("Senecio", phy$tip.label)])
-
-  # Branch length is average of distance from Senecio MRCA to Senecio sp.
-  el1 <- sapply(nodes[grep("Senecio_", names(nodes))], nodeheight, tree = phy)
-  el2 <- nodeheight(tree = phy, node = n)
-
-  phy <- bind.tip(phy, "Tephroseris_integrifolia", edge.length = mean(el1 - el2), where = n)
-
-  # Make Viola hirta as sister to V. reichenbachiana (both in Viola sect. Viola)
-  phy <- bind.tip(phy, "Viola_hirta", 
-                  edge.length = phy$edge.length[which(phy$tip.label == "Viola_reichenbachiana")],
-                  where = which(phy$tip.label == "Viola_reichenbachiana"))
-
-  # Write second modified phylogeny
-  write.nexus(phy, file = paste0(pathProcData, "/Lim_etal_2014_mod2.nex"))
-
-  # Check for species missing from modified phylogeny
-  notInPhy <- which(!(stomata$species %in% phy$tip.label |
-                        stomata$acceptedname %in% phy$tip.label))
-  stomata[notInPhy, ]; length(notInPhy)
-
-  rm("edge.lengths", "el1", "el2", "l1", "l2", "n", "nodes", "notInPhy", "tips", "tmp")
-
-  # Run PATHd8 (Britton et al. 2002)
-  cat(sprintf("Sequence length = 4692;
-
-              %s
-
-              mrca: Azolla_filiculoides, Solanum_nigrum, maxage=454;
-              mrca: Osmunda_regalis, Asplenium_onopteris, minage=354;
-              mrca: Pinus_sylvestris, Solanum_nigrum, minage=310;
-              mrca: Nymphaea_alba, Solanum_nigrum, minage=140;
-              mrca: Nymphaea_alba, Solanum_nigrum, maxage=217;
-              mrca: Nymphaea_alba, Nuphar_lutea, minage=115;
-              mrca: Chelidonium_majus, Solanum_nigrum, fixage=121;
-              mrca: Quercus_robur, Bryonia_dioica, minage=84;
-
-              name of mrca: Azolla_filiculoides, Solanum_nigrum, name=tracheophytes;
-              name of mrca: Osmunda_regalis, Asplenium_onopteris, name=monilophytes;
-              name of mrca: Pinus_sylvestris, Solanum_nigrum, name=seedplants;
-              name of mrca: Nymphaea_alba, Solanum_nigrum, name=angiosperms;
-              name of mrca: Nymphaea_alba, Nuphar_lutea, name=nymphaceae;
-              name of mrca: Chelidonium_majus, Solanum_nigrum, name=eudicots;
-              name of mrca: Quercus_robur, Bryonia_dioica, name=curcibitales_fagales;
-
-              ", write.tree(phy)), file = paste0(pathMain, "/PATHd8/pathd8_in"))
-
-  # Run PATHd8
-  system("cd ~/Google\\ Drive/StomatalRatio/BEF/PATHd8; ./PATHd8 pathd8_in pathd8_out")
-
-  # Import ultrametric tree
-  pathd8_out <- readLines(paste0(pathMain, "/PATHd8/pathd8_out"))
-  phy <- read.tree(text = substr(pathd8_out[11], 14, nchar(pathd8_out[11])))
-
-  rm("pathd8_out")
-
-@
 

### 2017-07-11 R/header.R

@@ -1,8 +1,8 @@
 rm(list = ls())
 graphics.off()
 
-compilte_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
-if (compilte_pathd8) {
+compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
+if (compile_pathd8) {
   system("cd PATHd8; cc PATHd8.c -O3 -lm -o PATHd8")
 }
 

### 2017-07-11 R/make-phylogeny.R

@@ -120,8 +120,9 @@ phy <- bind.tip(phy, "Viola_hirta",
 write.nexus(phy, file = paste0(path_proc_data, "/Lim_etal_2014_mod2.nex"))
 
 # Check for species missing from modified phylogeny
-notInPhy <- which(!stomata$species1 %in% phy$tip.label)
-stomata[notInPhy, ]
+notInPhy <- which(!(stomata$species %in% phy$tip.label |
+                      stomata$acceptedname %in% phy$tip.label))
+stomata[notInPhy, ]; length(notInPhy)
 
 rm("edge.lengths", "el1", "el2", "l1", "l2", "n", "nodes", "notInPhy", "tips", "tmp")
 

### 2017-07-11 R/process-data.R

@@ -1,48 +1,155 @@
 source("R/header.R")
 
 # British Ecological Flora
-
 stomata <- read_csv(str_c(path_raw_data, "/stomata.csv"))
 
 # Salisbury 1927
-
 salisbury <- read_csv(str_c(path_raw_data, "/salisbury_1927.csv"))
 
 # Taxize Salisbury dataset - this is done as a loop to prevent timing out
-tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
-for (i in 2:nrow(salisbury)) {
-  tnr %<>% bind_rows(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
-}
-write_csv(tnr, path = str_c(path_raw_data, "/taxized_salisbury_1927.csv"))
+# tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
+# for (i in 2:nrow(salisbury)) {
+#   tnr %<>% bind_rows(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
+# }
+# write_csv(tnr, path = str_c(path_raw_data, "/taxized_salisbury_1927.csv"))
 tnr <- read_csv(str_c(path_raw_data, "/taxized_salisbury_1927.csv"))
-############## WORKING TO ADD IN CODE FROM MS.RNW WHERE ###################
+
+# When TNRS does not give an accepted name, I am using these based on searching The Plant List
+# http://www.theplantlist.org/
+# Plant List corrections
+old_name <- c("Betula alba",            "Betula pubescens",         "Fagus sylvatica",
+              "Ulmus montana",          "Corylus avellana",         "Crataegus oxyacanthoides",
+              "Rubus corylifolius",     "Cephalanthera pallens",    "Conopodium denudatum",
+              "Helleborus foetidus",    "Primula acaulis",          "Byronia dioica",
+              "Geranium phaeum",        "Geranium striatum",        "Hieracium sylvaticum",
+              "Hypericum pulchrum",     "Lathyrus macrorrhizus",    "Luzula forsteri",
+              "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
+              "Tilia parvifolia",       "Tilia vulgaris",           "Vicia sylvatica",
+              "Hordeum sylvaticum")
+new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylvatica",
[…]

### 2017-07-11 ms/ms.Rnw
rejiggered lf-light analysis

@@
 % \pagenumbering{gobble}
-
-<<Violin plot of Raunkiaer life form versus stomatal ratio, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # chamaephyte = subshrub (woody)
-  # geophyte = cryptophyte resting in dry ground
-  # hemicryptophyte = perennial with overwintering buds at soil surface
-  # hydrophyte = cryptophyte resting in marshy ground
-  # phanerophyte = normally woody perennial
-  # therophyte = annual
-
-  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
-  names(lf) = c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
-  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-  pdf(paste0(pathFigures, "/FigureS_violin.pdf"), 8, 5.25)
-  par(mai = c(1.75, 1.5, 0.25, 0.25), cex.lab = 1.5, las = 1, cex = 1.5, cex.axis = 0.5)
-
-  vioplot(subset(stomata$sr_propAd, stomata$lifeform == names(lf[1]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[2]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[3]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[4]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[5]), na.rm = TRUE),
-          subset(stomata$sr_propAd, stomata$lifeform == names(lf[6]), na.rm = TRUE),
-          col = "grey", colMed = "black", names = rep("", 6))
-
-  clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
-       grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
-  # Sample sizes
-  text(1:6 + 0.25, grconvertY(1.5, "inches", "user"), labels = lf, srt = 30, pos = 2)
-  mtext(table(stomata$lifeform)[names(lf)], 3, at = 1:6, line = 0)
-  title(ylab = "Stomatal Ratio",  line = 3.5)
-  title(xlab = "Raunkiaer Life Form", line = 4.5)
-  title(ylab = expression(SR[propAd] == SD[ad] / SD[total]), line = 2, cex.lab = 1) 
-  dev.off()
-
-  rm("lf")
-
-@
-
-<<Remove hydrophytes + c4 + cam + nonangiosperm, eval=TRUE, echo=FALSE, results=hide>>=
-
-  nHydrophyte <- length(which(stomata$lifeform == "Hy"))
-  nC4 <- length(which(stomata$photo == "c4"))
-  nCAM <- length(which(stomata$photo == "cam"))
-
-  x1 <- c("Ph", "Ch", "hc", "Gn", "Th")
-  stomata <- stomata[which(stomata$lifeform %in% x1 & stomata$photo == "c3"), ]
-
-  angioPhy <- extract.clade(phy, 
-                            node = length(phy$tip.label) + which(phy$node.label == "angiosperms"))
-
-  nNonangio <- length(which(!(stomata$species %in% angioPhy$tip.label | 
-                                stomata$acceptedname %in% angioPhy$tip.label)))
-
-  stomata <- stomata[stomata$species %in% angioPhy$tip.label |
-                       stomata$acceptedname %in% angioPhy$tip.label, ]
-
-  # Export new dataset and tree
-  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
-  write.nexus(angioPhy, file = paste0(pathProcData, "/angioPhy.nex"))
-
-  rm("x1", "phy")
-
-@
 
@@
      "s1", "s2", "slopeCI", "x")
-
-@
-
- <<Testing leaf size model, eval=TRUE, echo=FALSE, results=hide>>=
-
-  stomata$leaf_size %<>% replace(. == 0, NA)
-  stomata %<>% mutate(log_leaf_size = log10(leaf_size))
-  tmp <- drop.tip(modAngioPhy, stomata$species[is.na(stomata$leaf_size)])
-
-  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light * log_leaf_size, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
-  fitSRb <- phylolm(sr_even ~ lifeform * ellenberg_light + log_leaf_size, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
-  fitSRc <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
-                    data = stomata[!is.na(stomata$leaf_size), ], phy = tmp, upper.bound = 10)
 

### 2017-07-11 R/run-all.R

@@ -4,15 +4,15 @@
 # source("R/session-info.R")
 
 # Prepare data
-source("R/process-data.R")
-source("R/make-phylogeny.R")
-source("R/plot-lifeform.R")
-source("R/filter-data.R")
-source("R/modify-phylogeny.R")
+source("R/01-process-data.R")
+source("R/02-make-phylogeny.R")
+source("R/03-plot-lifeform.R")
+source("R/04-filter-data.R") # starting here (if not before) need place to export stuff for use in ms
+source("R/05-modify-phylogeny.R")
 
 # Analyze data
-source("R/analyse-light-vs-lifeform.R")
-source("R/analyse-stomata.R")
+source("R/06-analyse-light-vs-lifeform.R")
+source("R/07-analyse-stomata.R")
 
 # NEED TO GET PATH ANALYSIS STUFF
 # NEED TO KNIT REPORT

### 2017-07-11 r/03-plot-lifeform.R

@@ -1,7 +1,6 @@
 source("R/header.R")
 
 stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"))
-phy <- read.nexus(file = str_c(path_proc_data, "/Lim_etal_2014_final.nex"))
 
 ##### Plot lifeform versus stomatal ratio -----
 

### 2017-07-11 r/04-filter-data.R

@@ -5,11 +5,14 @@ phy <- read.nexus(file = str_c(path_proc_data, "/Lim_etal_2014_final.nex"))
 
 ##### Filter hydrophytes, helophytes, c4, cam, and nonangiosperm plants -----
 
+nHydrophyte <- length(which(stomata$lifeform == "Hy"))
+nC4 <- length(which(stomata$photo == "c4"))
+nCAM <- length(which(stomata$photo == "cam"))
+
 stomata %<>% filter(lifeform %in% c("Ph", "Ch", "hc", "Gn", "Th"),
                     photo == "c3") %>%
   select(-photo)
 
-stomata$species %<>% str_replace_all(" ", "_")
 angio_phy <- extract.clade(phy, node = length(phy$tip.label) + 
                              which(phy$node.label == "angiosperms"))
 

### 2017-07-11 r/06-analyse-light-vs-lifeform.R

@@ -5,19 +5,18 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
 
 ##### Raunikaer life form versus Ellenberg light indicator values -----
 
-# Phylogenetic ANOVA (takes a minute)
-# fitLFvEL_pp <- phylopars.lm(ellenberg_light ~ lifeform,
-#                             trait_data = stomata[, c("species", "ellenberg_light", "lifeform")],
-#                             tree = phy, model = "OU", REML = F)
-# write_rds(fitLFvEL_pp, path = str_c(path_objects, "/fitLFvEL_pp.rds"))
-fitLFvEL_pp <- read_rds(str_c(path_objects, "/fitLFvEL_pp.rds"))
-anova(fitLFvEL_pp)
-
+# phylolm shows that there is no phylogenetic signal in this regression (very high alpha)
 stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
 fitLFvEL_pl <- phylolm(ellenberg_light ~ -1 + lifeform, model = "OUrandomRoot", 
-                       data = stomata, phy = phy)
+                       data = stomata, phy = phy, upper.bound = 1e10)
 summary(fitLFvEL_pl)
-confint(fitLFvEL_pl)
+
+# therefore using standard ANOVA
+fitLFvEL <- lm(ellenberg_light ~ lifeform, data = stomata)
+aovLFvEL <- anova(fitLFvEL)
+
+# for plotting confidence intervals
+fitLFvEL <- lm(ellenberg_light ~ -1 + lifeform, data = stomata)
 
 lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
 names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")

### 2017-07-11 ms/ms.Rnw
added stomatal density analysis scripts

@@
 
-<<Raunkiaer life form and Ellenberg light indicator values versus SR_even, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # Phylogenetic regression model comparison
-  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRb <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRd <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
-                    data = stomata, phy = modAngioPhy, upper.bound = 10)
-  aicSR <- sapply(list(fitSRa, fitSRb, fitSRc, fitSRd, fitSRe), AIC)
-
-  # Parametric bootstrap CIs of full model for p-values
-  # set.seed(36502)
-  # fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, model = "OUrandomRoot",
-  #                    data = stomata, phy = modAngioPhy, boot = 1e4, upper.bound = 10)
-  # saveRDS(fitSR_pl, file = paste0(pathR, "/fitSR_pl.rds"))
-  fitSR_pl <- readRDS(file = paste0(pathR, "/fitSR_pl.rds"))
-
-  # Figure
-
-  pdf(paste0(pathFigures, "/Figure_SRmultReg.pdf"), 4, 7)
-  par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
-
-  for (i in 5:1) {
-
-    plot(0, 0, type = "n", xlim = c(1, 9), ylim = c(-0.05, 1.05), xlab = "", ylab = "", axes = F,
-         frame.plot = F)
-    if (i %in% seq(1, 5, 2)) axis(2, at = seq(0, 1, 0.5), lwd = 0, lwd.ticks = 1, las = 1)
-
-    # polygon of confidence intervals
-    x <- seq(min(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]),
-             max(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]), length.out = 1e3)
-    bs <- sapply(x, function(X) {
-      fitSR_pl$bootstrap[, sprintf('lifeform%s', names(lf)[i])] + 
-        fitSR_pl$bootstrap[, sprintf('lifeform%s:ellenberg_light', names(lf)[i])] * X
-    } )
-    ci <- apply(bs, 2, quantile, probs = c(0.025, 0.5, 0.975))
-    polygon(c(x, rev(x)), c(ci[1, ], rev(ci[3, ])), col = "grey", border = NA)
-    points(x, ci[2, ], lwd = 2, type = "l") # median
-    points(x, ci[1, ], lwd = 2, lty = 2, type = "l") # lower 95%
-    points(x, ci[3, ], lwd = 2, lty = 2, type = "l") # upper 95%
-
-    # plot points
-    with(subset(stomata, stomata$lifeform == names(lf)[i]), 
-         points(jitter(ellenberg_light), sr_even, pch = 21, bg = rgb(0, 0, 0, 0.1), cex = 2))
-
-    # label life forms, slope, and statistical significance
-    s1 <- sprintf("%s (%s)", lf[i], table(stomata$lifeform)[names(lf)[i]])
-    text(1 + 0.5 * strwidth(s1), 1.05, labels = s1, pos = 1, adj = 1)
-    slopeCI <- confint(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", names(lf)[i]), ]
-    pval <- 1 - length(which(fitSR_pl$bootstrap[, sprintf("lifeform%s:ellenberg_light", 
-                                                          names(lf)[i])] > 0)) / fitSR_pl$boot 
-    s2 <- bquote(slope == .(round(coef(fitSR_pl)[sprintf("lifeform%s:ellenberg_light", 
-                                                         names(lf)[i])], 3))^.(sigStar(pval)))
-    text(1 + 0.5 * strwidth(s2), 1.05 - 1.5 * strheight(s1), labels = s2, pos = 1, adj = 1)
-
-  }
-
-  # frame plot
-  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
-       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
-  abline(h = grconvertY(0:5, "npc", "user"))
-  lines(rep(grconvertX(0, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))
-  lines(rep(grconvertX(1, "npc", "user"), 2), grconvertY(c(0, 5), "npc", "user"))
-
-  # axis labels
-  axis(1, at = 1:9, labels = 1:9, lwd = 0, line = -1)
-  axis(1, at = c(2, 5, 8), lwd = 0, labels = c("Shade", "Partial Shade", "Sun"),
-       line = 0)
-  mtext(1, text = "Ellenberg light indicator", cex = 1.5, line = 3, outer = T)
-
-  mtext(2, text = "Stomatal Ratio",  line = 4.5, outer = T, cex = 1.5)
-
-  mtext(2, text = expression(
-    SR[even] == plain(min)~group("(", list(SD[ab], SD[ad]), ")") /
-      plain(max)~group("(", list(SD[ab], SD[ad]), ")")), outer = T, line = 2.5)
-
-  dev.off()
-
-  # Parametric bootstrap statistical signicance of deltaAIC between model 
-  # with interaction (fitSRa) and model without interaction (fitSRb)
-  # set.seed(444294)
-  # bootModB <- predict(fitSRb) + rTrait(n = 1e3, modAngioPhy, "OU", 
-  #                                     list(sigma2 = fitSRb$sigma2, alpha = fitSRb$optpar))
-  # colnames(bootModB) <- paste0("b", 1:ncol(bootModB))
-
-  # fit model of model A to data simulated from model B
-  # fitA2B <- lapply(colnames(bootModB), function(col) {
-  #  stomata$Y <- bootModB[, col]
-  #  phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata, 
-  #          phy = modAngioPhy, upper.bound = 10)
-  #  })
-
-  # fit model of model A to data simulated from model B
-  # fitB2B <- lapply(colnames(bootModB), function(col) {
-  #  stomata$Y <- bootModB[, col]
-  #  phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata, 
-  #          phy = modAngioPhy, upper.bound = 10)
-  #  })
-
-  # refit models at parameter bounds
-  # for (i in which(round(sapply(fitA2B, function(X) X$optpar), 2) == 10)) {
-  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
-  #  fitA2B[[i]] <- phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata,
-  #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200, 
-  #                         starting.value = 10)
-  # }
-
-  # refit models at parameter bounds
-  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 10)) {
-  #  stomata$Y <- bootModB[, colnames(bootModB)[i]]
-   #  fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
-   #                        phy = modAngioPhy, lower.bound = 10, upper.bound = 200,
-   #                        starting.value = 10)
-   # }
-
-  # for some reason, this fixes those with alpha estimated at upper bound
-  # for (i in which(round(sapply(fitB2B, function(X) X$optpar), 2) == 200)) {
-  #   stomata$Y <- bootModB[, colnames(bootModB)[i]]
-  #   fitB2B[[i]] <- phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata,
-  #                          phy = modAngioPhy, lower.bound = 0, upper.bound = 300, 
-  #                          starting.value = 1)
-  # }
-
-  # saveRDS(fitA2B, paste0(pathR, "/fitA2B.rds"))
-  # saveRDS(fitB2B, paste0(pathR, "/fitB2B.rds"))
-
-  fitA2B <- readRDS(paste0(pathR, "/fitA2B.rds"))
-  fitB2B <- readRDS(paste0(pathR, "/fitB2B.rds"))
-
-  # Distribution of deltaAIC
-  obsDeltaAIC <- fitSRa$aic - fitSRb$aic
-  distDeltaAIC <- sapply(fitA2B, function(X) X$aic) - sapply(fitB2B, function(X) X$aic)
-  pInteraction <- length(which(obsDeltaAIC > distDeltaAIC)) / 1e3
-
-  rm("bs", "ci", "distDeltaAIC", "fitA2B", "fitB2B", "i", "lf", "obsDeltaAIC", "pval",
-     "s1", "s2", "slopeCI", "x")
-
-@
-
 I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
@@
 I tested whether parameter estimates were significantly different than zero using $z$-scores.
-
-<<Stomatal density + light + growth form, eval=TRUE, echo=FALSE, results=hide>>=
-
-  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", 
-          "phanerophyte", "therophyte")
-  names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
-  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-  # density of stomatal density by side and L-value
-  dnsAb <- tapply(log10(stomata$ab_density + 1), stomata$ellenberg_light, function(X) {
-    bp <- boxplot(X, plot = F)
-    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
-                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
-    ret$stats <- bp$stats
-    ret$out <- bp$out
-    ret
-  } )
-
-  dnsAd <- tapply(log10(stomata$ad_density + 1), stomata$ellenberg_light, function(X) {
-    bp <- boxplot(X, plot = F)
-    ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
-                   adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
-    ret$stats <- bp$stats
-    ret$out <- bp$out
-    ret
-  } )
-
-  pdf(paste0(pathFigures, "/Figure_SD-light.pdf"), 8, 10)
-  par(mar = rep(0, 4), oma = c(5, 6, 1, 1), las = 1, mfrow = c(2, 1))
-  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
-
-  mtext(1, text = "Ellenberg light indicator value", outer = T, cex = 2, line = 3)
-  mtext(2, text = expression(paste("Stomatal density (no. ", mu, m^-2, ") [log scale]")), 
-        outer = T, cex = 2, las = 0, line = 3)
-  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))
-
-  # lines
-  abline(v = 3:9, col = "grey")
-
-  # adaxial
-  for (i in 3:9) {
-
-    adDns <- dnsAd[[sprintf("%s", i)]]
-
-    # scale density for plotting  
-    adDns$d <- adDns$y * 0.4 / max(adDns$y)
-
-    # polygon
-    polygon(c(i + adDns$d, rev(i - adDns$d)), c(adDns$x, rev(adDns$x)), col = "grey")
-
-    # outliers
-    points(rep(i, length(adDns$out)), adDns$out, pch = 19)
-
-    # bars (0.25 and 0.75 quantiles?)
-    lines(rep(i, 2), adDns$stats[c(2, 4), 1], lwd = 5, lend = 1)
-
-    # median
-    points(i, adDns$stats[3, 1], pch = 19, cex = 1.5)
-
-    # labels
-    s <- "A. adaxial (upper)"
-    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
-         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)
-
-  }
-
-  plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
-
-  axis(1, at = 3:9)
-  axis(2, at = 0:3, labels = c(1, 10, 100, 1000))
-
-  # lines
-  abline(v = 3:9, col = "grey")
-
-  # abaxial
-  for (i in 3:9) {
-
-    abDns <- dnsAb[[sprintf("%s", i)]]
-
-    # scale density for plotting  
-    abDns$d <- abDns$y * 0.4 / max(abDns$y)
-
-    # polygons
-    polygon(c(i - abDns$d, rev(i + abDns$d)), c(abDns$x, rev(abDns$x)), col = "grey")
-
-    # outliers
-    points(rep(i, length(adDns$out)), adDns$out, pch = 19)
-
-    # bars (0.25 and 0.75 quantiles?)
-    lines(rep(i, 2), abDns$stats[c(2, 4), 1], lwd = 5, lend = 1)
-
-    # median
-    points(i, abDns$stats[3, 1], pch = 19, cex = 1.5)
-
-    # labels
-    s <- "B. abaxial (lower)"
-    text(grconvertX(0.025, "nfc", "user") + strwidth(s, cex = 2) / 2, 
-         grconvertY(0.975, "nfc", "user"), labels = s, cex = 2, pos = 1)
-
-  }
-
-  dev.off()
-
-  rm("abDns", "adDns", "dnsAb", "dnsAd", "i", "lf", "s")
-
-@
-
-<<Phylogenetic structural equation model, eval=TRUE, echo=FALSE, results=hide>>=
-
-  # small data.frame for Rphylopars
-  td <- data.frame(species = stomata$species)
-  td$logSDab <- log(stomata$ab_density + 1)
-  td$logSDad <- log(stomata$ad_density + 1)
-  td$ellenberg_light <- stomata$ellenberg_light
-
-  # ppOU <- phylopars(td, modAngioPhy, model = "mvOU")
-  # saveRDS(ppOU, paste0(pathR, "/ppOU.rds"))
-  ppOU <- readRDS(paste0(pathR, "/ppOU.rds"))
-  phyMat <- ppOU$pars$phylocov
-
-  # define (co)variances for use in text
-  varSDab <- phyMat["logSDab", "logSDab"]
-  varSDad <- phyMat["logSDad", "logSDad"]
-  covSD <- phyMat["logSDab", "logSDad"]
-  varSR <- varSDab + varSDad - 2 * covSD
-
-  ppBM <- phylopars(td, modAngioPhy, model = "BM")
-
-  # AIC comparison of OU and BM
-  AIC(ppOU, ppBM)
-
-  # Strucutural equation model using lavaan
-  model <- 'logSDab ~ ellenberg_light
-            logSDad ~ ellenberg_light'
-
-  fitSEM <- sem(model, sample.cov = ppOU$pars$phylocov, sample.nobs = nrow(td))
-
-  # Calculate p-values based on z-scores
-  pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, log = TRUE)
-  names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)
-
-  rm("td")
-
-@
-
-<<COMPADRE, eval=TRUE, echo=FALSE, results=hide>>=
-
-compadre <- read.csv(paste0(pathRawData, "/Demographics - Feb 17 2017.csv"),
-                     stringsAsFactors = FALSE)
-compadre$SpeciesAccepted %<>% str_replace_all(" ", "_")
-compadre %<>% rename(species = SpeciesAccepted)
-compadre %<>% inner_join(stomata, by = "species")
-
-plot(compadre$T[compadre$T > 0], compadre$sr_even[compadre$T > 0])
-plot(compadre$T[compadre$T > 0], compadre$sr_propAd[compadre$T > 0])
-cor.test(compadre$Efec, compadre$sr_even)
-
-@
 

### 2017-07-11 R/run-all.R

@@ -12,7 +12,8 @@ source("R/05-modify-phylogeny.R")
 
 # Analyze data
 source("R/06-analyse-light-vs-lifeform.R")
-source("R/07-analyse-stomata.R")
+source("R/07-analyse-stomatal-ratio.R")
+source("R/08-analyse-stomatal-density.R")
 
 # NEED TO GET PATH ANALYSIS STUFF
 # NEED TO KNIT REPORT

### 2017-07-11 r/07-analyse-stomata.R

@@ -1,85 +1,0 @@
-source("R/header.R")
-
-phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
-stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
-
-##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
-
-# Phylogenetic ANOVA (takes several minutes to run)
-# fitSR_pp <- phylopars.lm(sr_even ~ ellenberg_light * lifeform,
-#                          trait_data = stomata[, c("species", "sr_even", "ellenberg_light", "lifeform")],
-#                          tree = phy, model = "OU", REML = F)
-# write_rds(fitSR_pp, path = str_c(path_objects, "/fitSR_pp.rds"))
-fitSR_pp <- read_rds(str_c(path_objects, "/fitSR_pp.rds"))
-anova(fitSR_pp)
-
-# Parametric bootstrap CIs of full model for plotting
-# NOTE - this could be done with Rphylopars, which might be better because parameter estimates are slightly different for some reason
-# set.seed(36502)
-# fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, 
-#                     model = "OUrandomRoot", data = stomata, phy = phy, boot = 1e3)
-# write_rds(fitSR_pl, path = str_c(path_objects, "/fitSR_pl.rds"))
-fitSR_pl <- read_rds(str_c(path_objects, "/fitSR_pl.rds"))
-
-# Figure
-
-pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7)
-par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
-
-lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
-names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
-lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
-
-for (i in 5:1) {
-  
-  plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(-0.05, 1.05), xlab = "", 
-       ylab = "", axes = F, frame.plot = F)
-  if (i %in% seq(1, 5, 2)) {
[…]

### 2017-07-11 r/07-analyse-stomatal-ratio.R

@@ -1,0 +1,150 @@
+source("R/header.R")
+
+phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+
+##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
+
+# Phylogenetic regression model comparison
+stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
+fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
+                  data = stomata, phy = phy, upper.bound = 10)
+fitSRb <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
+                  data = stomata, phy = phy, upper.bound = 10)
+fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
+                  data = stomata, phy = phy, upper.bound = 10)
+fitSRd <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
+                  data = stomata, phy = phy, upper.bound = 10)
+fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
+                  data = stomata, phy = phy, upper.bound = 10)
+aicSR <- sapply(list(fitSRa, fitSRb, fitSRc, fitSRd, fitSRe), AIC)
+
+# Parametric bootstrap CIs of full model for p-values
+set.seed(36502)
+fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, 
+                    model = "OUrandomRoot", data = stomata, phy = phy, 
+                    boot = 1e4, upper.bound = 10)
+write_rds(fitSR_pl, file = str_c(path_objects, "/fitSR_pl.rds"))
+fitSR_pl <- read_rds(file = paste0(path_objects, "/fitSR_pl.rds"))
+
+# Figure
+
+pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7)
+par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
+
+for (i in 5:1) {
+  
+  plot(0, 0, type = "n", xlim = c(1, 9), ylim = c(-0.05, 1.05), xlab = "", 
[…]

### 2017-07-11 r/08-analyse-stomatal-density.R

@@ -1,0 +1,106 @@
+source("R/header.R")
+
+phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+
+##### Stomatal density, light, and lifeform -----
+  
+lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", 
+        "phanerophyte", "therophyte")
+names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
+lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+
+# density of stomatal density by side and L-value
+dnsAb <- tapply(log10(stomata$ab_density + 1), stomata$ellenberg_light, function(X) {
+  bp <- boxplot(X, plot = F)
+  ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
+                 adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
+  ret$stats <- bp$stats
+  ret$out <- bp$out
+  ret
+} )
+
+dnsAd <- tapply(log10(stomata$ad_density + 1), stomata$ellenberg_light, function(X) {
+  bp <- boxplot(X, plot = F)
+  ret <- density(X[!X %in% bp$out], kernel = "gaussian", n = 1e3, 
+                 adjust = 2, from = bp$stats[1, 1], to = bp$stats[5, 1])
+  ret$stats <- bp$stats
+  ret$out <- bp$out
+  ret
+} )
+
+pdf(str_c(path_figures, "/figure_SD-light.pdf"), 8, 10)
+par(mar = rep(0, 4), oma = c(5, 6, 1, 1), las = 1, mfrow = c(2, 1))
+plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
+
+mtext(1, text = "Ellenberg light indicator value", outer = T, cex = 2, line = 3)
+mtext(2, text = expression(paste("Stomatal density (no. ", mu, m^-2, ") [log scale]")), 
+      outer = T, cex = 2, las = 0, line = 3)
+axis(2, at = 0:3, labels = c(1, 10, 100, 1000))
+
[…]

### 2017-07-11 r/09-analyse-phyloSEM.R

@@ -1,0 +1,44 @@
+source("R/header.R")
+
+phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+
+##### Phylogenetic structural equation model ---
+  
+  # small data.frame for Rphylopars
+td <- stomata
+
+  td <- data.frame(species = stomata$species)
+  td$logSDab <- log(stomata$ab_density + 1)
+  td$logSDad <- log(stomata$ad_density + 1)
+  td$ellenberg_light <- stomata$ellenberg_light
+  
+  # ppOU <- phylopars(td, modAngioPhy, model = "mvOU")
+  # saveRDS(ppOU, paste0(pathR, "/ppOU.rds"))
+  ppOU <- readRDS(paste0(pathR, "/ppOU.rds"))
+  phyMat <- ppOU$pars$phylocov
+  
+  # define (co)variances for use in text
+  varSDab <- phyMat["logSDab", "logSDab"]
+  varSDad <- phyMat["logSDad", "logSDad"]
+  covSD <- phyMat["logSDab", "logSDad"]
+  varSR <- varSDab + varSDad - 2 * covSD
+  
+  ppBM <- phylopars(td, modAngioPhy, model = "BM")
+  
+  # AIC comparison of OU and BM
+  AIC(ppOU, ppBM)
+  
+  # Strucutural equation model using lavaan
+  model <- 'logSDab ~ ellenberg_light
+  logSDad ~ ellenberg_light'
+  
+  fitSEM <- sem(model, sample.cov = ppOU$pars$phylocov, sample.nobs = nrow(td))
+  
+  # Calculate p-values based on z-scores
+  pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, log = TRUE)
+  names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)
+  
+  rm("td")
+  
+  @

### 2017-07-12 ms/ms.Rnw
exported objects for ms

@@
 
+<<Import objects, echo = FALSE>>=
+
+  source("r/header.R")
+
+  import2ms()
+
+@
+
 \title{Light and life form interact to shape stomatal ratio among British angiosperms$^1$}
@@
 \\
-Short title: Shedding light on British herbs\\
+Short title: Shedding light on stomatal evolution\\
 \\
@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = \Sexpr{pInteraction}$).
+Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
 

### 2017-07-12 R/functions.R

@@ -1,5 +1,28 @@
 # R functions associated with Muir 2017
 
+# Export objects to ms
+export2ms <- function(x, path_export = getOption("path_export", "/export")) {
+  
+  for (i in 1:length(x)) {
+    stopifnot(is.character(x[i]))
+    tmp <- eval(parse(text = x[i]))
+    path <- str_c(path_ms, path_export)
+    if (!dir.exists(path)) dir.create(path)
+    write_rds(tmp, str_c(path, "/", x[i]))
+  }
+  
+}
+
+# Import objects to ms
+import2ms <- function(path_export = getOption("path_export", "/export")) {
+  
+  path <- str_c(path_ms, path_export)
+  files <- list.files(path)
+  eval(parse(text = str_c(files, " <- read_rds('", path, "/", files, "')")),
+       envir = .GlobalEnv)
+
+}
+
 # Calculate sr_even
 calc_sr_even <- function(ab_density, ad_density) {
   apply(cbind(ab_density, ad_density), 1, min) / 

### 2017-07-12 R/header.R

@@ -28,9 +28,9 @@ library(vioplot)
 path_raw_data <- "raw-data"
 path_proc_data <- "proc-data"
 path_r <- "r"
-  path_objects <- "r/objects"
-pathM_ms <- "ms"
-  path_figures <- "ms/figures"
+  path_objects <- str_c(path_r, "/objects")
+path_ms <- "ms"
+  path_figures <- str_c(path_ms, "/figures")
 
 
 

### 2017-07-12 r/04-filter-data.R

@@ -1,4 +1,4 @@
-source("R/header.R")
+source("r/header.R")
 
 stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"))
 phy <- read.nexus(file = str_c(path_proc_data, "/Lim_etal_2014_final.nex"))
@@ -16,8 +16,21 @@ stomata %<>% filter(lifeform %in% c("Ph", "Ch", "hc", "Gn", "Th"),
 angio_phy <- extract.clade(phy, node = length(phy$tip.label) + 
                              which(phy$node.label == "angiosperms"))
 
+# Change species to acceptedname is latter is in phylogeny
+stomata$species[which(!stomata$species %in% phy$tip.label)] <-
+  stomata$acceptedname[which(!stomata$species %in% phy$tip.label)]
+stomata %<>% select(-acceptedname)
+
+n <- nrow(stomata)
+
 stomata %<>% filter(species %in% angio_phy$tip.label)
+
+nNonangio <- n - nrow(stomata)
+rm(n)
 
 # Export new dataset and tree
 write_csv(stomata, str_c(path_proc_data, "/stomata_filtered.csv"))
 write.nexus(angio_phy, file = str_c(path_proc_data, "/angio_phy.nex"))
+
+# Export objects to ms
+export2ms(c("nC4", "nCAM", "nHydrophyte", "nNonangio", "stomata"))

### 2017-07-12 r/06-analyse-light-vs-lifeform.R

@@ -61,3 +61,5 @@ for (i in 1:5) {
 
 dev.off()
 
+# Export objects to ms
+export2ms(c("aovLFvEL"))

### 2017-07-12 r/07-analyse-stomatal-ratio.R

@@ -92,6 +92,9 @@ mtext(2, text = expression(
 
 dev.off()
 
+# Export objects to ms
+export2ms(c("aicSR", "fitSRa", "fitSRb", "fitSRc", "fitSRd", "fitSRe"))
+
 # NOTE: Matt suggested deleting this, just using AIC
 # Parametric bootstrap statistical signicance of deltaAIC between model 
 # with interaction (fitSRa) and model without interaction (fitSRb)

### 2017-07-12 r/09-analyse-phyloSEM.R

@@ -37,3 +37,6 @@ fitSEM <- sem(model, sample.cov = ppOU$pars$phylocov, sample.nobs = nrow(td))
 # Calculate p-values based on z-scores
 pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, log = TRUE)
 names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)
+
+# Export objects to ms
+export2ms(c("covSD", "pvalSEM", "varSDab", "varSDad", "varSR"))

### 2017-07-12 ms/ms.Rnw
added ability to knit ms

@@
 
-<<Import objects, echo = FALSE>>=
-
-  source("r/header.R")
-
-  import2ms()
-
-@
-
 \title{Light and life form interact to shape stomatal ratio among British angiosperms$^1$}
@@
 \# Figures and \# Tables, \# references
+
+<<Import-objects, echo = FALSE>>=
+
+  setwd("..")
+  source("r/header.R")
+
+  import2ms()
+  setwd("ms")
+
+@
 
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_lf-light.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}
 \caption{Lifeforms have different tolerances for sun and shade among British angiosperms.
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SRmultReg.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
 \caption{The effect of light on stomatal ratio depends on Raunki\ae r life form.
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SD-light.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SD-light.pdf}}
 \caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis).
@@
 \begin{figure}[ht]
-\centerline{\includegraphics{Figures/FigureS_violin.pdf}}
+\centerline{\includegraphics{figures/figureS_violin.pdf}}
 \caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$).

### 2017-07-12 ms/ms.tex
added ability to knit ms

@@
+% Notes from call with Rob
+% van Tienderen - sensitivity approach better for evolution
+% send list of species to rob
+
 % TASKS
@@
 % in methods - mention stuff about Salisbury
+% leaf size
 
@@
 In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 374 species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 372 species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
@@
 I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
-The final dataset contained 374 species.
+The final dataset contained 372 species.
 The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.
@@
 
-Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 369}$ = 18.4, $P$ = 8.1 $\times10^{-14}$).
+Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
 Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = 0.002$).
+Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
 
@@
 Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio.
-The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.39, and -0.53, respectively.
+The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively.
 Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive.
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 8.8 $\times10^{-7}$ and 4.9 $\times10^{-5}$, respectively).
-However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.2 versus 0.1).
-Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.22-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
-The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.1 $\times10^{-11}$).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively).
+However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1).
+Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
+The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$).
 These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.
@@
   & 12
-  & -32.6
-  & 89.2
+  & -33.3
+  & 90.6
   & 0 \\
@@
   & 0.47
-  & 0.071
+  & 0.072
   & 0.32
   & 8
-  & -39.8
-  & 95.6
-  & 6.3 \\
+  & -40.3
+  & 96.5
+  & 6 \\
     \el
-  & 0.65
-  & 0.109
+  & 0.64
+  & 0.108
   & 0.26
   & 4
-  & -58.9
-  & 125.9
-  & 36.6 \\
+  & -59.3
+  & 126.6
+  & 36.1 \\
     lifeform
   & 0.34
-  & 0.066
+  & 0.067
   & 0.15
@@
   & 172.4
-  & 83.2 \\
+  & 81.8 \\
     1
   & 0.29
-  & 0.068
+  & 0.067
   & 0
   & 3
-  & -107.9
-  & 221.7
-  & 132.5 \\
+  & -107.5
+  & 221
+  & 130.5 \\
 	\bottomrule

### 2017-07-12 R/functions.R

@@ -6,7 +6,7 @@ export2ms <- function(x, path_export = getOption("path_export", "/export")) {
   for (i in 1:length(x)) {
     stopifnot(is.character(x[i]))
     tmp <- eval(parse(text = x[i]))
-    path <- str_c(path_ms, path_export)
+    path <- normalizePath(str_c(path_ms, path_export))
     if (!dir.exists(path)) dir.create(path)
     write_rds(tmp, str_c(path, "/", x[i]))
   }
@@ -16,7 +16,7 @@ export2ms <- function(x, path_export = getOption("path_export", "/export")) {
 # Import objects to ms
 import2ms <- function(path_export = getOption("path_export", "/export")) {
   
-  path <- str_c(path_ms, path_export)
+  path <- normalizePath(str_c(path_ms, path_export))
   files <- list.files(path)
   eval(parse(text = str_c(files, " <- read_rds('", path, "/", files, "')")),
        envir = .GlobalEnv)

### 2017-07-12 R/header.R

@@ -6,11 +6,10 @@ if (compile_pathd8) {
   system("cd PATHd8; cc PATHd8.c -O3 -lm -o PATHd8")
 }
 
-source("R/functions.R")
-
 # Libraries
 library(ape)
 library(dplyr)
+library(knitr)
 library(lavaan)
 library(magrittr)
 library(phylolm)
@@ -25,12 +24,11 @@ library(taxize)
 library(vioplot)
 
 # Directories
-path_raw_data <- "raw-data"
-path_proc_data <- "proc-data"
-path_r <- "r"
-  path_objects <- str_c(path_r, "/objects")
-path_ms <- "ms"
-  path_figures <- str_c(path_ms, "/figures")
+path_raw_data <- "raw-data" %>% normalizePath()
+path_proc_data <- "proc-data" %>% normalizePath()
+path_r <- "r" %>% normalizePath()
+path_objects <- str_c(path_r, "/objects") %>% normalizePath()
+path_ms <- "ms" %>% normalizePath()
+path_figures <- str_c(path_ms, "/figures") %>% normalizePath()
 
-
-
+source(str_c(path_r, "/functions.R"))

### 2017-07-12 R/install-packages.R

@@ -1,5 +1,6 @@
 install.packages("ape")
 install.packages("dplyr")
+install.packages("knitr")
 install.packages("lavaan")
 install.packages("magrittr")
 install.packages("phylolm")

### 2017-07-12 R/run-all.R

@@ -3,17 +3,22 @@
 
 # source("R/session-info.R")
 
+source("r/header.R")
+
 # Prepare data
 source("R/01-process-data.R")
 source("R/02-make-phylogeny.R")
 source("R/03-plot-lifeform.R")
-source("R/04-filter-data.R") # starting here (if not before) need place to export stuff for use in ms
+source("R/04-filter-data.R")
 source("R/05-modify-phylogeny.R")
 
 # Analyze data
 source("R/06-analyse-light-vs-lifeform.R")
 source("R/07-analyse-stomatal-ratio.R")
 source("R/08-analyse-stomatal-density.R")
+source("R/09-analyse-phyloSEM.R")
 
-# NEED TO GET PATH ANALYSIS STUFF
+knit("ms/ms.Rnw")
+
+# UPDATE BIBLIOGRAPHY
 # NEED TO KNIT REPORT

### 2017-07-12 r/01-process-data.R

@@ -38,9 +38,15 @@ new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylva
 
 tnr$acceptedname[match(old_name, tnr$submittedname)] <- new_name
 tnr %<>% mutate(species = submittedname)
+
+# Changed one taxonomic name (does not affect trait or phylogeny data)
+tnr$acceptedname %<>% str_replace("Arrhenatherum elatius var. elatius", 
+                                  "Arrhenatherum elatius")
+
 # tnr$acceptedname %>% str_split(" ") %>% sapply(length) %>% is_greater_than(1) %>% all()
 salisbury %<>% join(tnr, by = "species")
 rm(new_name, old_name, tnr)
+
 
 # Find which Salisbury data are already in BEF data
 already_in <- which(salisbury$submittedname %in% stomata$species |
@@ -83,15 +89,12 @@ stomata %<>%
 stomata %<>% 
   bind_rows(salisbury[!(x1 | x2), colnames(.)]) 
 
-# Remove duplicates
+# Remove duplicates and stop if varietes are left
 stomata %<>% filter(!duplicated(species))
-rm(already_in, salisbury, tmp, x, x1, x2)
-
-# Changed one taxonomic name (does not affect trait or phylogeny data)
-stomata$acceptedname %<>% str_replace("Arrhenatherum elatius var. elatius", 
-                                      "Arrhenatherum elatius")
 stopifnot(str_detect(c(stomata$species, stomata$acceptedname), " var. ") %>%
             na.omit() %>% any() %>% not())
+
+rm(already_in, salisbury, tmp, x, x1, x2)
 
 # Calculate stomatal ratio
 stomata %<>% mutate(sr_propAd = ad_density / (ab_density + ad_density),

### 2017-07-12 r/03-plot-lifeform.R

@@ -4,13 +4,6 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"))
 
 ##### Plot lifeform versus stomatal ratio -----
 
-# chamaephyte = subshrub (woody)
-# geophyte = cryptophyte resting in dry ground
-# hemicryptophyte = perennial with overwintering buds at soil surface
-# hydrophyte = cryptophyte resting in marshy ground
-# phanerophyte = normally woody perennial
-# therophyte = annual
-
 lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
 names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

### 2017-07-12 r/06-analyse-light-vs-lifeform.R

@@ -1,4 +1,4 @@
-source("R/header.R")
+source("r/header.R")
 
 phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
 stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
@@ -22,7 +22,7 @@ lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyt
 names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
-pdf(str_c(path_figures, "/figureS_lf-light.pdf"), 4, 7)
+pdf(str_c(path_figures, "/figure_lf-light.pdf"), 4, 7)
 par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
 plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(0, 5), xlab = "", ylab = "", axes = F)  
 

### 2017-07-12 r/07-analyse-stomatal-ratio.R

@@ -5,6 +5,10 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
 
 ##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
 
+lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
+names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
+lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+
 # Phylogenetic regression model comparison
 stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
 fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
@@ -24,8 +28,8 @@ set.seed(36502)
 fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, 
                     model = "OUrandomRoot", data = stomata, phy = phy, 
                     boot = 1e4, upper.bound = 10)
-write_rds(fitSR_pl, file = str_c(path_objects, "/fitSR_pl.rds"))
-fitSR_pl <- read_rds(file = paste0(path_objects, "/fitSR_pl.rds"))
+write_rds(fitSR_pl, path = str_c(path_objects, "/fitSR_pl.rds"))
+fitSR_pl <- read_rds(str_c(path_objects, "/fitSR_pl.rds"))
 
 # Figure
 

### 2017-07-12 r/09-analyse-phyloSEM.R

@@ -12,8 +12,8 @@ td <- stomata %>% select(species, ab_density, ad_density, ellenberg_light) %>%
   as.data.frame() %>% 
   set_rownames(.$species) # need to change row names for phylolm
 
-# ppOU <- phylopars(td, phy, model = "mvOU")
-# write_rds(ppOU, str_c(path_objects, "/ppOU.rds"))
+ppOU <- phylopars(td, phy, model = "mvOU")
+write_rds(ppOU, str_c(path_objects, "/ppOU.rds"))
 ppOU <- read_rds(str_c(path_objects, "/ppOU.rds"))
 phy_mat <- ppOU$pars$phylocov
   
@@ -39,4 +39,4 @@ pvalSEM <- pnorm(fitSEM@ParTable$est / fitSEM@ParTable$se, lower.tail = FALSE, l
 names(pvalSEM) <- paste(fitSEM@ParTable$lhs, fitSEM@ParTable$op, fitSEM@ParTable$rhs)
 
 # Export objects to ms
-export2ms(c("covSD", "pvalSEM", "varSDab", "varSDad", "varSR"))
+export2ms(c("covSD", "fitSEM", "pvalSEM", "varSDab", "varSDad", "varSR"))

### 2017-07-13 ms/ms.Rnw
edited introduction

@@
+% Things to do before submission:
+% Sally's notes
+% Matt's notes
+% Remove code to copy refs
+% Github DOI?
+% Finish abstract
+% Finish citations
+% Licenses (see http://ropensci.github.io/reproducibility-guide/sections/introduction/)
+
 % Notes from call with Rob
@@
 % leaf size
-
-% OPTIONAL
-% Synthesis model
-% total stomatal density versus growth form (as supplement???)
-% used ordered multinomial rather than lm for lf-light corr?
-% COMPADRE?
-% Model adequacy?
-% FIGURE FOR SEM?
 
@@
 
+  # Remove before publication
+  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
+            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+
   setwd("..")
@@
 \section*{Acknowledgements}
-
+I thank Sally Otto and Matt Pennell for feedback on this manuscript.
 % \section*{Author contribution statement}
@@
 In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on \Sexpr{nrow(stomata)} species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
@@
 While the density and size of stomata have been researched extensively [CITATIONS], the adaptive significance of stomatal distribution is less well understood.
-Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}.
+Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}.
 Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions.
@@
 These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}.
-However, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986}, though empirical studies suggest great water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}.
-Hence, amphistomy clearly should be favored when CO$_2$ limits photosynthetic rate, but the open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?
+On the other hand, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986, Buckley_etal_2015}.
+While transition to amphistomy is thus thought to increase transpiration, empirical studies suggest greater water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}.
+Hence, amphistomy appears to benefit a plant's carbon use relative to water loss and should be favored when CO$_2$ limits photosynthetic rate.
+The open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?
 
@@
 The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs.
-\citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitat (i.e.
-high light) of the British flora.
+\citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora.
 These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants.
@@
 Stomatal ratio is also associated with growth form.
-In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitat are amphistomatous.
+In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous.
 This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}.
@@
 Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another.
-Open habitat and growth form are probably not independent because open habitat by definition consists of more short-statured, herbaceous plants.
-Even attempts to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927} are problematic because high light habitats select for faster life history stratagies \citep{Galloway_Etterson_2007} [<-- OTHER REFS IN THAT PAPER?].
-Open habitat and groth form may also interact with one another.
-For example, amphistomy may only favored when CO$_2$ strongly limits photosynthetic rate \textit{and} photosynthetic rate strongly limits fitness.
+Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants.
+Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}.
+However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species.
+Finally, open habitat and groth form may also interact with one another.
+For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g.
+in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g.
+in herbs).
 
@@
 
-To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Fitter_Peat_1994a, BEF} using phylogenetic comparative methods.
+To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods.
 The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation.
 Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods.
-For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep{Kelly_Beerling_1995}.
-A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.
+For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}.
+A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.
+
+% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.
 
@@
 \section*{METHODS}
+
+Data and annotated source code to generate this manuscript are available on GitHub (https://github.com/cdmuir/britstom) and Dryad \citep{Muir_dryad}.
 

### 2017-07-13 ms/ms.tex
edited introduction

@@
 
+\begin{Schunk}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
+
 %--------------------------------------------------
@@
 In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 372 species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
@@
 
-To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Fitter_Peat_1994a, BEF} using phylogenetic comparative methods.
+To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods.
 The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation.
 Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods.
-For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep{Kelly_Beerling_1995}.
-A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.
+For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}.
+A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me for the first time to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.
+
+% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.
 

### 2017-07-13 ms/ms.Rnw
edited ms

@@
 % Finish abstract
-% Finish citations
-% Licenses (see http://ropensci.github.io/reproducibility-guide/sections/introduction/)
+% Finish citations (add citations only mentioned online to MS)
 
@@
 \newcommand{\el}{L-value} 
+\newcommand{\els}{L-values}
 
@@
 \section*{Acknowledgements}
-I thank Sally Otto and Matt Pennell for feedback on this manuscript.
+I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript.
+
 % \section*{Author contribution statement}
@@
 
-I obtained data on ab- and adaxial stomatal density on 370 species from British Ecological Flora \citep{Fitter_Peat_1994a, BEF}.
-I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
+Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.
@@
 I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105).
-Seventeen species (4.6\%) in the dataset were not present in the phylogeny.
-For nine species, I used the position a congeneric species as a proxy for the focal species.
-When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species.
-For the remaining eight missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree.
+\Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny.
+For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species.
+When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.
+For the remaining \Sexpr{nNotInPhy2} missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}.
 Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree.
@@
 
-I excluded data from hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig.
+I excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig.
 \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
-I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants..
+I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants.
 I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data.
 The final dataset contained \Sexpr{nrow(stomata)} species.
-The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.
+The R code accompanying this paper documents these decisions with citations to the relevant literature.
 
@@
 
-% NOTES
-
-%  papers on Ellenberg light value:
-%  Bartelheimer and Poschlod 2015
-%  Diekmann, 2003
-%  Ellenberg, 1974
-%  Ellenberg et al., 1991
-%  Silvertown et al., 2006
-
 \subsection*{Testing for an association between open habitat and growth form}
 
-I tested whether Raunki\ae r life form was associated \el~values among British angiosperms using ANOVA with Type-2 sum of squares.
+I tested whether Raunki\ae r life form was associated \el~among British angiosperms using ANOVA with Type-2 sum of squares.
 I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
-See the R code accompanying this paper for further detail \citep{Muir_dryad}.
-I predicted that species with faster life histories, especially therophytes (annuals), would have greater \el~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.
+See the R code accompanying this paper for further detail.
+I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.
 
@@
 Because some species had zero adaxial stomata, I added one to all values prior to log-transformation.
-For simplicity, I have defined $\mathrm{SR_{even}}$ here as the ratio of ad- to abaxial stomatal density because in most cases adaxial stomatal density is lower than abaxial (see Eq.~\ref{eq:SReven1}).
-The variance in $\mathrm{sr_{even}}$ can be decomposed into contributions of $\mathrm{sd_{ad}}$, $\mathrm{sd_{ab}}$, and their covariance:
+To make the variance decomposition calculations tractable, I have defined $\mathrm{SR_{even}}$ here as the ratio of ad- to abaxial stomatal density because in most cases adaxial stomatal density is lower than abaxial (see Eq.~\ref{eq:SReven1}).
+This was not done in previous analyses because I wanted to test what factors influenced the evenness of stomatal densities, regardless of which surface had higher density.
+With this modified form, the variance in $\mathrm{sr_{even}}$ can be decomposed into contributions of $\mathrm{sd_{ad}}$, $\mathrm{sd_{ab}}$, and their covariance:
 
@@
 To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}.
-I tested whether parameter estimates were significantly different than zero using $z$-scores.
+I tested whether parameter estimates were significantly different from zero using $z$-scores.
 
@@
 Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}).
-Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
+Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
 

### 2017-07-13 ms/ms.tex
edited ms

@@
+% Things to do before submission:
+% Sally's notes
+% Matt's notes
+% Remove code to copy refs
+% Github DOI?
+% Finish abstract
+% Finish citations (add citations only mentioned online to MS)
+
 % Notes from call with Rob
@@
 \section*{Acknowledgements}
+I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript.
 
@@
 For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}.
-A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me for the first time to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.
+A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.
 
@@
 
+Data and annotated source code to generate this manuscript are available on GitHub (https://github.com/cdmuir/britstom) and Dryad \citep{Muir_dryad}.
+
 \subsection*{Data on stomatal ratio, light habitat, growth form, and phylogenetic relationships}
 
-I obtained data on ab- and adaxial stomatal density on 370 species from British Ecological Flora \citep{Fitter_Peat_1994a, BEF}.
-I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
+Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.
@@
 I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105).
-Seventeen species (4.6\%) in the dataset were not present in the phylogeny.
-For nine species, I used the position a congeneric species as a proxy for the focal species.
-When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species.
-For the remaining eight missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree.
+14 species (3.5\%) in the dataset were not present in the phylogeny.
+For 8 of these species, I used the position a congeneric species as a proxy for the focal species.
+When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.
+For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}.
 Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree.
@@
 
-I excluded data from hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig.
+I excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig.
 \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
-I also excluded C$_4$ (3 species) and CAM (2 species) plants..
+I also excluded C$_4$ (3 species) and CAM (2 species) plants.
 I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
 The final dataset contained 372 species.
-The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.
+The R code accompanying this paper documents these decisions with citations to the relevant literature.
 
@@
 %  Diekmann, 2003
-%  Ellenberg, 1974
-%  Ellenberg et al., 1991
-%  Silvertown et al., 2006
 
@@
 I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
-See the R code accompanying this paper for further detail \citep{Muir_dryad}.
+See the R code accompanying this paper for further detail.
 I predicted that species with faster life histories, especially therophytes (annuals), would have greater \el~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.

### 2017-07-13 R/run-all.R

@@ -18,7 +18,4 @@ source("R/07-analyse-stomatal-ratio.R")
 source("R/08-analyse-stomatal-density.R")
 source("R/09-analyse-phyloSEM.R")
 
-knit("ms/ms.Rnw")
-
 # UPDATE BIBLIOGRAPHY
-# NEED TO KNIT REPORT

### 2017-07-13 r/01-process-data.R

@@ -156,3 +156,7 @@ rm(photo, plantatt, xa, xaa, xp, xpa)
 stomata$species %<>% str_replace_all(" ", "_")
 stomata$acceptedname %<>% str_replace_all(" ", "_")
 write_csv(stomata, str_c(path_proc_data, "/stomata.csv"))
+
+# Export objects to ms
+stomata_unfiltered <- stomata
+export2ms(c("stomata_unfiltered"))

### 2017-07-13 r/02-make-phylogeny.R

@@ -1,20 +1,31 @@
-source("R/header.R")
+source("r/header.R")
+
+stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"))
 
 ##### Import phylogeny of British Flora -----
 # from Lim et al 2014
 tmp <- read_lines(str_c(path_raw_data, "/S15105.nex"))
 l1 <- str_detect(tmp, "BEGIN CHARACTERS") %>% which() - 1
 l2 <- str_detect(tmp, "BEGIN TREES") %>% which() - 1
+
+# Lithospermum purpurocaeruleum is mispelled
+tmp %<>% str_replace_all("Lithospermum_purpureocaeruleum", "Lithospermum_purpurocaeruleum")
+
+# Changing Populus_nigra_sens.lat. to Populus nigra officinale 
+tmp %<>% str_replace_all("Populus_nigra_sens.lat.", "Populus_nigra")
+
+# Changing Rosa_canina_agg. to Rosa_canina officinale 
+tmp %<>% str_replace_all("Rosa_canina_agg.", "Rosa_canina")
+
 write_lines(tmp[c(1:l1, l2:length(tmp))], 
             path = str_c(path_raw_data, "/Lim_etal_2014.nex"))
 phy <- read_nexus_phylo(str_c(path_raw_data, "/Lim_etal_2014.nex"))
 
 ##### Check for species missing from phylogeny -----
-stomata <- read_csv(str_c(path_proc_data, "/stomata.csv")) %>%
-  mutate(species1 = species)
-stomata$species1 %<>% str_replace_all(" ", "_")
-notInPhy <- which(!stomata$species1 %in% phy$tip.label)
-stomata[notInPhy, ]; length(notInPhy)
+notInPhy <- which(!(stomata$species %in% phy$tip.label |
+                      stomata$acceptedname %in% phy$tip.label))
+stomata[notInPhy, ]
+nNotInPhy1 <- length(notInPhy)
 
 ##### Modify tip labels -----
 
@@ -24,26 +35,14 @@ tmp %<>% str_replace_all("Cerastium_brachypetalum", "Cerastium_cerastoides")
 # Using Coincya monensis as proxy for C. wrightii 
 tmp %<>% str_replace_all("Coincya_monensis", "Coincya_wrightii")
 
-# Using Geranium pusillum as proxy for G. versicolor based on similar divergence in Palazzesi et al. 2012 (Bio J of Linn Soc, 107(1): 67-85)
[…]

### 2017-07-13 ms/ms.Rnw
edited ms

@@
 % Things to do before submission:
-% Sally's notes
 % Matt's notes
@@
 % Finish citations (add citations only mentioned online to MS)
-
-% Notes from call with Rob
-% van Tienderen - sensitivity approach better for evolution
-% send list of species to rob
-
-% TASKS
-% what's going on with monocots?
-% in methods - mention stuff about Salisbury
-% leaf size
+% Finish conclusions
 
@@
 
-% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.
-
 % from previous paper
@@
 
-Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}).
+Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}).
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
 Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$.
-Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}).
+Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SRmultReg}).
 The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups.
@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
+% Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
 
@@
 The contributions of adaxial density, abaxial density, and their covariance are \Sexpr{round(varSDad / varSR, 2)}, \Sexpr{round(varSDab / varSR, 2)}, and \Sexpr{round(-2 * covSD / varSR, 2)}, respectively.
-Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive.
 This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.
@@
 The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $\Sexpr{pval2latex(exp(pvalSEM["logSDab ~~ logSDad"]))}).
-These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.
+These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.
 
@@
 Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}.
-However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are necessarily dominated by herbs -- or the fact that species are not independent because of shared evolutionaey hisory.
+However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory.
 Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods.
@@
 In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}).
-Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies \textit{and} photosynthesis strongly limits fitness.
-The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g.
-containing both annual and perennial forms).
-I predict that amphistomy will be favored much more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light.
+Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
+This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}.
+The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms).
+I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light.
 Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.
@@
 The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}.
-It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}.
-Since evaporative demand increases under high insolation, under these hypotheses we would expect plants in high light to be hypostomatous.
-Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands.
+It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface.
+Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous.
+Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}.
 Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}.
@@
 
-These observations from the British flora strongly support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015}, but also point out previously unknown subtleties.
+These observations from the British flora partially support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015} but also point out previously unknown subtleties.
 These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$).
-Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantiative variation.
+Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantitative variation.
 Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'.
@@
 Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio.
-The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$).
+The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution).
 This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal.
@@
 Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces.
+Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses.
 Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density.
-This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domesication \citep{Milla_etal_2013}; crops species tend to have higher adaxial stomatal density than their wild relatives.
-Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses.
+This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.
 
@@
 The first I refer to as the `real estate' hypothesis.
-In hypostomatous plants, the lower surface is already crowded with stomata, and hence plants must increase the real estate available for stomata by develping them on the upper surface whenever there is selection for greater stomatal density.
+In hypostomatous plants, the lower surface is already crowded with stomata, and hence plants must increase the real estate available for stomata by developing them on the upper surface whenever there is selection for greater stomatal density.
 When stomata are packed too densely on one surface, stomatal interference limits their functioning and hence may create a strong selective pressure for amphistomy \citep{Parlange_Waggoner_1970, Dow_etal_2014a}.
@@
  \setstretch{\stretchy}
-  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform shape stomatal ratio ($\mathrm{SR_{even}}$).
+  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$).
 I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$.
@@
   & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
-    1
+    null
   & \Sexpr{round(fitSRe$optpar, 2)}
@@
 Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data.
-Points have been jittered for visial clarity.}
+Points have been jittered for visual clarity.}
 \label{fig:SRmultReg}
@@
 The width of the grey polygons indicates the density of data.
-Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles.
+Length of grey polygon indicate the range of the data; the dot indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles.
 Points outside the polygons are statistical outliers.}

### 2017-07-13 ms/ms.tex
edited ms

@@
 
-% NOTES
-
-%  papers on Ellenberg light value:
-%  Bartelheimer and Poschlod 2015
-%  Diekmann, 2003
-
 \subsection*{Testing for an association between open habitat and growth form}
 
-I tested whether Raunki\ae r life form was associated \el~values among British angiosperms using ANOVA with Type-2 sum of squares.
+I tested whether Raunki\ae r life form was associated \el~among British angiosperms using ANOVA with Type-2 sum of squares.
 I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
 See the R code accompanying this paper for further detail.
-I predicted that species with faster life histories, especially therophytes (annuals), would have greater \el~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.
+I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.
 
@@
 Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
-Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
+Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
 

### 2017-07-14 ms/ms.Rnw
edited ms

@@
 % Remove code to copy refs
-% Github DOI?
-% Finish abstract
-% Finish citations (add citations only mentioned online to MS)
-% Finish conclusions
+% Remove notes
+% word count
 
@@
 
-\title{Light and life form interact to shape stomatal ratio among British angiosperms$^1$}
-\author{Christopher D. Muir$^2$}
+\title{Light and life form interact to shape stomatal ratio among British angiosperms}
+\author{Christopher D. Muir$^1$}
 \date{} % delete this line to display the current date
@@
 
-$^1$ Manuscript received \underline{     }; revision accepted \underline{     }. \\
-
-$^2$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
+$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
 \\
@@
 Word count: \\
-Summary:  \\
+Summary:  194\\
 Introduction:  \\
-Methods and Results:  \\
+Materials and Methods:  \\
+Results:  \\
 Discussion:  \\
-\# Figures and \# Tables, \# references
+Acknowledgement:  \\
+3 Figures and 1 Table, 1 Supplemental Figure
 
-<<Import-objects, echo = FALSE>>=
+<<Import-objects, echo = FALSE, results = hide>>=
 
@@
 
-%--------------------------------------------------
-% Acknowledgements and Contributions
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript.
-
-% \section*{Author contribution statement}
-% CDM designed the study, collected data, analyzed the data, and wrote the manuscript.
-
 \linenumbers
@@
 
-\section*{Abstract}
-
-% (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including key points of discussion?
+\section*{Summary}
 
@@
 	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy).
-Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood.
-In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
+High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
 	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item (return to this) key results: \el, growth form, and interaction are important
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and hemocryptophytic perennials, but less so in phanerophytes (mostly trees).
+Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
 	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
+These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
 \end{itemize}
-
-% POSSIBLE CONCLUSION: THIS TELLS US THAT CO2 LIMITS PHOTOSYNTHESIS AND HIGH LIGHT, BUT ONLY STRONG SELECTION IN FAST LIFE HISTORY (HERBS)...OR SOMETHING
 
@@
 
-\section*{INTRODUCTION}
+\section*{Introduction}
 
@@
 Natural selection shapes leaf anatomy in order to optimize its photosynthetic function in a given environment \citep{Haberlandt_1914, Givnish_1987, Smith_etal_1997}.
-By understanding the adaptive significance of leaf anatomical variation we can learn about natural history, find targets for crop improvement, and identify anatomical proxies for paleoclimates preserved in the fossil record [CITE].
-The size, density, and distribution of stomata on a leaf vary widely and impact functions like the maximum photosynthetic rate, water-use efficiency, photosynthetic nitrogen-use efficiency, and susceptibility to foliar pathogens that infect through stomata [CITATIONS].
-Hence, stomata have been especially useful in understanding plastic and evolutionary response to climate change and domestication (Royer, Ward, Woodward, Beerling, Milla et al...).
+By understanding the adaptive significance of leaf anatomical variation we can learn about natural history, find targets for crop improvement, and identify anatomical proxies for paleoclimates preserved in the fossil record \citep[e.g.][]{Wolfe_1971, Royer_2001, McElwain_Steinthorsdottir_2017}.
+The size, density, and distribution of stomata on a leaf vary widely and impact the flux of CO$_2$ and water vapour \citep[recently reviewed in][]{Sack_Buckley_2016}, as well as susceptibility to foliar pathogens that infect through stomata \citep{Mckown_etal_2014, Melotto_etal_2017}.
+Hence, stomata have been especially useful in understanding plastic and evolutionary response to climate change and domestication \citep{Woodward_1987, Beerling_Royer_2011, Milla_etal_2013}.
 
-While the density and size of stomata have been researched extensively [CITATIONS], the adaptive significance of stomatal distribution is less well understood.
+While the density and size of stomata have been researched extensively \citep[and references therin]{Sack_Buckley_2016}, the adaptive significance of stomatal distribution is less well understood.
 Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}.
@@
 The open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?
-
-% NOTES FOR FIRST TWO PARAGRAPHS
-% Stomatal ratio and photo/water use: Parkhurst papers, 
-% from Peat and Fitter 1994b "Jones (1985) found that amphistomaty may be advantageous as it allows greater CO2 transfer through the mesophyll, especially in thicker leaves, and greater leaf conductance in high light conditions when CO2 is limiting photosynthetic rate.
-However, his models showed that hypostomaty would be advantageous when there are high boundary layer resistances (for example in large leaves or deep within a plant canopy), high humidity or a temperature gradient across the leaf."
 
@@
 This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}.
-Across plants from ~90 familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
+Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
 These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates.
-\citep{Bazzaz_1979, Korner_etal_1989}.
-NEED MORE RECENT CITATIONS ON THIS.
-
-% NOTES FOR PARAGRAPHS 3-4:
-% From Mott et al 1982: ''Korner, Scheel \& Bauer (1979) have recently reviewed the literature on maximum stomatal conductance and photosynthetic rates for plants of specific habitats and growth forms.
-Low maximum photosynthetic rates are observed for succulents, evergreen conifers, deciduous woody plants, herbs from shady habitats, evergreen woody plants, and desert and steppe shrubs.
-Higher maximum values are obtained for deciduous fruit trees, wild grasses, cultivated C$_4$ and C$_3$ grasses, herbaceous crop plants, herbs from open habitats, and plants from aquatic habitats and swamps.
-Bazzaz (1979) cites evidence that the maximum photosynthetic rate declines through successional stages in a plant community, pointing to higher rates for open field plants and grasses and lower rates for canopy deciduous trees and shade habitat herbs.
-Early successional trees appear to be intermediate in photosynthetic rates.''
+\citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.
 
@@
 
-% from previous paper
-% Here, we examine whether stomata on the abaxial (`lower') surface of the leaf evolve independently of adaxial (`upper') stomata.
-Stomata are microscopic pores on the leaf surface formed by a pair of guard cells.
-The density, size, and arrangement of stomata on a leaf set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it \citep{Parkhurst_1978, Sack_etal_2003, Franks_Farquhar_2001, Galmes_etal_2013}.
-Hence, stomatal traits like density, size, and ratio of upper to lower stomata have strong effects on carbon assimilation and water-use efficiency.
-
-% The proportion of stomata found on the upper surface also tends to increase during domestication, even as the total stomatal density stays constant \citep{Milla_etal_2013}.
-Amphistomy increases CO$_2$ diffusion within the leaf by opening up a second parallel pathway in the intercellular airspace for diffusion from substomatal cavities to mesophyll cell walls.
-However, stomata on the upper surface in particular may be costly.
-For example, upper stomata increase the susceptibility to rust pathogens in \textit{Populus} \citep{Mckown_etal_2014}.
-Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits \citep{Buckley_etal_2015}.
-\cite{Muir_2015} reviewed the literature on other possible fitness costs.
-
 %--------------------------------------------------
-% Methods
+% Materials and Methods
 %--------------------------------------------------
 
-\section*{METHODS}
+\section*{Materials and Methods}
 
@@
 
-\section*{RESULTS}
+\section*{Results}
 
@@
 
-\section*{DISCUSSION}
+\section*{Discussion}
 
@@
 
-% phylo half-life estimate from Muir 2015 is 22my
-
 \subsection*{Why does adaxial stomatal density control stomatal ratio?}
@@
 Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio.
+A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}.
 
-% alternatives are different benefits and costs
-% leaf orientation \citep{Smith_etal_1998}, stomatal interference , leaf hydraulic conductance (Sack, Buckley, de Boer?), change in stomatal ratio may be important to CO2 in amphi species (Vaccinium myrttillus, \citep{Woodward_Bazzaz_1988})
+\subsection*{Conclusions}
 
-\subsection*{Conclusions - finish when analysis is complete}
+By revisiting this classic ecological dataset with modern phylogenetic comparative methods, I have shown that amphistomy is strongly associated with both light and growth form, but the interaction between these factors is also important.
+Furthermore, amphistomy and high stomatal density are closely connected in species from high light environments, suggesting selective for coordination between these traits.
 
-% From previous paper:
-% Adaptive evolution may be constrained if traits cannot evolve independently.
-In particular, if traits share developmental pathways, then they may be unable to respond differentially to selection.
-In this study, we examined whether stomata on the abaxial (lower) and adaxial (upper) surfaces can evolve independently.
-We adduce two new lines of evidence which suggest that stomatal function on each surface can readily respond to selection.
-First, species possess heritable variation that allows partially independent evolution of stomatal densities in response to selection; every study reviewed found loci which alter stomatal density on one surface but not the other.
-Second, the anatomical trait most closely connected to stomatal conductance, stomatal pore index \citep{Sack_etal_2003}, evolves independently on ab- and adaxial surfaces among wild tomato species.
-Together, these new lines of evidence demonstrate that natural selection on stomatal arrangement is not strongly constrained by development, although we lacked statistical power to detect weak constraint.
-It is therefore likely that variation in how stomatal conductance is partitioned between leaf surfaces is due to adaptive rather than nonadaptive forces.
+%--------------------------------------------------
+% Acknowledgements and Contributions
+%--------------------------------------------------
 
-% Indeed, much recent evidence indicates that selection finely tunes the ratio of stomata on the upper and lower leaf surface, although the adaptive significance of variation in stomatal ratio is unresolved.
-Stomatal ratio affects leaf function, increasing CO$_2$ diffusion \citep{Parkhurst_1978, Parkhurst_Mott_1990, Gutschick_1984b, Parkhurst_1994} and hydraulic conductance outside the xylem \citep{Buckley_etal_2015}.
-As predicted, amphistomy seems to be more common in circumstances when efficient CO$_2$ supply is important, such as high irradiance \citep{Mott_etal_1982, Gibson_1996, Smith_etal_1997, Jordan_etal_2014}, thick leaves \citep{Parkhurst_1978, Muir_etal_2014b}, herbaceous growth form \citep{Salisbury_1927, Muir_2015}, and domestication \citep{Milla_etal_2013}.
-Despite potential benefits of amphistomy, most plant species are hypostomatous, implying a fitness cost of upper stomata, such as increased infection by foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}.
-For example, `upside-down' (resupinate) leaves with the abaxial surface facing upward have re-evolved hypostomy \citep{Lyshede_2002}, strongly implying a cost of upward facing stomata.
+\section*{Acknowledgements}
+I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript. I was supported by an NSERC CREATE grant.
 
-% To optimally balance fitness costs and benefits, natural selection must be able to change stomatal traits on one surface independently of the other.
-The present study shows that this is likely true and strikingly consistent on micro- and macroevolutionary timescales.
-Among \textit{Populus trichocarpa} populations and \textit{Solanum} species, the ratio of adaxial to abaxial $\mathrm{SPI}$ ($\mathrm{SPI}$ ratio) evolves mostly by changes in stomatal density rather than guard cell size.
-Within \textit{Populus}, populations are more amphistomatous at Northern latitudes with shorter growing seasons that may select for faster carbon assimilation \citep{Mckown_etal_2014, Kaluthota_etal_2015}.
-Latitudinal variation \textit{Populus trichocarpa} is due mostly to adaptive variation in adaxial stomatal density \citep{Mckown_etal_2014, Porth_etal_2015}.
-Stomatal density rather than size may have responded more readily to selection because there is no genetic covariance between ab- and adaxial stomatal density, permitting independent evolution \citep{Porth_etal_2015}.
-In contrast ab- and adaxial guard cell length positively covary, likely constraining evolution.
-Similarly, we found that over macroevolutionary timescales most of the variation in $\mathrm{SPI}$ among wild tomato species is due to changes in adaxial stomatal density rather than size.
-Indeed, stomatal density on each surface evolved independently, whereas guard cell lengths positively covaried (Table~\ref{table:Table2}).
-Adaptive evolution will likely take advantage of traits that evolve independently because this minimizes antagonistic pleiotropy.
-In a previous study, we found that loci affecting adaxial stomatal density were likely fixed by selection, but we did not measure stomatal size \citep{Muir_etal_2014a}.
-Overall, patterns within and between species indicate that selection on $\mathrm{SPI}$ ratio leads to greater change in stomatal densities rather sizes on each surface.
-Based on the analysis here, we conclude that changing stomatal density on one surface incurs less cost than changing size because the former is less constrained by shared developmental pathways.
-
-% However, simulations show that model identification (e.g. Brownian motion versus Ornstein-Uhlenbeck) is usually correct, even when sample sizes are moderate \citep{Cressler_etal_2015, Ho_Ane_2014}.
+\section*{Author contribution statement}
+CDM designed the study, collected data, analyzed the data, and wrote the manuscript.
 
@@
 %--------------------------------------------------
-% Literature Cited
+% References
 %--------------------------------------------------

### 2017-07-14 ms/ms.tex
edited ms

@@
 % Things to do before submission:
-% Sally's notes
 % Matt's notes
 % Remove code to copy refs
-% Github DOI?
-% Finish abstract
-% Finish citations (add citations only mentioned online to MS)
-
-% Notes from call with Rob
-% van Tienderen - sensitivity approach better for evolution
-% send list of species to rob
-
-% TASKS
-% what's going on with monocots?
-% in methods - mention stuff about Salisbury
-% leaf size
+% Remove notes
+% word count
 
@@
 Word count: \\
-Summary:  \\
+Summary:  194\\
 Introduction:  \\
-Methods and Results:  \\
+Materials and Methods:  \\
+Results:  \\
 Discussion:  \\
-\# Figures and \# Tables, \# references
-
-\begin{Schunk}
-\begin{Soutput}
-[1] TRUE
-\end{Soutput}
-\end{Schunk}
-
-%--------------------------------------------------
-% Acknowledgements and Contributions
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript.
-
-% \section*{Author contribution statement}
-% CDM designed the study, collected data, analyzed the data, and wrote the manuscript.
+Acknowledgement:  \\
+3 Figures and 1 Table, 1 Supplemental Figure
 
@@
 	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy).
-Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood.
-In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
+High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
 	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item (return to this) key results: \el, growth form, and interaction are important
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and hemocryptophytic perennials, but less so in phanerophytes (mostly trees).
+Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
 	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
+These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
 \end{itemize}
-
-% POSSIBLE CONCLUSION: THIS TELLS US THAT CO2 LIMITS PHOTOSYNTHESIS AND HIGH LIGHT, BUT ONLY STRONG SELECTION IN FAST LIFE HISTORY (HERBS)...OR SOMETHING
 
@@
 This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}.
-Across plants from ~90 familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
+Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
 These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates.
-\citep{Bazzaz_1979, Korner_etal_1989}.
-NEED MORE RECENT CITATIONS ON THIS.
+\citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.
 
@@
 %--------------------------------------------------
-% Methods
+% Materials and Methods
 %--------------------------------------------------
 
-\section*{METHODS}
+\section*{Materials and Methods}
 
@@
 The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively.
-Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive.
 This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.
@@
 The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$).
-These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.
+These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.
 
@@
 Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio.
-The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = 1.5$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$).
+The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = 1.5$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution).
 This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal.
@@
   & 81.8 \\
-    1
+    null
   & 0.29

### 2017-07-14 R/header.R

@@ -9,16 +9,13 @@ if (compile_pathd8) {
 # Libraries
 library(ape)
 library(dplyr)
-library(knitr)
 library(lavaan)
 library(magrittr)
 library(phylolm)
 library(phytools)
 library(plyr)
 library(readr)
-library(rncl)
 library(Rphylopars)
-library(rstanarm)
 library(stringr)
 library(taxize)
 library(vioplot)

### 2017-07-14 R/install-packages.R

@@ -1,15 +1,12 @@
 install.packages("ape")
 install.packages("dplyr")
-install.packages("knitr")
 install.packages("lavaan")
 install.packages("magrittr")
 install.packages("phylolm")
 install.packages("phytools")
 install.packages("plyr")
 install.packages("readr")
-install.packages("rncl")
 install.packages("Rphylopars")
-install.packages("rstanarm")
 install.packages("stringr")
 install.packages("taxize")
 install.packages("vioplot")

### 2017-07-14 R/run-all.R

@@ -1,5 +1,4 @@
 # source("R/install-packages.R")
-# source("R/update-packages.R")
 
 # source("R/session-info.R")
 
@@ -17,5 +16,3 @@ source("R/06-analyse-light-vs-lifeform.R")
 source("R/07-analyse-stomatal-ratio.R")
 source("R/08-analyse-stomatal-density.R")
 source("R/09-analyse-phyloSEM.R")
-
-# UPDATE BIBLIOGRAPHY

### 2017-07-14 R/update-pacakges.R

@@ -1,1 +1,0 @@
-utils::update.packages(ask = FALSE)

### 2017-07-14 ms/ms.Rnw
edited methods

@@
 % Things to do before submission:
-% Matt's notes
 % Remove code to copy refs
@@
 \usepackage{topcapt}
-% \usepackage[labelformat=empty]{caption} % for empty caption labels
 \usepackage{caption}
@@
 \usepackage{multirow}
-% \usepackage[none]{hyphenat}
 
@@
 Word count: \\
-Summary:  194\\
+Summary:  193\\
 Introduction:  \\
@@
   # Remove before publication
-  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
-            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+  #file.copy(from = "~/Google Drive/CommonResources/refs.bib",
+  #          to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
@@
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and hemocryptophytic perennials, but less so in phanerophytes (mostly trees).
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and perennial herbs, but less so in phanerophytes (mostly trees).
 Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
@@
 
-%%%%% PARAGRAPH 1-2: general stuff about adaptation, stomata, stomatal ratio %%%%%
 Natural selection shapes leaf anatomy in order to optimize its photosynthetic function in a given environment \citep{Haberlandt_1914, Givnish_1987, Smith_etal_1997}.
@@
 
-%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
 The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs.
@@
 
-%%%% PARAGRAPH 5-6 %%%%%
 Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another.
@@
 Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree.
-I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.5-64 \citep{Revell_2012}.
+I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}.
 I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree.
@@
 For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution.
-I used a 10,000 parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
-Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models).
+I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
+% REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models).
 Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction.
@@
 
-I used two complementary phylogenetic methods to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution.
-The contribution of each can be formalized using standard variance decomposition methods as derived below.
+I used two related phylogenetic methods, variance decomposition and structural equation modeling (SEM), to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution.
+First, the contribution of each ab- versus adaxial stomatal density can be calculated using phylogenetic variance decomposition methods as derived below.
 Because stomatal density is highly skewed, I log-transformed values for normality:
@@
 To make the variance decomposition calculations tractable, I have defined $\mathrm{SR_{even}}$ here as the ratio of ad- to abaxial stomatal density because in most cases adaxial stomatal density is lower than abaxial (see Eq.~\ref{eq:SReven1}).
-This was not done in previous analyses because I wanted to test what factors influenced the evenness of stomatal densities, regardless of which surface had higher density.
-With this modified form, the variance in $\mathrm{sr_{even}}$ can be decomposed into contributions of $\mathrm{sd_{ad}}$, $\mathrm{sd_{ab}}$, and their covariance:
+This differs from analyses described above because in those I wanted to test what factors influenced the evenness of stomatal densities, regardless of which surface had higher density.
+With this modified form, the variance in $\mathrm{sr_{even}}$ can readily be decomposed into contributions of $\mathrm{sd_{ad}}$, $\mathrm{sd_{ab}}$, and their covariance:
 
@@
 
-I estimated the phylogenetic covariance matrix between \el, $\mathrm{sd_{ab}}$, and $\mathrm{sd_{ad}}$ using a multivariate Ornstein-Uhlenbeck model fit in \pkg{Rphylopars} version 0.2.9 \citep{Goolsby_etal_2016, Goolsby_etal_2017}.
+I did not use the raw covariance, but rather estimated the phylogenetic covariance matrix between \el, $\mathrm{sd_{ab}}$, and $\mathrm{sd_{ad}}$ using a multivariate Ornstein-Uhlenbeck model fit in \pkg{Rphylopars} version 0.2.9 \citep{Goolsby_etal_2016, Goolsby_etal_2017}.
 From the covariance matrix, I estimated the contribution of abaxial density, adaxial density, and their covariance as:
 
-\begin{equation} \label{eq:contribution}
-	\frac{\mathrm{Var(sd_{ad})}}{\mathrm{Var(sr_{even})}}, \frac{\mathrm{Var(sd_{ab})}}{\mathrm{Var(sr_{even})}},~\textrm{and}~\frac{\mathrm{Cov(sd_{ad}, sd_{ab})}}{\mathrm{Var(sr_{even})}},
+\begin{equation} \label{eq:contribution_ad}
+	\mathrm{Contribution~of~sd_{ad}} = \frac{\mathrm{Var(sd_{ad})}}{\mathrm{Var(sr_{even})}}
+\end{equation}
+
+\begin{equation} \label{eq:contribution_ab}
+ \mathrm{Contribution~of~sd_{ab}} = \frac{\mathrm{Var(sd_{ab})}}{\mathrm{Var(sr_{even})}}
+\end{equation}
+
+\begin{equation} \label{eq:contribution_cov}
+ \mathrm{Contribution~of~Cov(sd_{ad}, sd_{ab})} = \frac{\mathrm{Cov(sd_{ad}, sd_{ab})}}{\mathrm{Var(sr_{even})}}
 \end{equation}
@@
 
-I was interested in whether light-mediated evolution of stomatal ratio acted mostly by increasing adaxial stomatal density while maintaining abaxial density, or keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface.
+I also wanted to test whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface.
 The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$.
 The second scenario predicts that \el~acts similarly on both and that there is a negative covariance $\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$.
-I tested these competing predictions by fitting a simple phylogenetic structural equation model (SEM).
-The model uses the phylogenetic covariance matrix to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e.
+I tested these competing predictions by fitting a very simple phylogenetic SEM.
+The model uses the phylogenetic covariance matrix, as described above, to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e.
 estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$).
@@
 By revisiting this classic ecological dataset with modern phylogenetic comparative methods, I have shown that amphistomy is strongly associated with both light and growth form, but the interaction between these factors is also important.
-Furthermore, amphistomy and high stomatal density are closely connected in species from high light environments, suggesting selective for coordination between these traits.
+Furthermore, amphistomy and high stomatal density are closely connected in species from high light environments, suggesting selection for coordination between these traits.
 
@@
 Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes.
-The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples.
+The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples.
 Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset.

### 2017-07-14 ms/ms.tex
reran everything to check
edited methods

@@
 
-% NOTES FOR FIRST TWO PARAGRAPHS
-% Stomatal ratio and photo/water use: Parkhurst papers, 
-% from Peat and Fitter 1994b "Jones (1985) found that amphistomaty may be advantageous as it allows greater CO2 transfer through the mesophyll, especially in thicker leaves, and greater leaf conductance in high light conditions when CO2 is limiting photosynthetic rate.
-However, his models showed that hypostomaty would be advantageous when there are high boundary layer resistances (for example in large leaves or deep within a plant canopy), high humidity or a temperature gradient across the leaf."
-
-%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
 The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs.
@@
 
-% NOTES FOR PARAGRAPHS 3-4:
-% From Mott et al 1982: ''Korner, Scheel \& Bauer (1979) have recently reviewed the literature on maximum stomatal conductance and photosynthetic rates for plants of specific habitats and growth forms.
-Low maximum photosynthetic rates are observed for succulents, evergreen conifers, deciduous woody plants, herbs from shady habitats, evergreen woody plants, and desert and steppe shrubs.
-Higher maximum values are obtained for deciduous fruit trees, wild grasses, cultivated C$_4$ and C$_3$ grasses, herbaceous crop plants, herbs from open habitats, and plants from aquatic habitats and swamps.
-Bazzaz (1979) cites evidence that the maximum photosynthetic rate declines through successional stages in a plant community, pointing to higher rates for open field plants and grasses and lower rates for canopy deciduous trees and shade habitat herbs.
-Early successional trees appear to be intermediate in photosynthetic rates.''
-
-%%%% PARAGRAPH 5-6 %%%%%
 Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another.
@@
 
-% from previous paper
-% Here, we examine whether stomata on the abaxial (`lower') surface of the leaf evolve independently of adaxial (`upper') stomata.
-Stomata are microscopic pores on the leaf surface formed by a pair of guard cells.
-The density, size, and arrangement of stomata on a leaf set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it \citep{Parkhurst_1978, Sack_etal_2003, Franks_Farquhar_2001, Galmes_etal_2013}.
-Hence, stomatal traits like density, size, and ratio of upper to lower stomata have strong effects on carbon assimilation and water-use efficiency.
-
-% The proportion of stomata found on the upper surface also tends to increase during domestication, even as the total stomatal density stays constant \citep{Milla_etal_2013}.
-Amphistomy increases CO$_2$ diffusion within the leaf by opening up a second parallel pathway in the intercellular airspace for diffusion from substomatal cavities to mesophyll cell walls.
-However, stomata on the upper surface in particular may be costly.
-For example, upper stomata increase the susceptibility to rust pathogens in \textit{Populus} \citep{Mckown_etal_2014}.
-Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits \citep{Buckley_etal_2015}.
-\cite{Muir_2015} reviewed the literature on other possible fitness costs.
-
 %--------------------------------------------------

### 2017-07-14 r/run-all.R

@@ -1,18 +1,18 @@
-# source("R/install-packages.R")
+# source("r/install-packages.R")
 
-# source("R/session-info.R")
+# source("r/session-info.R")
 
 source("r/header.R")
 
 # Prepare data
-source("R/01-process-data.R")
-source("R/02-make-phylogeny.R")
-source("R/03-plot-lifeform.R")
-source("R/04-filter-data.R")
-source("R/05-modify-phylogeny.R")
+source("r/01-process-data.R")
+source("r/02-make-phylogeny.R")
+source("r/03-plot-lifeform.R")
+source("r/04-filter-data.R")
+source("r/05-modify-phylogeny.R")
 
 # Analyze data
-source("R/06-analyse-light-vs-lifeform.R")
-source("R/07-analyse-stomatal-ratio.R")
-source("R/08-analyse-stomatal-density.R")
-source("R/09-analyse-phyloSEM.R")
+source("r/06-analyse-light-vs-lifeform.R")
+source("r/07-analyse-stomatal-ratio.R")
+source("r/08-analyse-stomatal-density.R")
+source("r/09-analyse-phyloSEM.R")

### 2017-07-14 ms/ms.Rnw
edited ms

@@
 % Remove code to copy refs
-% Remove notes
 % word count
+% plain language summary
 
@@
 \usepackage{setspace}
-\newcommand{\stretchy}{2}
+\newcommand{\stretchy}{1.5}
 \usepackage{lineno}
@@
 
-\title{Light and life form interact to shape stomatal ratio among British angiosperms}
+\title{Light and growth form interact to shape stomatal ratio among British angiosperms}
 \author{Christopher D. Muir$^1$}
@@
 Word count: \\
-Summary:  193\\
+Summary:  190\\
 Introduction:  \\
@@
   # Remove before publication
-  #file.copy(from = "~/Google Drive/CommonResources/refs.bib",
-  #          to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
+            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
@@
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and perennial herbs, but less so in phanerophytes (mostly trees).
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees).
 Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
 	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
+These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
 \end{itemize}
@@
 
-Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, stomata, stomatal ratio
+Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r lifeform, stomata, stomatal ratio
 
@@
 Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion.
-Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}.
+Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}.
 Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.
@@
 Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
-These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates.
-\citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.
+These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.
 
@@
 
-The final question is important for telling whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume.
-If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be limited.
+The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume.
+If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small.
 In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well.
@@
 \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny.
-For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species.
+For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
 When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.
@@
 I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
+Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code).
 I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.
@@
 I used two related phylogenetic methods, variance decomposition and structural equation modeling (SEM), to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution.
-First, the contribution of each ab- versus adaxial stomatal density can be calculated using phylogenetic variance decomposition methods as derived below.
+First, the contribution of ab- versus adaxial stomatal density can be calculated using phylogenetic variance decomposition methods as derived below.
 Because stomatal density is highly skewed, I log-transformed values for normality:
@@
 The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$.
-The second scenario predicts that \el~acts similarly on both and that there is a negative covariance $\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$.
-I tested these competing predictions by fitting a very simple phylogenetic SEM.
+The second scenario predicts that \el~acts similarly on both and that there is a negative covariance ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$).
+I tested these competing predictions by fitting a very simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}.
 The model uses the phylogenetic covariance matrix, as described above, to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e.
@@
 Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}.
-However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory.
+However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory.
 Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods.
@@
 I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light.
-Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.
+Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.
 
@@
 
-The nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information.
+The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information.
 For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested.

### 2017-07-14 ms/ms.tex
edited ms

@@
 14 species (3.5\%) in the dataset were not present in the phylogeny.
-For 8 of these species, I used the position a congeneric species as a proxy for the focal species.
+For 8 of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
 When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.

### 2017-07-14 ms/ms.Rnw
finalized ms for biorxiv

@@
-% Things to do before submission:
-% Remove code to copy refs
-% word count
-% plain language summary
-
 \documentclass[12pt, oneside]{article}
@@
 Summary: 190\\
-Introduction:  \\
-Materials and Methods:  \\
-Results:  \\
-Discussion:  \\
-Acknowledgement:  \\
+Introduction: 1179\\
+Materials and Methods: 1176\\
+Results: 433\\
+Discussion: 1504\\
+Acknowledgement: 24\\
 3 Figures and 1 Table, 1 Supplemental Figure
@@
 <<Import-objects, echo = FALSE, results = hide>>=
-
-  # Remove before publication
-  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
-            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
@@
 \setstretch{\stretchy}
-% \pagenumbering{gobble}
+%\pagenumbering{gobble}
 
@@
 
-While the density and size of stomata have been researched extensively \citep[and references therin]{Sack_Buckley_2016}, the adaptive significance of stomatal distribution is less well understood.
+While the density and size of stomata have been researched extensively \citep[and references therein]{Sack_Buckley_2016}, the adaptive significance of stomatal distribution is less well understood.
 Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}.
@@
 These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants.
-Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}.
+Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, especially in thick leaves \citep{Jones_1985}.
 Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion.
@@
 Stomatal ratio is also associated with growth form.
-In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous.
+In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous.
 This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}.
-Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
+Across plants from $\sim 90$ families worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}.
 These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.
@@
 Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants.
-Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}.
+Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habitats \citep{Salisbury_1927}.
 However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species.
-Finally, open habitat and groth form may also interact with one another.
+Finally, open habitat and growth form may also interact with one another.
 For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g.
@@
 
-To better understand the adapative significance of stomatal ratio, I asked three main questions:
+To better understand the adaptive significance of stomatal ratio, I asked three main questions:
 
@@
 The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups.
-There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s.
+There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s.
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
@@
 Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}.
-However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory.
+However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary history.
 Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods.
@@
 In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}).
-Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
+Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
 This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}.
@@
 Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantitative variation.
-Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'.
+Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantitative indicator of vegetation type, rather than simply `open' or `closed'.
 A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

### 2017-07-14 ms/ms.Rnw
updated ms

@@
 I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
-Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.

### 2017-07-15 ms/ms.tex
updated ms

@@
 I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
-Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.
@@
 Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio.
-The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively.
+The contributions of adaxial density, abaxial density, and their covariance are 1.12, 0.38, and -0.5, respectively.
 This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively).
-However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1).
-Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
-The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.8 $\times10^{-7}$, respectively).
+However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12).
+Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.27-fold change in adaxial stomatal density versus a 1.13-fold change in abaxial stomatal density.
+The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $2.5 $\times10^{-10}$).
 These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.
@@
   & 12
-  & -33.3
-  & 90.6
+  & -33.2
+  & 90.4
   & 0 \\
     \el~+ lifeform
-  & 0.47
-  & 0.072
+  & 0.46
+  & 0.071
   & 0.32
   & 8
-  & -40.3
-  & 96.5
+  & -40.2
+  & 96.4
   & 6 \\
@@
   & 0.64
-  & 0.108
+  & 0.107
   & 0.26
@@
   & 126.6
-  & 36.1 \\
+  & 36.2 \\
     lifeform
@@
   & 172.4
-  & 81.8 \\
+  & 82 \\
     null
@@
   & -107.5
-  & 221
-  & 130.5 \\
+  & 221.1
+  & 130.7 \\
 	\bottomrule
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
+%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
 \caption{The effect of light on stomatal ratio depends on Raunki\ae r life form.

### 2017-07-15 r/06-analyse-light-vs-lifeform.R

@@ -22,7 +22,7 @@ lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyt
 names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
-pdf(str_c(path_figures, "/figure_lf-light.pdf"), 4, 7)
+pdf(str_c(path_figures, "/figure_lf-light.pdf"), 4, 7, useDingbats = FALSE)
 par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
 plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(0, 5), xlab = "", ylab = "", axes = F)  
 

### 2017-07-15 r/07-analyse-stomatal-ratio.R

@@ -33,7 +33,8 @@ fitSR_pl <- read_rds(str_c(path_objects, "/fitSR_pl.rds"))
 
 # Figure
 
-pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7)
+pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7, useDingbats = FALSE)
+#postscript(str_c(path_figures, "/figure_SRmultReg.ps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {
@@ -98,60 +99,3 @@ dev.off()
 
 # Export objects to ms
 export2ms(c("aicSR", "fitSRa", "fitSRb", "fitSRc", "fitSRd", "fitSRe"))
-
-# NOTE: Matt suggested deleting this, just using AIC
-# Parametric bootstrap statistical signicance of deltaAIC between model 
-# with interaction (fitSRa) and model without interaction (fitSRb)
-# set.seed(444294)
-# bootModB <- predict(fitSRb) + rTrait(n = 1e3, modAngioPhy, "OU", 
-#                                     list(sigma2 = fitSRb$sigma2, alpha = fitSRb$optpar))
-# colnames(bootModB) <- paste0("b", 1:ncol(bootModB))
-
-# fit model of model A to data simulated from model B
-# fitA2B <- lapply(colnames(bootModB), function(col) {
-#  stomata$Y <- bootModB[, col]
-#  phylolm(Y ~ lifeform * ellenberg_light, model = "OUrandomRoot", data = stomata, 
-#          phy = modAngioPhy, upper.bound = 10)
-#  })
-
-# fit model of model A to data simulated from model B
-# fitB2B <- lapply(colnames(bootModB), function(col) {
-#  stomata$Y <- bootModB[, col]
-#  phylolm(Y ~ lifeform + ellenberg_light, model = "OUrandomRoot", data = stomata, 
-#          phy = modAngioPhy, upper.bound = 10)
-#  })
-
-# refit models at parameter bounds
-# for (i in which(round(sapply(fitA2B, function(X) X$optpar), 2) == 10)) {
-#  stomata$Y <- bootModB[, colnames(bootModB)[i]]
[…]

### 2017-07-15 r/08-analyse-stomatal-density.R

@@ -29,11 +29,11 @@ dnsAd <- tapply(log10(stomata$ad_density + 1), stomata$ellenberg_light, function
   ret
 } )
 
-pdf(str_c(path_figures, "/figure_SD-light.pdf"), 8, 10)
+pdf(str_c(path_figures, "/figure_SD-light.pdf"), 8, 10, useDingbats = FALSE)
 par(mar = rep(0, 4), oma = c(5, 6, 1, 1), las = 1, mfrow = c(2, 1))
 plot(0, 0, xlim = c(2.5, 9.5), ylim = c(0, 3), type = "n", axes = F, frame.plot = T)
 
-mtext(1, text = "Ellenberg light indicator value", outer = T, cex = 2, line = 3)
+mtext(1, text = "Ellenberg light indicator", outer = T, cex = 2, line = 3)
 mtext(2, text = expression(paste("Stomatal density (no. ", mu, m^-2, ") [log scale]")), 
       outer = T, cex = 2, las = 0, line = 3)
 axis(2, at = 0:3, labels = c(1, 10, 100, 1000))

### 2017-09-26 ms/ms.Rnw
updated ms
adding phylo figure

@@
 \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny.
-For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
+For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
 When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.
@@
 
-I tested whether Raunki\ae r life form was associated \el~among British angiosperms using ANOVA with Type-2 sum of squares.
+I tested whether Raunki\ae r life form was associated with \el~among British angiosperms using ANOVA with Type-2 sum of squares.
 I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.

### 2017-09-26 r/XX-plot-phylogeny.R

@@ -1,0 +1,34 @@
+source("R/header.R")
+
+phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+
+# Arrange stomata in same order as phylogeny for plotting
+stomata %<>% extract(match(phy$tip.label, .$species), 1:ncol(.))
+
+# Add blank rows for internal node data
+blank_rows <- matrix(NA, nrow = nrow(stomata) - 1, ncol = ncol(stomata)) %>%
+  as.data.frame() %>%
+  set_colnames(colnames(stomata))
+
+stomata %<>% bind_rows(blank_rows)
+rm(blank_rows)
+
+##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
+
+lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
+names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
+lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+
+plot(phy, type = "fan", show.tip.label = FALSE)
+
+
+##### Scratch -----
+
+stomata$lifeform %<>% 
+  factor(levels = names(lf))
+gp <- ggtree(phy, layout = "circular") +
+  geom_tippoint(aes(color = stomata$lifeform), shape = 18, size = 3) +
+  scale_colour_brewer("Growth form", type= "seq", palette = "Greens") +
+  theme(legend.position = "right")
+print(gp)

### 2017-09-26 r/header.R

@@ -9,6 +9,7 @@ if (compile_pathd8) {
 # Libraries
 library(ape)
 library(dplyr)
+library(ggtree)
 library(lavaan)
 library(magrittr)
 library(phylolm)

### 2017-09-26 r/install-packages.R

@@ -1,5 +1,7 @@
 install.packages("ape")
+source("https://bioconductor.org/biocLite.R")
 install.packages("dplyr")
+biocLite("ggtree")
 install.packages("lavaan")
 install.packages("magrittr")
 install.packages("phylolm")

### 2017-09-27 ms/ms.tex
finished first draft of phylo figure

@@
 14 species (3.5\%) in the dataset were not present in the phylogeny.
-For 8 of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
+For 8 of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}.
 When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}.
@@
 I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
-The final dataset contained 372 species.
+The final dataset contained 372 species (Fig.~\ref{fig:phylo}).
 The R code accompanying this paper documents these decisions with citations to the relevant literature.
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.8 $\times10^{-7}$, respectively).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.9 $\times10^{-7}$, respectively).
 However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12).
@@
   & 3
-  & -107.5
+  & -107.6
   & 221.1
@@
 %--------------------------------------------------
+
+\begin{figure}[ht]
+\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo.pdf}}
+\caption{CAPTION} 
+\label{fig:phylo}
+\end{figure}
 

### 2017-09-27 r/06-plot-phylogeny.R

@@ -6,33 +6,12 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
 # Arrange stomata in same order as phylogeny for plotting
 stomata %<>% extract(match(phy$tip.label, .$species), 1:ncol(.))
 
-# Add blank rows for internal node data (this is for ggtree - remove if unused)
-# blank_rows <- matrix(NA, nrow = nrow(stomata) - 1, ncol = ncol(stomata)) %>%
-#   as.data.frame() %>%
-#   set_colnames(colnames(stomata))
-# 
-# stomata %<>% bind_rows(blank_rows)
-# rm(blank_rows)
-
-##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
-
 lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
 names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
 # Factor lifeform for plotting
 stomata$lifeform %<>% factor(levels = names(lf))
-
-##### Scratch -----
-
-# gp <- ggtree(phy, layout = "rectangular") +
-#   geom_tippoint(aes(color = stomata$lifeform), shape = 19, size = 3) +
-#   scale_colour_brewer("Growth form", type= "seq", palette = "Greens") +
-#   theme(legend.position = "right")
-# print(gp)
-
-
-#### I'm liking this appraoch. ggtree not flexible enough ####
 
 # Rescale phylogeny for plotting
 phy$edge.length %<>% divide_by(max(nodeHeights(phy)[, 2]))
@@ -65,6 +44,8 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   ## Legend
   clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
        grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
+  text(grconvertX(5.5, "in", "user"), grconvertY(0.85, "ndc", "user"),
+       labels = "Growth form", cex = 1.5)
   points(rep(grconvertX(5, "in", "user"), 5),
          seq(0.1, 0.9, 0.2) * 1.25,
[…]

### 2017-09-27 r/run-all.R

@@ -10,9 +10,10 @@ source("r/02-make-phylogeny.R")
 source("r/03-plot-lifeform.R")
 source("r/04-filter-data.R")
 source("r/05-modify-phylogeny.R")
+source("r/06-plot-phylogeny.R")
 
 # Analyze data
-source("r/06-analyse-light-vs-lifeform.R")
-source("r/07-analyse-stomatal-ratio.R")
-source("r/08-analyse-stomatal-density.R")
-source("r/09-analyse-phyloSEM.R")
+source("r/07-analyse-light-vs-lifeform.R")
+source("r/08-analyse-stomatal-ratio.R")
+source("r/09-analyse-stomatal-density.R")
+source("r/10-analyse-phyloSEM.R")

### 2017-10-01 ms/ms.Rnw
finished first draft of phylo figure
adding growth form comparison

@@
 I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
-Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+Following recent comparative analyses \citep[e.g.][]{Niinemets_Valladares_2006, Bartelheimer_Poschlod_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} as measures of light habitat.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.
-Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).
-Ellenberg light indicator values are hereafter abbreviated \el.
+Ellenberg light indicator values, hereafter abbreviated \el, have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).
+
+There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes.
+First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
+Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
+a species with both annual and perennial forms would be classified as a biennial here).
+Woody species were classified as shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m).
+Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}.
+[NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]
+
 I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105).
@@
 I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data.
-The final dataset contained \Sexpr{nrow(stomata)} species.
+The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo}).
 The R code accompanying this paper documents these decisions with citations to the relevant literature.
@@
 \begin{figure}[ht]
+\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
+\caption{CAPTION} 
+\label{fig:phylo}
+\end{figure}
+
+\begin{figure}[ht]
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}

### 2017-10-01 ms/ms.tex
adding growth form comparison

@@
 I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}.
-Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively.
+Following recent comparative analyses \citep[e.g.][]{Niinemets_Valladares_2006, Bartelheimer_Poschlod_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} as measures of light habitat.
 Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found.
-Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).
-Ellenberg light indicator values are hereafter abbreviated \el.
+Ellenberg light indicator values, hereafter abbreviated \el, have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).
+
+There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes.
+First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
+Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
+a species with both annual and perennial forms would be classified as a biennial here).
+Woody species were classified as shrubs (plant height less than 4 m) or trees (plant height greater than 4 m).
+Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}.
+[NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]
+
 I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105).
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo.pdf}}
+\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
 \caption{CAPTION} 

### 2017-10-01 r/01-process-data.R

@@ -194,9 +194,10 @@ plantatt$growthform[x] <- "perennial"
 x <- which(plantatt$woodiness == "w")
 plantatt$growthform[x] <- "shrub"
 
-# Reclassify as tree if height > 1 m
+# Reclassify as tree if height > 4 m
 message("need to determine height cutoff for trees, or alternative classification scheme")
-x <- which(plantatt$woodiness == "w" & plantatt$height >= 1000)
+shrub_cutoff <- 4
+x <- which(plantatt$woodiness == "w" & plantatt$height >= (shrub_cutoff * 1e3))
 plantatt$growthform[x] <- "tree"
 
 #summary(plantatt$growthform)
@@ -242,4 +243,4 @@ write_csv(stomata, str_c(path_proc_data, "/stomata.csv"))
 
 # Export objects to ms
 stomata_unfiltered <- stomata
-export2ms(c("stomata_unfiltered"))
+export2ms(c("shrub_cutoff", "stomata_unfiltered"))

### 2017-10-01 r/06-plot-phylogeny.R

@@ -14,7 +14,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                     ))
 
 # Arrange stomata in same order as phylogeny for plotting
-stomata %<>% extract(match(phy$tip.label, .$species), 1:ncol(.))
+stomata %<>% magrittr::extract(match(phy$tip.label, .$species), 1:ncol(.))
 
 gf <- c("tree", "shrub", "perennial", "biennial", "annual")
 
@@ -31,7 +31,7 @@ phy$edge.length %<>% divide_by(max(nodeHeights(phy)[, 2]))
 
 # Angle for calculating position of data in fan phylogeny
 theta_mid <- seq(0, 2 * pi, length.out = Ntip(phy) + 1)
-theta_mid %<>% extract(1:Ntip(phy))
+theta_mid %<>% magrittr::extract(1:Ntip(phy))
 theta_left <- theta_mid - pi / Ntip(phy)
 theta_right <- theta_mid + pi / Ntip(phy)
 

### 2017-10-01 r/08-analyse-stomatal-ratio.R

@@ -1,7 +1,133 @@
 source("R/header.R")
 
 phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
-stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
+                    col_types = cols(
+                      species = col_character(),
+                      ab_density = col_double(),
+                      ad_density = col_double(),
+                      sr_propAd = col_double(),
+                      sr_even = col_double(),
+                      lifeform = col_character(),
+                      ellenberg_light = col_integer(),
+                      growthform = col_character()
+                    ))
+
+##### Growth form and Ellenberg light indicator values versus sr_even -----
+
+gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+
+# Phylogenetic regression model comparison
+stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
+fitSR_gf_a <- phylolm(sr_even ~ growthform * ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_gf_b <- phylolm(sr_even ~ growthform + ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_gf_c <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_gf_d <- phylolm(sr_even ~ growthform, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_gf_e <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+aicSR_gf <- sapply(list(fitSR_gf_a, fitSR_gf_a, fitSR_gf_a, fitSR_gf_a, fitSR_gf_a), AIC)
+
[…]

### 2017-10-01 r/09-analyse-stomatal-density.R

@@ -1,14 +1,19 @@
 source("R/header.R")
 
 phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
-stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
+                    col_types = cols(
+                      species = col_character(),
+                      ab_density = col_double(),
+                      ad_density = col_double(),
+                      sr_propAd = col_double(),
+                      sr_even = col_double(),
+                      lifeform = col_character(),
+                      ellenberg_light = col_integer(),
+                      growthform = col_character()
+                    ))
 
-##### Stomatal density, light, and lifeform -----
-  
-lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", 
-        "phanerophyte", "therophyte")
-names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
-lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+##### Stomatal density and light -----
 
 # density of stomatal density by side and L-value
 dnsAb <- tapply(log10(stomata$ab_density + 1), stomata$ellenberg_light, function(X) {

### 2017-10-01 r/functions.R

@@ -1,5 +1,23 @@
 # R functions associated with Muir 2017
 
+# Convenience function to remove combinations of same growth form
+remove_diag <- function(df) {
+  
+  df <- df[which(df[, 1] != df[, 2]), ]
+  df
+  
+}
+
+# Compare bootstrap samples to determine whether slopes are significantly different
+compare_slopes <- function(s1, s2) {
+  d <- s1 - s2
+  ret <- mean(d)
+  ret %<>% c(median(d))
+  ret %<>% c(quantile(d, probs = c(0.025, 0.975)))
+  ret %<>% set_names(c("mean", "median", "lower", "upper"))
+  ret
+}
+
 # Draw 'histofan' around phylogeny
 draw_histofan <- function(y, r1, r2) {
   
@@ -24,7 +42,7 @@ draw_histofan <- function(y, r1, r2) {
          col = "grey", type = "l") # sr_even = 0
   points(cos(theta_mid) * r2, sin(theta_mid) * r2, col = "grey", type = "l") # sr_even = 0
 
-  theta_mid %<>% extract(1:n)
+  theta_mid %<>% magrittr::extract(1:n)
   theta_left <- theta_mid - pi / (n + 1)
   theta_right <- theta_mid + pi / (n + 1)
   

### 2017-10-01 r/header.R

@@ -1,6 +1,10 @@
 rm(list = ls())
 graphics.off()
 
+# remove before final publication
+file.copy("~/Google Drive/CommonResources/refs.bib",
+          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {
   system("cd PATHd8; cc PATHd8.c -O3 -lm -o PATHd8")
@@ -9,7 +13,6 @@ if (compile_pathd8) {
 # Libraries
 library(ape)
 library(dplyr)
-library(ggtree)
 library(lavaan)
 library(magrittr)
 library(phylolm)
@@ -21,6 +24,7 @@ library(RColorBrewer)
 library(Rphylopars)
 library(stringr)
 library(taxize)
+library(tidyr)
 library(vioplot)
 
 # Directories

### 2017-10-01 r/install-packages.R

@@ -1,7 +1,5 @@
 install.packages("ape")
-source("https://bioconductor.org/biocLite.R")
 install.packages("dplyr")
-biocLite("ggtree")
 install.packages("lavaan")
 install.packages("magrittr")
 install.packages("phylolm")
@@ -13,4 +11,5 @@ install.packages("RColorBrewer")
 install.packages("Rphylopars")
 install.packages("stringr")
 install.packages("taxize")
+install.packages("tidyr")
 install.packages("vioplot")

### 2017-10-04 r/01-process-data.R
added figure comparing growth forms

@@ -197,7 +197,7 @@ plantatt$growthform[x] <- "shrub"
 # Reclassify as tree if height > 4 m
 message("need to determine height cutoff for trees, or alternative classification scheme")
 shrub_cutoff <- 4
-x <- which(plantatt$woodiness == "w" & plantatt$height >= (shrub_cutoff * 1e3))
+x <- which(plantatt$woodiness == "w" & plantatt$height >= (shrub_cutoff * 1e2))
 plantatt$growthform[x] <- "tree"
 
 #summary(plantatt$growthform)

### 2017-10-04 r/05-plot-overlap.R

@@ -1,0 +1,34 @@
+source("r/header.R")
+
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"), 
+                    col_types = cols(
+                      species = col_character(),
+                      ab_density = col_double(),
+                      ad_density = col_double(),
+                      sr_propAd = col_double(),
+                      sr_even = col_double(),
+                      lifeform = col_character(),
+                      ellenberg_light = col_integer(),
+                      growthform = col_character()
+                    ))
+
+gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+
+lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
+names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
+lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+
+palette(brewer.pal(5, "Greens"))
+
+palette(brewer.pal(5, "Greens"))
+
+pdf(str_c(path_figures, "/figureS_compare-growth-forms.pdf"), 
+    width = 4, height = 6, useDingbats = FALSE)
+
+  par(mar = c(5, 5, 1, 1))
+  tab <- prop_table(stomata$lifeform, stomata$growthform)
+  tab <- tab[, names(lf)]
+  tab %<>% set_colnames(lf)
+  dotchart(tab, bg = 1:5, pt.cex = 1.5, xlab = "Proportion", cex.lab = 1.5)
+
+dev.off()

### 2017-10-04 r/functions.R

@@ -1,5 +1,56 @@
 # R functions associated with Muir 2017
 
+# Make cross tab table of growth form and life form
+prop_table <- function(x1, x2) {
+  
+  tab <- table(x1, x2) %>%
+    apply(1, function(X) X / sum(X)) %>% 
+    as.matrix()
+  tab
+  
+}
+
+# Draw connector to show overlap between Raunkiaer life form and growth form
+draw_connector <- function(x, y1, y2, col, nslice = 1e2) {
+  
+  # for debug
+  # x <- c(0.2, 0.8)
+  # y1 <- c(0.3, 0.5)
+  # y2 <- c(0.2, 0.6)
+  # nslice = 1e2
+  # col <- c("#EDF8E9", "#006D2C")
+  
+  stopifnot(length(x) == 2L)
+  stopifnot(length(y1) == 2L)
+  stopifnot(length(y2) == 2L)
+  stopifnot(length(col) == 2L)
+  
+  col <- colorRamp(col)(seq(0, 1, length.out = nslice))
+  
+  x_left <- seq(x[1], x[2], length.out = nslice + 1)
+  nudge <- diff(x_left)[1] / 10
+  x_right <- x_left[2:(nslice + 1)] %>% add(nudge)
+  x_left <- x_left[1:nslice] %>% subtract(nudge)
+  
+  slope_bottom <- (y2[1] - y1[1]) / (x[2] - x[1])
+  slope_top <- (y2[2] - y1[2]) / (x[2] - x[1])
+
+  y_bottom_left <- slope_bottom * (x_left - x[1]) + y1[1] 
+  y_bottom_right <- slope_bottom * (x_right - x[1]) + y1[1] 
+  y_top_left <- slope_top * (x_left - x[1]) + y1[2]
+  y_top_right <- slope_top * (x_right - x[1]) + y1[2]
+  
+  for (i in 1:nslice) {
+    
+    polygon(c(x_left[i], x_right[i], x_right[i], x_left[i]), 
+            c(y_bottom_left[i], y_bottom_right[i], y_top_right[i], y_top_left[i]), 
+            col = rgb(col[i, 1], col[i, 2], col[i, 3], maxColorValue = 255), 
+            border = NA, lty = 0)
+
+  }
+  
+}
+
 # Convenience function to remove combinations of same growth form
 remove_diag <- function(df) {
   

### 2017-10-04 r/run-all.R

@@ -9,11 +9,12 @@ source("r/01-process-data.R")
 source("r/02-make-phylogeny.R")
 source("r/03-plot-lifeform.R")
 source("r/04-filter-data.R")
-source("r/05-modify-phylogeny.R")
-source("r/06-plot-phylogeny.R")
+source("r/05-plot-overlap.R")
+source("r/06-modify-phylogeny.R")
+source("r/07-plot-phylogeny.R")
 
 # Analyze data
-source("r/07-analyse-light-vs-form.R")
-source("r/08-analyse-stomatal-ratio.R")
-source("r/09-analyse-stomatal-density.R")
-source("r/10-analyse-phyloSEM.R")
+source("r/08-analyse-light-vs-form.R")
+source("r/09-analyse-stomatal-ratio.R")
+source("r/10-analyse-stomatal-density.R")
+source("r/11-analyse-phyloSEM.R")

### 2017-10-04 ms/ms.Rnw
added figure comparing growth forms
modified table 1

@@
 Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio.
-The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution).
+The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSR_lf_a$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution).
 This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal.
@@
   \midrule  
-    \el~$\times$ lifeform
-  & \Sexpr{round(fitSRa$optpar, 2)}
-  & \Sexpr{round(fitSRa$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRa$fitted.values, fitSRa$fitted.values + fitSRa$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRa)) + 2}
-  & \Sexpr{round(fitSRa$logLik, 1)}
-  & \Sexpr{round(fitSRa$aic, 1)}
-  & \Sexpr{round(fitSRa$aic - min(aicSR), 1)} \\
-    \el~+ lifeform
-  & \Sexpr{round(fitSRb$optpar, 2)}
-  & \Sexpr{round(fitSRb$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRb$fitted.values, fitSRb$fitted.values + fitSRb$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRb)) + 2}
-  & \Sexpr{round(fitSRb$logLik, 1)}
-  & \Sexpr{round(fitSRb$aic, 1)}
-  & \Sexpr{round(fitSRb$aic - min(aicSR), 1)} \\
+    \el~$\times$ Raunki\ae r lifeform
+  & \Sexpr{round(fitSR_lf_a$optpar, 2)}
+  & \Sexpr{round(fitSR_lf_a$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_lf_a$fitted.values, fitSR_lf_a$fitted.values + fitSR_lf_a$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_lf_a)) + 2}
+  & \Sexpr{round(fitSR_lf_a$logLik, 1)}
+  & \Sexpr{round(fitSR_lf_a$aic, 1)}
+  & \Sexpr{round(fitSR_lf_a$aic - min(aicSR), 1)} \\
+    \el~$\times$ growth form
+  & \Sexpr{round(fitSR_gf_f$optpar, 2)}
+  & \Sexpr{round(fitSR_gf_f$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_gf_f$fitted.values, fitSR_gf_f$fitted.values + fitSR_gf_f$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_gf_f)) + 2}
+  & \Sexpr{round(fitSR_gf_f$logLik, 1)}
+  & \Sexpr{round(fitSR_gf_f$aic, 1)}
+  & \Sexpr{round(fitSR_gf_f$aic - min(aicSR), 1)} \\
+    \el~+ Raunki\ae r lifeform
+  & \Sexpr{round(fitSR_lf_b$optpar, 2)}
+  & \Sexpr{round(fitSR_lf_b$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_lf_b$fitted.values, fitSR_lf_b$fitted.values + fitSR_lf_b$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_lf_b)) + 2}
+  & \Sexpr{round(fitSR_lf_b$logLik, 1)}
+  & \Sexpr{round(fitSR_lf_b$aic, 1)}
+  & \Sexpr{round(fitSR_lf_b$aic - min(aicSR), 1)} \\
+     \el~+ growth form
+  & \Sexpr{round(fitSR_gf_g$optpar, 2)}
+  & \Sexpr{round(fitSR_gf_g$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_gf_g$fitted.values, fitSR_gf_g$fitted.values + fitSR_gf_g$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_gf_g)) + 2}
+  & \Sexpr{round(fitSR_gf_g$logLik, 1)}
+  & \Sexpr{round(fitSR_gf_g$aic, 1)}
+  & \Sexpr{round(fitSR_gf_g$aic - min(aicSR), 1)} \\
+    Raunki\ae r lifeform
+  & \Sexpr{round(fitSR_lf_d$optpar, 2)}
+  & \Sexpr{round(fitSR_lf_d$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_lf_d$fitted.values, fitSR_lf_d$fitted.values + fitSR_lf_d$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_lf_d)) + 2}
+  & \Sexpr{round(fitSR_lf_d$logLik, 1)}
+  & \Sexpr{round(fitSR_lf_d$aic, 1)}
+  & \Sexpr{round(fitSR_lf_d$aic - min(aicSR), 1)} \\
+    growth form
+  & \Sexpr{round(fitSR_gf_h$optpar, 2)}
+  & \Sexpr{round(fitSR_gf_h$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_gf_h$fitted.values, fitSR_gf_h$fitted.values + fitSR_gf_h$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_gf_h)) + 2}
+  & \Sexpr{round(fitSR_gf_h$logLik, 1)}
+  & \Sexpr{round(fitSR_gf_h$aic, 1)}
+  & \Sexpr{round(fitSR_gf_h$aic - min(aicSR), 1)} \\
     \el
@@
   & \Sexpr{round(fitSRc$aic - min(aicSR), 1)} \\
-    lifeform
-  & \Sexpr{round(fitSRd$optpar, 2)}
-  & \Sexpr{round(fitSRd$sigma2, 3)}
-  & \Sexpr{round(cor(fitSRd$fitted.values, fitSRd$fitted.values + fitSRd$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSRd)) + 2}
-  & \Sexpr{round(fitSRd$logLik, 1)}
-  & \Sexpr{round(fitSRd$aic, 1)}
-  & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
     null
@@
 \begin{figure}[ht]
+\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
+\caption{CAPTION}
+\label{fig:compare-growth-forms}
+\end{figure}
+
+\begin{figure}[ht]
 \centerline{\includegraphics{figures/figureS_violin.pdf}}

### 2017-10-04 r/09-analyse-stomatal-ratio.R

@@ -13,30 +13,139 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       growthform = col_character()
                     ))
 
+##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
+
+lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
+names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
+lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
+
+# Phylogenetic regression model comparison
+stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
+fitSR_lf_a <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_lf_b <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSR_lf_d <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
+                      data = stomata, phy = phy, upper.bound = 10)
+
+# Parametric bootstrap CIs of full model for p-values
+# set.seed(36502)
+# fitSR_lf_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, 
+#                        model = "OUrandomRoot", data = stomata, phy = phy, 
+#                        boot = 1e4, upper.bound = 10)
+# write_rds(fitSR_lf_pl, path = str_c(path_objects, "/fitSR_lf_pl.rds"))
+fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
+
+# Figure
+
[…]

### 2017-10-04 ms/ms.Rnw
updated phylo methods section

@@
 There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes.
-First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
-Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
+First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
+Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
 a species with both annual and perennial forms would be classified as a biennial here).
-Woody species were classified as shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m).
-Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}.
-[NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]
+Woody species are shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m).
+Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}.
+I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig.
+\ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
+The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).
 
@@
 
-I excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig.
-\ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
-I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants.
+I excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants.
 I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data.
-The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo}).
-The R code accompanying this paper documents these decisions with citations to the relevant literature.
+The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}).
+The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.
 
@@
 
-I tested whether Raunki\ae r life form was associated with \el~among British angiosperms using ANOVA with Type-2 sum of squares.
-I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
-See the R code accompanying this paper for further detail.
-I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.
+I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms.
+I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).
+I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
+However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail).
+Specifically, the estimated $\alpha$ parameter was extremely high.
+In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e.
+phylogenetic signal).
+When there is no phylogenetic signal (i.e.
+high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates.
+Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}.
+Therefore, I used a regular ANOVA with Type-2 sum of squares.
 
@@
 
-I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
+I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
 Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code).
@@
 \caption{CAPTION} 
-\label{fig:phylo}
+\label{fig:phylo-lifeform}
 \end{figure}
@@
 \begin{figure}[ht]
-\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
-\caption{CAPTION}
-\label{fig:compare-growth-forms}
-\end{figure}
-
-\begin{figure}[ht]
 \centerline{\includegraphics{figures/figureS_violin.pdf}}
@@
 
+\begin{figure}[ht]
+\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
+\caption{CAPTION}
+\label{fig:compare-growth-forms}
+\end{figure}
+
+\begin{figure}[ht]
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_gf-light.pdf}}
+\caption{Growth forms have different tolerances for sun and shade among British angiosperms.
+Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different growth forms.
+Height of the bars indicate the raw proportion of species in each bin for that lifeform.
+The sample size for each lifeform is listed next in parentheses.
+The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.}
+\label{fig:gf-light}
+\end{figure}
+
 \end{document}

### 2017-10-04 ms/ms.tex
added figure comparing growth forms
modified table 1
updated phylo methods section

@@
 There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes.
-First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
-Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
+First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
+Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
 a species with both annual and perennial forms would be classified as a biennial here).
-Woody species were classified as shrubs (plant height less than 4 m) or trees (plant height greater than 4 m).
-Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}.
-[NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]
+Woody species are shrubs (plant height less than 4 m) or trees (plant height greater than 4 m).
+Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}.
+I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig.
+\ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
+The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).
 
@@
 
-I excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig.
-\ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only.
-I also excluded C$_4$ (3 species) and CAM (2 species) plants.
+I excluded C$_4$ (3 species) and CAM (2 species) plants.
 I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
-The final dataset contained 372 species (Fig.~\ref{fig:phylo}).
-The R code accompanying this paper documents these decisions with citations to the relevant literature.
+The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}).
+The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.
 
@@
   \midrule  
-    \el~$\times$ lifeform
+    \el~$\times$ Raunki\ae r lifeform
   & 0.46
@@
   & 0 \\
-    \el~+ lifeform
+    \el~$\times$ growth form
+  & 0.46
+  & 0.07
+  & 0.32
+  & 12
+  & -38.2
+  & 100.4
+  & 9.9 \\
+    \el~+ Raunki\ae r lifeform
   & 0.46
@@
   & 6 \\
+     \el~+ growth form
+  & 0.51
+  & 0.08
+  & 0.31
+  & 8
+  & -43.4
+  & 102.7
+  & 12.3 \\
+    Raunki\ae r lifeform
+  & 0.34
+  & 0.067
+  & 0.15
+  & 7
+  & -79.2
+  & 172.4
+  & 82 \\
+    growth form
+  & 0.35
+  & 0.069
+  & 0.13
+  & 7
+  & -82.5
+  & 179.1
+  & 88.6 \\
     \el
@@
   & 36.2 \\
-    lifeform
-  & 0.34
-  & 0.067
-  & 0.15
-  & 7
-  & -79.2
-  & 172.4
-  & 82 \\
     null

### 2017-10-04 r/header.R

@@ -2,8 +2,8 @@ rm(list = ls())
 graphics.off()
 
 # remove before final publication
-file.copy("~/Google Drive/CommonResources/refs.bib",
-          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+# file.copy("~/Google Drive/CommonResources/refs.bib",
+#           "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {

### 2017-10-04 ms/ms.Rnw
updated light-growth form results

@@
 I did not use the raw covariance, but rather estimated the phylogenetic covariance matrix between \el, $\mathrm{sd_{ab}}$, and $\mathrm{sd_{ad}}$ using a multivariate Ornstein-Uhlenbeck model fit in \pkg{Rphylopars} version 0.2.9 \citep{Goolsby_etal_2016, Goolsby_etal_2017}.
+The phylogenetic covariance measures how strongly a set of traits evolve together over macroevolutionary timescales.
 From the covariance matrix, I estimated the contribution of abaxial density, adaxial density, and their covariance as:
@@
 I tested these competing predictions by fitting a very simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}.
-The model uses the phylogenetic covariance matrix, as described above, to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e.
+In general, SEMs attempt to determine whether variables are related causally or whether a relationship is mediated by another correlated variable.
+Phylogenetic SEMs use the phylogenetic covariance matrix, as described above, rather than the raw covariance.
+Here, I used a phylogenetic SEM to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e.
 estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$).
@@
 
-\subsection*{Light tolerance varies among Raunki\ae r life forms}
+\subsection*{Light tolerance varies among growth forms}
 
-Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}).
-Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
+Ellenberg light indicator values (\el) differed significantly among growth forms.
+Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}).
+Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{\Sexpr{aovGFvEL$Df[1]}, \Sexpr{aovGFvEL$Df[2]}}$ = \Sexpr{round(aovGFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovGFvEL$`Pr(>F)`[1])})
 

### 2017-10-04 ms/ms.tex
updated light-growth form results

@@
 
-\subsection*{Light tolerance varies among Raunki\ae r life forms}
+\subsection*{Light tolerance varies among growth forms}
 
-Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
-Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).
+Ellenberg light indicator values (\el) differed significantly among growth forms.
+Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
+Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~ than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)
 

### 2017-10-04 r/run-all.R

@@ -1,7 +1,5 @@
 # source("r/install-packages.R")
 
-# source("r/session-info.R")
-
 source("r/header.R")
 
 # Prepare data

### 2017-10-05 ms/ms.Rnw
updated results

@@
 
-\subsection*{Interactions between light and Raunki\ae r life form determine stomatal ratio}
+\subsection*{Interactions between light and growth form determine stomatal ratio}
 
-Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}).
+Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
+When classified as annual, biennial, perennial, shrub, or tree, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_gf_g) - AIC(fitSR_gf_f), 1)}$; Fig.~\ref{fig:SR-gf}).
+Raunki\ae r life form explained variation in stomatal ratio best (lower AIC; Table~\ref{table:srAIC}) , therefore we focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
 Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$.
-Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SRmultReg}).
+Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}).
 The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups.
@@
 Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantitative indicator of vegetation type, rather than simply `open' or `closed'.
-A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).
+A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SR-lf}).
 
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
 %\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
@@
 Points have been jittered for visual clarity.}
-\label{fig:SRmultReg}
+\label{fig:SR-lf}
 \end{figure}
@@
 
+\begin{figure}[ht]
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-gf.pdf}}
+%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
+\caption{The effect of light on stomatal ratio depends on growth form.
+Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in annual, biennual, and perennial herbs, but not shrubs or trees.
+The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples.
+Numbers in parentheses next to growth form are the sample sizes in the final dataset.
+Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data.
+Points have been jittered for visual clarity.}
+\label{fig:SR-gf}
+\end{figure}
 \end{document}

### 2017-10-05 ms/ms.tex
updated results

@@
 Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
-Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~ than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)
+Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)
 
-\subsection*{Interactions between light and Raunki\ae r life form determine stomatal ratio}
+\subsection*{Interactions between light and growth form determine stomatal ratio}
 
-Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}).
+Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
+When classified as annual, biennial, perennial, shrub, or tree, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-gf}).
+Raunki\ae r life form explained variation in stomatal ratio best (lower AIC; Table~\ref{table:srAIC}) , therefore we focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).

### 2017-10-05 r/09-analyse-stomatal-ratio.R

@@ -42,7 +42,7 @@ fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
 
 # Figure
 
-pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7, useDingbats = FALSE)
+pdf(str_c(path_figures, "/figure_SR-lf.pdf"), 4, 7, useDingbats = FALSE)
 #postscript(str_c(path_figures, "/figure_SRmultReg.ps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
@@ -150,8 +150,8 @@ fitSR_gf_pl <- read_rds(str_c(path_objects, "/fitSR_gf_pl.rds"))
 
 # Figure
 
-pdf(str_c(path_figures, "/figure_SRmultReg.pdf"), 4, 7, useDingbats = FALSE)
-#postscript(str_c(path_figures, "/figure_SRmultReg.ps"), width = 4, height = 7) # need for submitting to journal website
+pdf(str_c(path_figures, "/figure_SR-gf.pdf"), 4, 7, useDingbats = FALSE)
+#postscript(str_c(path_figures, "/figure_SR-gf.ps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {

### 2017-10-06 ms/ms.Rnw
changed plant habit terminology

@@
 
-There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes.
-First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree).
+There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes based on plant habit and Raunki\ae r life form.
+First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form based on habit.
+I categorized herbaceous species as annual, biennial, or perennial and woody species as shrub or tree.
 Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g.
@@
 I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.
-I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC).
+I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC).
 For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution.
@@
 
-I also wanted to test whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface.
+I also tested whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface.
 The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$.
@@
 Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{\Sexpr{aovLFvEL$Df[1]}, \Sexpr{aovLFvEL$Df[2]}}$ = \Sexpr{round(aovLFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovLFvEL$`Pr(>F)`[1])}).
-Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{\Sexpr{aovGFvEL$Df[1]}, \Sexpr{aovGFvEL$Df[2]}}$ = \Sexpr{round(aovGFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovGFvEL$`Pr(>F)`[1])})
+Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:hf-light}; ANOVA - $F_{\Sexpr{aovHFvEL$Df[1]}, \Sexpr{aovHFvEL$Df[2]}}$ = \Sexpr{round(aovHFvEL$`F value`[1], 1)}, $P$ = \Sexpr{pval2latex(aovHFvEL$`Pr(>F)`[1])})
 
@@
 Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
-When classified as annual, biennial, perennial, shrub, or tree, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_gf_g) - AIC(fitSR_gf_f), 1)}$; Fig.~\ref{fig:SR-gf}).
-Raunki\ae r life form explained variation in stomatal ratio best (lower AIC; Table~\ref{table:srAIC}) , therefore we focus hereafter on those analyses.
+When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}).
+Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
 However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary history.
-Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods.
-As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}).
-Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms.
+By bringing together datasets on stomata, light tolerance, growth form, and phylogeny of British angiosperms, I tested new hypotheses and reevaluated previous results using modern phylogenetic comparative methods.
+As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Fig.~\ref{fig:lf-light}, Fig.~\ref{fig:hf-light}).
+Nevertheless, both \el~and growth form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms.
+Finally, I show for the first time that adaxial stomatal density in particular accounts for most of the coordinated evolution between light tolerance and stomatal density.
 These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.
@@
 
+% Revise and put after 2nd sentence: Although I repeated certain analyses from previous papers, this is an important advance because previous analyses did not include phylogeny or did so improperly...
 Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy.
-In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}).
+In British angiosperms, both factors are important, though light may be a stronger determinant than growth form based on information criterion comparisons (Table~\ref{table:srAIC}).
 Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
@@
     \el~$\times$ growth form
-  & \Sexpr{round(fitSR_gf_f$optpar, 2)}
-  & \Sexpr{round(fitSR_gf_f$sigma2, 3)}
-  & \Sexpr{round(cor(fitSR_gf_f$fitted.values, fitSR_gf_f$fitted.values + fitSR_gf_f$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSR_gf_f)) + 2}
-  & \Sexpr{round(fitSR_gf_f$logLik, 1)}
-  & \Sexpr{round(fitSR_gf_f$aic, 1)}
-  & \Sexpr{round(fitSR_gf_f$aic - min(aicSR), 1)} \\
+  & \Sexpr{round(fitSR_hf_f$optpar, 2)}
+  & \Sexpr{round(fitSR_hf_f$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_hf_f$fitted.values, fitSR_hf_f$fitted.values + fitSR_hf_f$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_hf_f)) + 2}
+  & \Sexpr{round(fitSR_hf_f$logLik, 1)}
+  & \Sexpr{round(fitSR_hf_f$aic, 1)}
+  & \Sexpr{round(fitSR_hf_f$aic - min(aicSR), 1)} \\
     \el~+ Raunki\ae r lifeform
@@
      \el~+ growth form
-  & \Sexpr{round(fitSR_gf_g$optpar, 2)}
-  & \Sexpr{round(fitSR_gf_g$sigma2, 3)}
-  & \Sexpr{round(cor(fitSR_gf_g$fitted.values, fitSR_gf_g$fitted.values + fitSR_gf_g$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSR_gf_g)) + 2}
-  & \Sexpr{round(fitSR_gf_g$logLik, 1)}
-  & \Sexpr{round(fitSR_gf_g$aic, 1)}
-  & \Sexpr{round(fitSR_gf_g$aic - min(aicSR), 1)} \\
+  & \Sexpr{round(fitSR_hf_g$optpar, 2)}
+  & \Sexpr{round(fitSR_hf_g$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_hf_g$fitted.values, fitSR_hf_g$fitted.values + fitSR_hf_g$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_hf_g)) + 2}
+  & \Sexpr{round(fitSR_hf_g$logLik, 1)}
+  & \Sexpr{round(fitSR_hf_g$aic, 1)}
+  & \Sexpr{round(fitSR_hf_g$aic - min(aicSR), 1)} \\
     Raunki\ae r lifeform
@@
     growth form
-  & \Sexpr{round(fitSR_gf_h$optpar, 2)}
-  & \Sexpr{round(fitSR_gf_h$sigma2, 3)}
-  & \Sexpr{round(cor(fitSR_gf_h$fitted.values, fitSR_gf_h$fitted.values + fitSR_gf_h$residuals) ^ 2, 2)}
-  & \Sexpr{length(coef(fitSR_gf_h)) + 2}
-  & \Sexpr{round(fitSR_gf_h$logLik, 1)}
-  & \Sexpr{round(fitSR_gf_h$aic, 1)}
-  & \Sexpr{round(fitSR_gf_h$aic - min(aicSR), 1)} \\
+  & \Sexpr{round(fitSR_hf_h$optpar, 2)}
+  & \Sexpr{round(fitSR_hf_h$sigma2, 3)}
+  & \Sexpr{round(cor(fitSR_hf_h$fitted.values, fitSR_hf_h$fitted.values + fitSR_hf_h$residuals) ^ 2, 2)}
+  & \Sexpr{length(coef(fitSR_hf_h)) + 2}
+  & \Sexpr{round(fitSR_hf_h$logLik, 1)}
+  & \Sexpr{round(fitSR_hf_h$aic, 1)}
+  & \Sexpr{round(fitSR_hf_h$aic - min(aicSR), 1)} \\
     \el
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_gf-light.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_hf-light.pdf}}
 \caption{Growth forms have different tolerances for sun and shade among British angiosperms.
@@
 The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.}
-\label{fig:gf-light}
+\label{fig:hf-light}
 \end{figure}
@@
 \begin{figure}[ht]
-\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-gf.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
 %\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
@@
 Points have been jittered for visual clarity.}
-\label{fig:SR-gf}
+\label{fig:SR-hf}
 \end{figure}

### 2017-10-06 ms/ms.tex
changed plant habit terminology

@@
 Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$).
-Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)
+Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:hf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)
 
@@
 Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
-When classified as annual, biennial, perennial, shrub, or tree, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-gf}).
-Raunki\ae r life form explained variation in stomatal ratio best (lower AIC; Table~\ref{table:srAIC}) , therefore we focus hereafter on those analyses.
+When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}).
+Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
 Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio.
-The contributions of adaxial density, abaxial density, and their covariance are 1.12, 0.38, and -0.5, respectively.
+The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively.
 This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.9 $\times10^{-7}$, respectively).
-However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12).
-Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.27-fold change in adaxial stomatal density versus a 1.13-fold change in abaxial stomatal density.
-The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $2.5 $\times10^{-10}$).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively).
+However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1).
+Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
+The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$).
 These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.
@@
   & 12
-  & -33.2
-  & 90.4
+  & -33.3
+  & 90.6
   & 0 \\
@@
   & -38.2
-  & 100.4
-  & 9.9 \\
+  & 100.3
+  & 9.8 \\
     \el~+ Raunki\ae r lifeform
-  & 0.46
-  & 0.071
+  & 0.47
+  & 0.072
   & 0.32
   & 8
-  & -40.2
-  & 96.4
+  & -40.3
+  & 96.5
   & 6 \\
@@
   & 8
-  & -43.4
+  & -43.3
   & 102.7
-  & 12.3 \\
+  & 12.1 \\
     Raunki\ae r lifeform
@@
   & 172.4
-  & 82 \\
+  & 81.8 \\
     growth form
@@
   & -82.5
-  & 179.1
-  & 88.6 \\
+  & 178.9
+  & 88.4 \\
     \el
   & 0.64
-  & 0.107
+  & 0.108
   & 0.26
@@
   & 126.6
-  & 36.2 \\
+  & 36.1 \\
     null
@@
   & 3
-  & -107.6
-  & 221.1
-  & 130.7 \\
+  & -107.5
+  & 221
+  & 130.5 \\
 	\bottomrule

### 2017-10-06 r/01-process-data.R

@@ -166,13 +166,13 @@ plantatt <- read_csv(str_c(path_raw_data, "/plantatt.csv"),
          lifeform = LF1, perennation1 = P1, perennation2 = P2, woodiness = W, height = Hght,
          ellenberg_light = L)
 
-# Define growth form
-plantatt$growthform <- factor(NA, levels = c("annual", "biennial", "perennial",
-                                             "shrub", "tree"))
+# Define habit
+plantatt$habit <- factor(NA, levels = c("annual", "biennial", "perennial",
+                                        "shrub", "tree"))
   
 # Annual if P1 == annual and is.na(P2)
 x <- which(plantatt$perennation1 == "a" & is.na(plantatt$perennation2))
-plantatt$growthform[x] <- "annual"
+plantatt$habit[x] <- "annual"
 
 # Biennial if P1 == biennial & is.na(P2)
 # or P1 == annual & !is.na(P2)l
@@ -184,25 +184,23 @@ x <- which(plantatt$perennation1 == "b" & is.na(plantatt$perennation2) |
              plantatt$perennation1 == "b" & !is.na(plantatt$perennation2) |
              plantatt$perennation1 == "p" & !is.na(plantatt$perennation2))
 
-plantatt$growthform[x] <- "biennial"
+plantatt$habit[x] <- "biennial"
 
 # Perennial if P1 == perennial and is.na(P2)
 x <- which(plantatt$perennation1 == "p" & is.na(plantatt$perennation2))
-plantatt$growthform[x] <- "perennial"
+plantatt$habit[x] <- "perennial"
 
 # Reclassify as shrub if woody
 x <- which(plantatt$woodiness == "w")
-plantatt$growthform[x] <- "shrub"
+plantatt$habit[x] <- "shrub"
 
 # Reclassify as tree if height > 4 m
-message("need to determine height cutoff for trees, or alternative classification scheme")
 shrub_cutoff <- 4
 x <- which(plantatt$woodiness == "w" & plantatt$height >= (shrub_cutoff * 1e2))
-plantatt$growthform[x] <- "tree"
+plantatt$habit[x] <- "tree"
 
-#summary(plantatt$growthform)
 rm(x)
-plantatt %<>% select(species, lifeform, ellenberg_light, growthform)
+plantatt %<>% select(species, lifeform, ellenberg_light, habit)
 
 # stomata$species[!(stomata$species %in% plantatt$species |
[…]

### 2017-10-06 r/02-make-phylogeny.R

@@ -11,7 +11,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"),
                       photo = col_character(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 ##### Import phylogeny of British Flora -----

### 2017-10-06 r/03-plot-lifeform.R

@@ -11,7 +11,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"),
                       photo = col_character(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 ##### Plot lifeform versus stomatal ratio -----

### 2017-10-06 r/04-filter-data.R

@@ -11,7 +11,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata.csv"),
                       photo = col_character(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 phy <- read.nexus(file = str_c(path_proc_data, "/Lim_etal_2014_final.nex"))
 

### 2017-10-06 r/05-plot-overlap.R

@@ -9,10 +9,10 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
-gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+hf <- c("tree", "shrub", "perennial", "biennial", "annual")
 
 lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
 names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
@@ -20,13 +20,11 @@ lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
 palette(brewer.pal(5, "Greens"))
 
-palette(brewer.pal(5, "Greens"))
-
 pdf(str_c(path_figures, "/figureS_compare-growth-forms.pdf"), 
     width = 4, height = 6, useDingbats = FALSE)
 
   par(mar = c(5, 5, 1, 1))
-  tab <- prop_table(stomata$lifeform, stomata$growthform)
+  tab <- prop_table(stomata$lifeform, stomata$habit)
   tab <- tab[, names(lf)]
   tab %<>% set_colnames(lf)
   dotchart(tab, bg = 1:5, pt.cex = 1.5, xlab = "Proportion", cex.lab = 1.5)

### 2017-10-06 r/06-modify-phylogeny.R

@@ -9,7 +9,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 angio_phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy.nex"))

### 2017-10-06 r/07-plot-phylogeny.R

@@ -10,20 +10,20 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 # Arrange stomata in same order as phylogeny for plotting
 stomata %<>% magrittr::extract(match(phy$tip.label, .$species), 1:ncol(.))
 
-gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+hf <- c("tree", "shrub", "perennial", "biennial", "annual")
 
 lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
 names(lf) <- c("Ch", "Gn", "hc", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
-# Factor growthform and lifeform for plotting
-stomata$growthform %<>% factor(levels = gf)
+# Factor habit and lifeform for plotting
+stomata$habit %<>% factor(levels = hf)
 stomata$lifeform %<>% factor(levels = names(lf))
 
 # Rescale phylogeny for plotting
@@ -35,10 +35,10 @@ theta_mid %<>% magrittr::extract(1:Ntip(phy))
 theta_left <- theta_mid - pi / Ntip(phy)
 theta_right <- theta_mid + pi / Ntip(phy)
 
-##### Using growth form -----
+##### Using habit -----
 
 # Main plot
-pdf(str_c(path_figures, "/figure_phylo-growthform.pdf"), w = 6.5, h = 6.5)#, 
+pdf(str_c(path_figures, "/figure_phylo-habit.pdf"), w = 6.5, h = 6.5)#, 
 # useDingbats = FALSE)
 par(mai = c(1, 0, 1, 2))
 gp <- plot(phy, type = "fan", show.tip.label = FALSE,
@@ -53,7 +53,7 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
     draw_wedge(theta_left[i], theta_right[i], r1 = 1.025, r2 = 1.075, 
-               col = stomata$growthform[i])
+               col = stomata$habit[i])
   }  
 
[…]

### 2017-10-06 r/08-analyse-light-vs-form.R

@@ -10,44 +10,47 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
-##### Growth form versus Ellenberg light indicator values -----
+##### Habit versus Ellenberg light indicator values -----
 
 # phylolm shows that there is no phylogenetic signal in this regression (very high alpha)
 stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
-fitGFvEL_pl <- phylolm(ellenberg_light ~ -1 + growthform, model = "OUrandomRoot", 
+fitHFvEL_pl <- phylolm(ellenberg_light ~ -1 + habit, model = "OUrandomRoot", 
                        data = stomata, phy = phy, upper.bound = 1e10)
-summary(fitGFvEL_pl)
+summary(fitHFvEL_pl)
 
 # therefore using standard ANOVA
-fitGFvEL <- lm(ellenberg_light ~ growthform, data = stomata)
-aovGFvEL <- anova(fitGFvEL)
+fitHFvEL <- lm(ellenberg_light ~ habit, data = stomata)
+aovHFvEL <- anova(fitHFvEL)
 
 # for plotting confidence intervals
-fitGFvEL <- lm(ellenberg_light ~ -1 + growthform, data = stomata)
+fitHFvEL <- lm(ellenberg_light ~ -1 + habit, data = stomata)
 
-gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+hf <- c("tree", "shrub", "perennial", "biennial", "annual")
 
-pdf(str_c(path_figures, "/figure_gf-light.pdf"), 4, 7, useDingbats = FALSE)
+pdf(str_c(path_figures, "/figure_hf-light.pdf"), 4, 7, useDingbats = FALSE)
 par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
[…]

### 2017-10-06 r/09-analyse-stomatal-ratio.R

@@ -10,7 +10,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 ##### Raunikaer life form and Ellenberg light indicator values versus sr_even -----
@@ -125,33 +125,33 @@ for (i in 1:nrow(lf)) {
   
 }
 
-##### Growth form and Ellenberg light indicator values versus sr_even -----
+##### Habit and Ellenberg light indicator values versus sr_even -----
 
-gf <- c("tree", "shrub", "perennial", "biennial", "annual")
+hf <- c("tree", "shrub", "perennial", "biennial", "annual")
 
 # Phylogenetic regression model comparison
 stomata %<>% as.data.frame() %>% set_rownames(.$species) # need to change row names for phylolm
-fitSR_gf_f <- phylolm(sr_even ~ growthform * ellenberg_light, model = "OUrandomRoot",
+fitSR_hf_f <- phylolm(sr_even ~ habit * ellenberg_light, model = "OUrandomRoot",
                       data = stomata, phy = phy, upper.bound = 10)
-fitSR_gf_g <- phylolm(sr_even ~ growthform + ellenberg_light, model = "OUrandomRoot",
+fitSR_hf_g <- phylolm(sr_even ~ habit + ellenberg_light, model = "OUrandomRoot",
                       data = stomata, phy = phy, upper.bound = 10)
-fitSR_gf_h <- phylolm(sr_even ~ growthform, model = "OUrandomRoot",
+fitSR_hf_h <- phylolm(sr_even ~ habit, model = "OUrandomRoot",
                       data = stomata, phy = phy, upper.bound = 10)
 aicSR <- sapply(list(fitSR_lf_a, fitSR_lf_b, fitSRc, fitSR_lf_d, fitSRe,
-                     fitSR_gf_f, fitSR_gf_g, fitSR_gf_h), AIC)
+                     fitSR_hf_f, fitSR_hf_g, fitSR_hf_h), AIC)
 
 # Parametric bootstrap CIs of full model for p-values
[…]

### 2017-10-06 r/10-analyse-stomatal-density.R

@@ -10,7 +10,7 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
                       sr_even = col_double(),
                       lifeform = col_character(),
                       ellenberg_light = col_integer(),
-                      growthform = col_character()
+                      habit = col_character()
                     ))
 
 ##### Stomatal density and light -----

### 2017-10-06 r/11-analyse-phyloSEM.R

@@ -1,7 +1,17 @@
 source("R/header.R")
 
 phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
-stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
+                    col_types = cols(
+                      species = col_character(),
+                      ab_density = col_double(),
+                      ad_density = col_double(),
+                      sr_propAd = col_double(),
+                      sr_even = col_double(),
+                      lifeform = col_character(),
+                      ellenberg_light = col_integer(),
+                      habit = col_character()
+                    ))
 
 ##### Phylogenetic structural equation model ---
   

### 2017-10-14 r/10-analyse-variance.R
Added analysis of phylogenetic variance

@@ -1,0 +1,46 @@
+source("R/header.R")
+
+phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
+stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
+                    col_types = cols(
+                      species = col_character(),
+                      ab_density = col_double(),
+                      ad_density = col_double(),
+                      sr_propAd = col_double(),
+                      sr_even = col_double(),
+                      lifeform = col_character(),
+                      ellenberg_light = col_integer(),
+                      habit = col_character()
+                    ))
+
+# Import fitted model
+fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
+
+# Prepare phylogeny and data for OUwie
+phy$node.label <- rep("1", Nnode(phy))
+OUwie_dat <- data.frame(species = phy$tip.label, regime = "1", x = 0)
+
+# Parameteric bootstrap parameters
+par <- fitSR_lf_pl$bootstrap[, c("sigma2", "alpha")]
+
+# Simulate data (takes several minutes)
+# set.seed(493151857)
+# actual_tree <- apply(par, 1, function(x) {
+#   tmp <- OUwie.sim(phy, data = OUwie_dat, alpha = rep(x["alpha"], 2), 
+#                    sigma.sq = rep(x["sigma2"], 2), theta = c(0, 0), theta0 = 0)
+#   var(tmp$X)
+# })
+# 
+# stationary_tree <- apply(par, 1, function(x) {
+#   var(rnorm(Ntip(phy), 0, sqrt(x["sigma2"] / (2 * x["alpha"]))))
+# })
+# 
+# var_comp <- data.frame(actual_tree, stationary_tree)
+# write_rds(var_comp, str_c(path_objects, "/var_comp.rds"))
+
+var_comp <- read_rds(str_c(path_objects, "/var_comp.rds"))
+
+ttest <- t.test(var_comp$actual_tree, var_comp$stationary_tree, paired = TRUE)
+
+# Export objects to ms
+export2ms(c("ttest", "var_comp"))

### 2017-10-14 r/XX-OUwie.R

@@ -1,35 +1,0 @@
-source("R/header.R")
-
-phy <- read.nexus(file = str_c(path_proc_data, "/angio_phy_modified.nex"))
-stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
-                    col_types = cols(
-                      species = col_character(),
-                      ab_density = col_double(),
-                      ad_density = col_double(),
-                      sr_propAd = col_double(),
-                      sr_even = col_double(),
-                      lifeform = col_character(),
-                      ellenberg_light = col_integer(),
-                      habit = col_character()
-                    ))
-
-library(OUwie)
-
-# Import fitted model
-fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
-
-# Prepare phylogeny and data for OUwie
-phy$node.label <- rep("1", Nnode(phy))
-OUwie_dat <- data.frame(species = phy$tip.label, regime = "1", X = 0)
-
-#OUwie.sim(phy, alpha = fitSR_lf_a$optpar, sigma.sq = fitSR_lf_a$sigma2,
-#          theta = c(0, 0), theta0 = 0)
-
-# Sample root state randomly from stationary distribution
-nsim <- 1e1
-rnorm(nsim, 0, sqrt(fitSR_lf_pl$sigma2 / (2 * fitSR_lf_pl$optpar)))
-sim_data <- OUwie.sim(phy, data = OUwie_dat, alpha = rep(fitSR_lf_pl$optpar, 2), 
-                      sigma.sq = rep(fitSR_lf_pl$sigma2, 2),
-                      theta = c(0, 0), theta0 = 0)
-sd(sim_data$X)
-sd(resid(fitSR_lf_pl))

### 2017-10-14 r/functions.R

@@ -142,15 +142,27 @@ export2ms <- function(x, path_export = getOption("path_export", "/export")) {
 }
 
 # Import objects to ms
+# import2ms <- function(path_export = getOption("path_export", "/export")) {
+#   
+#   path <- normalizePath(str_c(path_ms, path_export))
+#   files <- list.files(path)
+#   eval(parse(text = str_c(files, " <- read_rds('", path, "/", files, "')")),
+#        envir = .GlobalEnv)
+# 
+# }
+
+# Import objects to ms
 import2ms <- function(path_export = getOption("path_export", "/export")) {
   
   path <- normalizePath(str_c(path_ms, path_export))
-  files <- list.files(path)
-  eval(parse(text = str_c(files, " <- read_rds('", path, "/", files, "')")),
+  objects <- list.files(path)
+  files <- str_c(path, "/", objects) %>% 
+    normalizePath() %>% 
+    str_replace_all("\\\\", "\\\\\\\\")
+  eval(parse(text = str_c(objects, " <- read_rds('", files, "')")),
        envir = .GlobalEnv)
-
+  
 }
-
 # Calculate sr_even
 calc_sr_even <- function(ab_density, ad_density) {
   apply(cbind(ab_density, ad_density), 1, min) / 

### 2017-10-14 r/header.R

@@ -2,8 +2,8 @@ rm(list = ls())
 graphics.off()
 
 # remove before final publication
-# file.copy("~/Google Drive/CommonResources/refs.bib",
-#           "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+file.copy("~/Google Drive/CommonResources/refs.bib",
+          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {
@@ -15,6 +15,7 @@ library(ape)
 library(dplyr)
 library(lavaan)
 library(magrittr)
+library(OUwie)
 library(phylolm)
 library(phytools)
 library(plyr)

### 2017-10-14 r/run-all.R

@@ -14,5 +14,6 @@ source("r/07-plot-phylogeny.R")
 # Analyze data
 source("r/08-analyse-light-vs-form.R")
 source("r/09-analyse-stomatal-ratio.R")
-source("r/10-analyse-stomatal-density.R")
-source("r/11-analyse-phyloSEM.R")
+source("r/10-analyse-variance.R")
+source("r/11-analyse-stomatal-density.R")
+source("r/12-analyse-phyloSEM.R")

### 2017-10-14 ms/ms.Rnw
Added analysis of phylogenetic variance

@@
 I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
+
+I also tested whether phylogenetic could explain the residual variation in stomatal ratio after accounting for growth form and \el.
+Specifically, I compared the expected residual variation given the actual tree versus a hypothetical tree where trait evolution has reached stationarity (i.e.
+a star phylogeny with infinite branch lengths).
+If phylogeny explains much of the variation, then the simulated residual variance from the actual tree should be greater than that of the stationary tree.
+I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r lifeform, \el, and their interaction; see Results).
+I performed the first set of simulations using the actual phylogenetic tree in \pkg{OUwie} version 1.50 \citep{Beaulieu_OMeara_2016}.
+Each simulation used a different bootstrap parameter sample of $\alpha$ and $\sigma ^ 2$, where $\alpha$ is the return rate to the mean and $\sigma ^ 2$ is the diffusion rate.
+At stationarity, the variance of an Ornstein-Uhlenbeck trait is equal to $\sigma ^ 2 / 2 \alpha$.
+Therefore, I simulated stationary data by assuming a normal distribution with this variance estimated from the bootstrap samples.
+For comparability, I set the mean of simulations from both actual phylogeny and the stationary to zero.
+I compared the actual to stationary variance across simulated datasets using a paired $t$-test.
+
 % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models).
@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-% Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
+
+Although there was significant phylogenetic signal in the residual variation of stomatal ratio (see R code), the total variation among these species was consistent with a trait at stationarity.
+Specifically, the simulated residual trait variation, after accounting for Raunki\ae r life form and \el, from the actual tree was not significantly greater than that simulated from a tree where traits had reached stationarity (paired $t$-test, $P = \Sexpr{pval2latex(ttest$p.value)}$).
+Hence, phylogenetic nonindependence is an important statistical consideration, but phylogeny does not explain stomatal trait variation among British angiosperms.
 
@@
 
-% Revise and put after 2nd sentence: Although I repeated certain analyses from previous papers, this is an important advance because previous analyses did not include phylogeny or did so improperly...
-Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy.
-In British angiosperms, both factors are important, though light may be a stronger determinant than growth form based on information criterion comparisons (Table~\ref{table:srAIC}).
-Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
+Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing plants, but not shrubs and trees.
+This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindenpendence.
+For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle these effects.
+Later work by \cite{Peat_Fitter_1994b} demonstrated these trends again using family-level comparisons, a basic method to account for phylogenetic nonindependence \citep[see also][]{Mott_etal_1982, Beerling_Kelly_1996}.
+However, this approach is still problematic because traits like growth from can be highly phylogenetically conserved.
+For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonidendence.
+By combining trait, ecological, and phylogenetic datasets on British angiosperms, we now know that not only do both light and growth form influence stomatal ratio, but in fact their effects may reinforce one another.
+Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role o flight \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.
+
+The interaction between light and growth form among British angiosperms suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs).
 This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}.
-The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms).
+Along these lines, Raunki\ae r lifeform may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics.
+For example, on an axis of `fast' to `slow' life history, geophytes more closely resemble phanerophytes than do chamaephytes or hemicryptophytes \citep{Salguero-Gomez_etal_2016}.
+Similarly, the relationship between light and stomatal ratio for geophytes was intermediate between that for phanerophytes and chamaephytes/hemicryptophytes (Fig.~\ref{fig:hf-light}).
+These comparisons indirectly suggest that both high light and fast life history are necessary to induce strong selection for amphistomy.
+The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both therophyte/annual and perennial forms).
 I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light.
@@
 For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested.
-Likewise, phanerophytes (most tall trees) are almost always hypostomatous (see also \cite{Muir_2015}).
+Likewise, phanerophytes (mostly tall trees) are almost always hypostomatous \citep[see also][]{Muir_2015}.
 Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling.

### 2017-10-14 ms/ms.tex
Added analysis of phylogenetic variance

@@
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
-% Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
+
+Although there was significant phylogenetic signal in the residual variation of stomatal ratio (see R code), the total variation among these species was consistent with a trait at stationarity.
+Specifically, the simulated residual trait variation, after accounting for Raunki\ae r life form and \el, from the actual tree was not significantly greater than that simulated from a tree where traits had reached stationarity (paired $t$-test, $P = 0.331$).
+Hence, phylogenetic nonindependence is an important statistical consideration, but phylogeny does not explain stomatal trait variation among British angiosperms.
 
@@
 Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}).
-Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively).
-However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1).
-Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density.
-The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$).
+Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.9 $\times10^{-7}$, respectively).
+However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12).
+Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.27-fold change in adaxial stomatal density versus a 1.13-fold change in abaxial stomatal density.
+The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $2.5 $\times10^{-10}$).
 These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.
@@
   & 12
-  & -33.3
-  & 90.6
+  & -33.2
+  & 90.4
   & 0 \\
@@
   & -38.2
-  & 100.3
-  & 9.8 \\
+  & 100.4
+  & 9.9 \\
     \el~+ Raunki\ae r lifeform
-  & 0.47
-  & 0.072
+  & 0.46
+  & 0.071
   & 0.32
   & 8
-  & -40.3
-  & 96.5
+  & -40.2
+  & 96.4
   & 6 \\
@@
   & 8
-  & -43.3
+  & -43.4
   & 102.7
-  & 12.1 \\
+  & 12.3 \\
     Raunki\ae r lifeform
@@
   & 172.4
-  & 81.8 \\
+  & 82 \\
     growth form
@@
   & -82.5
-  & 178.9
-  & 88.4 \\
+  & 179.1
+  & 88.6 \\
     \el
   & 0.64
-  & 0.108
+  & 0.107
   & 0.26
@@
   & 126.6
-  & 36.1 \\
+  & 36.2 \\
     null
@@
   & 3
-  & -107.5
-  & 221
-  & 130.5 \\
+  & -107.6
+  & 221.1
+  & 130.7 \\
 	\bottomrule

### 2017-10-14 r/07-plot-phylogeny.R

@@ -38,8 +38,8 @@ theta_right <- theta_mid + pi / Ntip(phy)
 ##### Using habit -----
 
 # Main plot
-pdf(str_c(path_figures, "/figure_phylo-habit.pdf"), w = 6.5, h = 6.5)#, 
-# useDingbats = FALSE)
+pdf(str_c(path_figures, "/figure_phylo-habit.pdf"), w = 6.5, h = 6.5, 
+    useDingbats = FALSE)
 par(mai = c(1, 0, 1, 2))
 gp <- plot(phy, type = "fan", show.tip.label = FALSE,
            x.lim = c(-1.25, 1.25), y.lim = c(-1.25, 1.25))
@@ -118,8 +118,8 @@ dev.off()
 ##### Using Raunkiaer lifeform -----
 
 # Main plot
-pdf(str_c(path_figures, "/figure_phylo-lifeform.pdf"), w = 6.5, h = 6.5)#, 
-   # useDingbats = FALSE)
+pdf(str_c(path_figures, "/figure_phylo-lifeform.pdf"), w = 6.5, h = 6.5, 
+    useDingbats = FALSE)
 par(mai = c(1, 0, 1, 2))
 gp <- plot(phy, type = "fan", show.tip.label = FALSE,
            x.lim = c(-1.25, 1.25), y.lim = c(-1.25, 1.25))

### 2017-10-14 r/install-packages.R

@@ -2,6 +2,7 @@ install.packages("ape")
 install.packages("dplyr")
 install.packages("lavaan")
 install.packages("magrittr")
+install.packages("OUwie")
 install.packages("phylolm")
 install.packages("phytools")
 install.packages("plyr")

### 2017-10-14 r/01-process-data.R
fixing some bs

@@ -98,7 +98,7 @@ x <- tmp %>%
 tmp$ad_density.x[x] <- tmp$ad_density.y[x]
 
 tmp %<>% 
-  select(species, acceptedname, ab_density.x, ad_density.x) %>%
+  dplyr::select(species, acceptedname, ab_density.x, ad_density.x) %>%
   set_colnames(c("species", "acceptedname", "ab_density", "ad_density"))
 
 stomata %<>% 
@@ -162,9 +162,9 @@ plantatt <- read_csv(str_c(path_raw_data, "/plantatt.csv"),
                        N = col_integer(),
                        S = col_integer()
                      )) %>%
-  select(species = `Taxon name`, 
-         lifeform = LF1, perennation1 = P1, perennation2 = P2, woodiness = W, height = Hght,
-         ellenberg_light = L)
+  dplyr::select(species = `Taxon name`, 
+                lifeform = LF1, perennation1 = P1, perennation2 = P2, woodiness = W, 
+                height = Hght, ellenberg_light = L)
 
 # Define habit
 plantatt$habit <- factor(NA, levels = c("annual", "biennial", "perennial",
@@ -200,7 +200,7 @@ x <- which(plantatt$woodiness == "w" & plantatt$height >= (shrub_cutoff * 1e2))
 plantatt$habit[x] <- "tree"
 
 rm(x)
-plantatt %<>% select(species, lifeform, ellenberg_light, habit)
+plantatt %<>% dplyr::select(species, lifeform, ellenberg_light, habit)
 
 # stomata$species[!(stomata$species %in% plantatt$species |
 #                    stomata$acceptedname %in% plantatt$species)] # Missing data for several

### 2017-10-14 r/04-filter-data.R

@@ -23,7 +23,7 @@ nCAM <- length(which(stomata$photo == "cam"))
 
 stomata %<>% filter(lifeform %in% c("Ph", "Ch", "hc", "Gn", "Th"),
                     photo == "c3") %>%
-  select(-photo)
+  dplyr::select(-photo)
 
 angio_phy <- extract.clade(phy, node = length(phy$tip.label) + 
                              which(phy$node.label == "angiosperms"))
@@ -31,7 +31,7 @@ angio_phy <- extract.clade(phy, node = length(phy$tip.label) +
 # Change species to acceptedname is latter is in phylogeny
 stomata$species[which(!stomata$species %in% phy$tip.label)] <-
   stomata$acceptedname[which(!stomata$species %in% phy$tip.label)]
-stomata %<>% select(-acceptedname)
+stomata %<>% dplyr::select(-acceptedname)
 
 n <- nrow(stomata)
 

### 2017-10-14 r/09-analyse-stomatal-ratio.R

@@ -110,7 +110,7 @@ dev.off()
 
 lf <- data.frame(lifeform = names(lf))
 lf %<>% 
-  expand(lifeform, lifeform) %>%
+  tidyr::expand(lifeform, lifeform) %>%
   remove_diag()
 lf %<>% mutate(mean = numeric(nrow(.)),
                median = numeric(nrow(.)),

### 2017-10-14 r/12-analyse-phyloSEM.R

@@ -16,9 +16,9 @@ stomata <- read_csv(str_c(path_proc_data, "/stomata_filtered.csv"),
 ##### Phylogenetic structural equation model ---
   
 # small data.frame for Rphylopars
-td <- stomata %>% select(species, ab_density, ad_density, ellenberg_light) %>%
+td <- stomata %>% dplyr::select(species, ab_density, ad_density, ellenberg_light) %>%
   mutate(logSDab = log(ab_density + 1), logSDad = log(ad_density + 1)) %>%
-  select(-ab_density, -ad_density) %>%
+  dplyr::select(-ab_density, -ad_density) %>%
   as.data.frame() %>% 
   set_rownames(.$species) # need to change row names for phylolm
 

### 2017-10-17 ms/ms.Rnw
updated phylo figure

@@
 Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction.
+Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity.
+The variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable at stationarity under an Ornstein-Uhlenbeck process (see Results).
+This may not be the case for younger clades that have radiated in the past few million years.
 

### 2017-10-17 r/07-plot-phylogeny.R

@@ -46,9 +46,10 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
 
 # Growth form
 
-  ## Palette
+  ## Palette for growth form
   palette(brewer.pal(5, "Greens"))
-
+  palette(rev(palette()))
+  
   ## Plot wedges
   message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
@@ -59,50 +60,79 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   ## Legend
   clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
        grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
-  text(grconvertX(5.5, "in", "user"), grconvertY(0.85, "ndc", "user"),
+  text(grconvertX(5.5, "in", "user"), grconvertY(0.9, "ndc", "user"),
        labels = "Growth form", cex = 1.5)
   points(rep(grconvertX(5, "in", "user"), 5),
-         seq(0.1, 0.9, 0.2) * 1.25,
+         seq(0.15, 0.95, 0.2) * grconvertY(0.85, "ndc", "user"),
          pch = 21, col = "black", bg = 1:5, cex = 3)
   text(rep(grconvertX(5, "in", "user"), 5),
-       seq(0.1, 0.9, 0.2) * 1.25,
+       seq(0.15, 0.95, 0.2) * grconvertY(0.85, "ndc", "user"),
        labels = hf, pos = 4, offset = 1)
-
+  
+# Light
+  
+  ## Palette for light
+  palette(brewer.pal(length(unique(stomata$ellenberg_light)), "Oranges"))
+  palette(rev(palette()))
+  
+  ## Plot wedges
+  message("double check that wedges are drawn in same order as data")
+  for (i in 1:Ntip(phy)) {
+    draw_wedge(theta_left[i], theta_right[i], r1 = 1.075, r2 = 1.125, 
+               col = stomata$ellenberg_light[i] - min(stomata$ellenberg_light) + 1)
+  }  
+  
+  ## Legend
+  clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
+       grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
[…]

### 2017-10-21 ms/ms.Rnw
updated figure captions

@@
 High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees).
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees).
 Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
-	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
+	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution.
+First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
+Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange.
+These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
 \end{itemize}
@@
 The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs.
-\citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora.
+\cite{Salisbury_1927} first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora.
 These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants.
@@
 Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion.
-Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}.
+\cite{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}.
 Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.
@@
 Stomatal ratio is also associated with growth form.
-In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous.
+In the British flora, \cite{Salisbury_1927} found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous.
 This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}.
@@
 Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another.
-Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants.
+Open habitat and growth form may be confounded because open habitats generally consist of more short-statured, herbaceous plants.
 Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habitats \citep{Salisbury_1927}.
@@
 
+% MARKER (still revising, adding predictions here): Specifically, I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).
+
 The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume.
@@
 Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.
+Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.
 
@@
 For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}.
-A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.
+A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time a rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.
 
@@
 I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data.
-The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}).
+The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}, \ref{fig:phylo-habit}).
 The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.
@@
 I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms.
-I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).
 I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}.
@@
 For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution.
-I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
+I used 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
 
@@
 \centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
-\caption{CAPTION} 
+\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+At the center is the phylogenetic tree for \Sexpr{nrow(stomata)} species of British angiosperms.
+For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el.
+The outer circle indicates the stomatal ratio for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
 \label{fig:phylo-lifeform}
@@
 \begin{figure}[ht]
+\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
+\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+At the center is the phylogenetic tree for \Sexpr{nrow(stomata)} species of British angiosperms.
+For each species, the green wedges indicate plant habit and the orange wedges indicate \el.
+The outer circle indicates the stomatal ratio for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
+\label{fig:phylo-habit}
+\end{figure}
+
+\begin{figure}[ht]
 \centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
-\caption{CAPTION}
+\caption{Raunki\ae r lifeform and plant habit broadly overlap.
+The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit.
+For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
 \label{fig:compare-growth-forms}

### 2017-10-21 ms/ms.tex
updated figure captions

@@
 High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
-	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms.
+	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for 372 species of British angiosperms.
 I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
-	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees).
+	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees).
 Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
-	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
+	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution.
+First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
+Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange.
+These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
 \end{itemize}
@@
 I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data.
-The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}).
+The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}, \ref{fig:phylo-habit}).
 The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.
@@
 \centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
-\caption{CAPTION} 
+\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+At the center is the phylogenetic tree for 372 species of British angiosperms.
+For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el.
+The outer circle indicates the stomatal ratio for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
 \label{fig:phylo-lifeform}
@@
 \begin{figure}[ht]
+\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
+\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+At the center is the phylogenetic tree for 372 species of British angiosperms.
+For each species, the green wedges indicate plant habit and the orange wedges indicate \el.
+The outer circle indicates the stomatal ratio for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
+\label{fig:phylo-habit}
+\end{figure}
+
+\begin{figure}[ht]
 \centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
-\caption{CAPTION}
+\caption{Raunki\ae r lifeform and plant habit broadly overlap.
+The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit.
+For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
 \label{fig:compare-growth-forms}

### 2017-10-21 ms/ms.Rnw
revised introduction

@@
 
-% MARKER (still revising, adding predictions here): Specifically, I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).
-
-The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume.
+In answering these questions, I both reassessed previous hypotheses using newer phylogenetic comparative methods and evaluated previously untested hypotheses.
+I predicted \textit{a priori} that light habitat and growth would be correlated.
+Species with faster life histories, especially therophytes (annuals), would on average inhabitat sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees).
+Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence.
+Although these predictions have been tested previously, it is critical to reevaluate them here with updated methods because the subsequent untested hypotheses build on these results.
+The first novel hypothesis I tested predicts that light and growth form interact.
+Specifically, I hypothesized that both high light and herbaceous growth would be required to favor a more even stomatal ratio (i.e.
+amphistomy).
+Finally, I tested whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate.
+If high light and growth form favor amphistomy because it increases photosynthesis, then it follows that they should also favor other stomatal traits that reinforce this advantage.
 If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small.
 In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well.
-Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.
+Evolutionary coordination of amphistomy and high stomatal density would thus reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.
 Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

### 2017-10-21 ms/ms.Rnw
resubmitted

@@
 Word count: \\
-Summary: 190\\
-Introduction: 1179\\
-Materials and Methods: 1176\\
-Results: 433\\
-Discussion: 1504\\
-Acknowledgement: 24\\
-3 Figures and 1 Table, 1 Supplemental Figure
+Summary: 199\\
+Introduction: 1355\\
+Materials and Methods: 1698\\
+Results: 582\\
+Discussion: 1935\\
+Acknowledgement: 26\\
+4 Figures and 1 Table, 5 Supplemental Figures
 
@@
 %\pagenumbering{gobble}
+%\raggedright
+
+%\clearpage
 
@@
 \begin{itemize}
+
 	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy).
 High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
+
 	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms.
-I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
+I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio and density.
+
 	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees).
-Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
-	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution.
-First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange.
-These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
+Furthermore, amphistomy and stomatal density evolve together in response to light.
+
+	\item Comparative analyses of British angiosperms reveal two major insights.
+First, light and growth form interact to shape stomatal ratio; amphistomy is common under high light, but mostly for herbs.
+Second, coordinated evolution of adaxial stomatal density and light tolerance indicates that amphistomy helps to optimally balance light acquisition with gas exchange.
+Stomatal ratio may have potential as a functional trait for paleoecology and crop improvement.
+
 \end{itemize}
@@
 
-Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r lifeform, stomata, stomatal ratio
+Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio
 
@@
 I predicted \textit{a priori} that light habitat and growth would be correlated.
-Species with faster life histories, especially therophytes (annuals), would on average inhabitat sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees).
+Species with faster life histories, especially therophytes (annuals), would on average inhabit sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees).
 Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence.
@@
 high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates.
-Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}.
+Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists perform worse than nonphylogenetic tests \citep{Revell_2010}.
 Therefore, I used a regular ANOVA with Type-2 sum of squares.
@@
 
+% REMOVED: I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.
+
 I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
-Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code).
-I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.
-I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC).
+I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC), a common measure of model fit that penalizes additional parameters.
+Phylogenetic linear models simultaneously estimate the effect of continuous and categorical predictors while controlling for phylogenetic nonindependence.
 For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution.
+The Ornstein-Uhlenbeck model is characterized by a diffusion rate ($\sigma ^ 2$) and a return rate ($\alpha$), which describes the phylogenetic signal (see above).
 I used 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.
 
-I also tested whether phylogenetic could explain the residual variation in stomatal ratio after accounting for growth form and \el.
+I tested whether phylogenetic nonstationarity could explain the residual variation in stomatal ratio after accounting for growth form and \el.
 Specifically, I compared the expected residual variation given the actual tree versus a hypothetical tree where trait evolution has reached stationarity (i.e.
@@
 If phylogeny explains much of the variation, then the simulated residual variance from the actual tree should be greater than that of the stationary tree.
-I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r lifeform, \el, and their interaction; see Results).
+I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r life form, \el, and their interaction; see Results).
 I performed the first set of simulations using the actual phylogenetic tree in \pkg{OUwie} version 1.50 \citep{Beaulieu_OMeara_2016}.
@@
 Therefore, I simulated stationary data by assuming a normal distribution with this variance estimated from the bootstrap samples.
-For comparability, I set the mean of simulations from both actual phylogeny and the stationary to zero.
+For comparability, I set the mean of simulations from both actual phylogeny and the stationary `phylogeny' to zero.
 I compared the actual to stationary variance across simulated datasets using a paired $t$-test.
@@
 
-I also tested whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface.
-The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$.
-The second scenario predicts that \el~acts similarly on both and that there is a negative covariance ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$).
-I tested these competing predictions by fitting a very simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}.
+If light-mediated increases in adaxial stomatal density can evolve while abaxial density remains roughly constant, then the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ will be stronger than that for $\mathrm{sd_{ab}}$.
+Under this scenario, stomatal ratio and density evolve in a coordinated fashion in response to light.
+Alternatively, if greater \el~favors greater stomatal ratio but total stomatal density is roughly constant, then there will be a negative covariance between ab- and adaxial density ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$).
+I tested these competing predictions by fitting a simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}.
 In general, SEMs attempt to determine whether variables are related causally or whether a relationship is mediated by another correlated variable.
@@
 estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$).
-To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}.
+I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012} to fit the SEM by finding parameter estimates would lead to phylogenetic covariance close to that observed in the data.
 I tested whether parameter estimates were significantly different from zero using $z$-scores.
@@
 Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
-When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}).
-Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses.
+When classified based on plant habit, growth form interacted with \el~less strongly ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}).
+Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore I focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
 The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups.
-There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s.
+There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \els.
 In contrast, phanerophytes were nearly always hypostomatous regardless of \el.
@@
 
-The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear.
+The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance of this variation is not well understood.
 Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work.
@@
 As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Fig.~\ref{fig:lf-light}, Fig.~\ref{fig:hf-light}).
-Nevertheless, both \el~and growth form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms.
+Nevertheless, both \el~and growth form affect stomatal ratio, but these factors also interact.
+This new finding shows that the influence of \el~on stomatal ratio varies across forms.
 Finally, I show for the first time that adaxial stomatal density in particular accounts for most of the coordinated evolution between light tolerance and stomatal density.
@@
 
-Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing plants, but not shrubs and trees.
-This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindenpendence.
-For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle these effects.
+Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing herbs, but not shrubs and trees.
+This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindependence.
+For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle light and growth form.
 Later work by \cite{Peat_Fitter_1994b} demonstrated these trends again using family-level comparisons, a basic method to account for phylogenetic nonindependence \citep[see also][]{Mott_etal_1982, Beerling_Kelly_1996}.
 However, this approach is still problematic because traits like growth from can be highly phylogenetically conserved.
-For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonidendence.
+For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonindependence.
 By combining trait, ecological, and phylogenetic datasets on British angiosperms, we now know that not only do both light and growth form influence stomatal ratio, but in fact their effects may reinforce one another.
-Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role o flight \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.
+Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role of light \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.
 
@@
 This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}.
-Along these lines, Raunki\ae r lifeform may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics.
+Along these lines, Raunki\ae r life form may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics.
 For example, on an axis of `fast' to `slow' life history, geophytes more closely resemble phanerophytes than do chamaephytes or hemicryptophytes \citep{Salguero-Gomez_etal_2016}.
@@
 The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}.
-It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface.
+It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface.
 Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous.
@@
 Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}.
-The cost of adaxial stomata may be greater in the shade because greater leaf wetness and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.
+The cost of adaxial stomata may be greater in the shade because wetter leaves and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.
 
@@
 For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested.
-Likewise, phanerophytes (mostly tall trees) are almost always hypostomatous \citep[see also][]{Muir_2015}.
+Likewise, phanerophytes are almost always hypostomatous \citep[see also][]{Muir_2015}.
 Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling.
@@
 Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity.
-The variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable at stationarity under an Ornstein-Uhlenbeck process (see Results).
+The observed variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable from that expected for a trait at stationarity under an Ornstein-Uhlenbeck process (see Results).
 This may not be the case for younger clades that have radiated in the past few million years.
 
-\subsection*{Why does adaxial stomatal density control stomatal ratio?}
+\subsection*{Coordinated evolution of stomatal ratio and density in response to light}
 
@@
 Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density.
-This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.
+To my knowledge, this highly nonrandom pattern among British angiosperms has not been demonstrated before, but it parallels evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.
 
@@
 A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}.
+These results suggest that coordination between stomatal ratio and density might play a greater role than previously appreciated in optimizing CO$_2$ supply and demand under different light regimes \citep[see also][]{Beerling_Kelly_1996}.
 
@@
 \section*{Acknowledgements}
-I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript. I was supported by an NSERC CREATE grant.
+I thank Sally Otto, Matt Pennell, Rob Salguero-G\'{o}mez, and two anonymous reviewers for feedback on this manuscript. I was supported by an NSERC CREATE grant.
 
@@
  \setstretch{\stretchy}
-  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$).
+  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r life form predict stomatal ratio ($\mathrm{SR_{even}}$).
 I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$.
@@
   \midrule  
-    \el~$\times$ Raunki\ae r lifeform
+    \el~$\times$ Raunki\ae r life form
   & \Sexpr{round(fitSR_lf_a$optpar, 2)}
@@
   & \Sexpr{round(fitSR_hf_f$aic - min(aicSR), 1)} \\
-    \el~+ Raunki\ae r lifeform
+    \el~+ Raunki\ae r life form
   & \Sexpr{round(fitSR_lf_b$optpar, 2)}
@@
   & \Sexpr{round(fitSR_hf_g$aic - min(aicSR), 1)} \\
-    Raunki\ae r lifeform
+    Raunki\ae r life form
   & \Sexpr{round(fitSR_lf_d$optpar, 2)}
@@
 For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el.
-The outer circle indicates the stomatal ratio for each species.
-As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
+The outer circle indicates the stomatal ratio ($\mathrm{SR_{even}}$) for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ratio means that most stomata are on the lower surface.}
 \label{fig:phylo-lifeform}
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}
-\caption{Lifeforms have different tolerances for sun and shade among British angiosperms.
+\caption{Life forms have different tolerances for sun and shade among British angiosperms.
 Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms.
-Height of the bars indicate the raw proportion of species in each bin for that lifeform.
-The sample size for each lifeform is listed next in parentheses.
-The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.}
+Height of the bars indicate the raw proportion of species in each bin for that life form.
+The sample size for each life form is listed next in parentheses.
+The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each life form based on phylogenetic regression are above the histogram.}
 \label{fig:lf-light}
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
-%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
+%\centerline{\includegraphics[angle=270]{figures/figure_SR-lf.eps}} % Needed this to upload file to journal website
 \caption{The effect of light on stomatal ratio depends on Raunki\ae r life form.
@@
 \caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$).
-The violin plot shows stomatal ratio as a function of Raunki\ae r lifeform.
+The violin plot shows stomatal ratio as a function of Raunki\ae r life form.
 The width of the grey polygons indicates the density of data.
 Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles.
-Sample sizes per lifeform in the dataset are given above the upper plot margin.
+Sample sizes per life form in the dataset are given above the upper plot margin.
 $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
 \label{fig:violin}
+\end{figure}
+
+\begin{figure}[ht]
+\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
+\caption{Raunki\ae r life form and plant habit broadly overlap.
+The dot chart shows for each Raunki\ae r life form, the proportion that overlap with a given plant habit.
+For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
+\label{fig:compare-growth-forms}
 \end{figure}
@@
 For each species, the green wedges indicate plant habit and the orange wedges indicate \el.
-The outer circle indicates the stomatal ratio for each species.
-As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.}
+The outer circle indicates the stomatal ratio ($\mathrm{SR_{even}}$) for each species.
+As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ratio means that most stomata are on the lower surface.}
 \label{fig:phylo-habit}
-\end{figure}
-
-\begin{figure}[ht]
-\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
-\caption{Raunki\ae r lifeform and plant habit broadly overlap.
-The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit.
-For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
-\label{fig:compare-growth-forms}
 \end{figure}
@@
 \caption{Growth forms have different tolerances for sun and shade among British angiosperms.
-Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different growth forms.
-Height of the bars indicate the raw proportion of species in each bin for that lifeform.
-The sample size for each lifeform is listed next in parentheses.
-The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.}
+Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different plant habits.
+Height of the bars indicate the raw proportion of species in each bin for that habit.
+The sample size for each habit is listed next in parentheses.
+The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each habit based on phylogenetic regression are above the histogram.}
 \label{fig:hf-light}
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
-%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
+%\centerline{\includegraphics[angle=270]{figures/figure_SR-hf.eps}} % Needed this to upload file to journal website
 \caption{The effect of light on stomatal ratio depends on growth form.

### 2017-10-21 ms/ms.tex
resubmitted

@@
 \begin{itemize}
+
 	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy).
 High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
+
 	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for 372 species of British angiosperms.
-I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
+I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio and density.
+
 	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees).
-Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
-	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution.
-First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs.
-Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange.
-These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
+Furthermore, amphistomy and stomatal density evolve together in response to light.
+
+	\item Comparative analyses of British angiosperms reveal two major insights.
+First, light and growth form interact to shape stomatal ratio; amphistomy is common under high light, but mostly for herbs.
+Second, coordinated evolution of adaxial stomatal density and light tolerance indicates that amphistomy helps to optimally balance light acquisition with gas exchange.
+Stomatal ratio may have potential as a functional trait for paleoecology and crop improvement.
+
 \end{itemize}
@@
 
-% MARKER (still revising, adding predictions here): Specifically, I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).
-
-The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume.
+In answering these questions, I both reassessed previous hypotheses using newer phylogenetic comparative methods and evaluated previously untested hypotheses.
+I predicted \textit{a priori} that light habitat and growth would be correlated.
+Species with faster life histories, especially therophytes (annuals), would on average inhabit sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees).
+Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence.
+Although these predictions have been tested previously, it is critical to reevaluate them here with updated methods because the subsequent untested hypotheses build on these results.
+The first novel hypothesis I tested predicts that light and growth form interact.
+Specifically, I hypothesized that both high light and herbaceous growth would be required to favor a more even stomatal ratio (i.e.
+amphistomy).
+Finally, I tested whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate.
+If high light and growth form favor amphistomy because it increases photosynthesis, then it follows that they should also favor other stomatal traits that reinforce this advantage.
 If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small.
 In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well.
-Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.
+Evolutionary coordination of amphistomy and high stomatal density would thus reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.
 Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.
@@
 Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}).
-When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}).
-Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses.
+When classified based on plant habit, growth form interacted with \el~less strongly ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}).
+Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore I focus hereafter on those analyses.
 Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}).
@@
   \midrule  
-    \el~$\times$ Raunki\ae r lifeform
+    \el~$\times$ Raunki\ae r life form
   & 0.46
@@
   & 9.8 \\
-    \el~+ Raunki\ae r lifeform
+    \el~+ Raunki\ae r life form
   & 0.47
@@
   & 12.1 \\
-    Raunki\ae r lifeform
+    Raunki\ae r life form
   & 0.34

### 2017-10-21 r/07-plot-phylogeny.R

@@ -51,7 +51,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   palette(rev(palette()))
   
   ## Plot wedges
-  message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
     draw_wedge(theta_left[i], theta_right[i], r1 = 1.025, r2 = 1.075, 
                col = stomata$habit[i])
@@ -76,7 +75,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   palette(rev(palette()))
   
   ## Plot wedges
-  message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
     draw_wedge(theta_left[i], theta_right[i], r1 = 1.075, r2 = 1.125, 
                col = stomata$ellenberg_light[i] - min(stomata$ellenberg_light) + 1)
@@ -161,7 +159,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   palette(rev(palette()))
 
   ## Plot wedges
-  message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
     draw_wedge(theta_left[i], theta_right[i], r1 = 1.025, r2 = 1.075, 
                col = stomata$lifeform[i])
@@ -186,7 +183,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   palette(rev(palette()))
   
   ## Plot wedges
-  message("double check that wedges are drawn in same order as data")
   for (i in 1:Ntip(phy)) {
     draw_wedge(theta_left[i], theta_right[i], r1 = 1.075, r2 = 1.125, 
                col = stomata$ellenberg_light[i] - min(stomata$ellenberg_light) + 1)

### 2017-10-21 r/09-analyse-stomatal-ratio.R

@@ -43,7 +43,7 @@ fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
 # Figure
 
 pdf(str_c(path_figures, "/figure_SR-lf.pdf"), 4, 7, useDingbats = FALSE)
-#postscript(str_c(path_figures, "/figure_SRmultReg.ps"), width = 4, height = 7) # need for submitting to journal website
+#postscript(str_c(path_figures, "/figure_SR-lf.eps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {
@@ -151,7 +151,7 @@ fitSR_hf_pl <- read_rds(str_c(path_objects, "/fitSR_hf_pl.rds"))
 # Figure
 
 pdf(str_c(path_figures, "/figure_SR-hf.pdf"), 4, 7, useDingbats = FALSE)
-#postscript(str_c(path_figures, "/figure_SR-hf.ps"), width = 4, height = 7) # need for submitting to journal website
+#postscript(str_c(path_figures, "/figure_SR-hf.eps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {

### 2017-10-21 r/header.R

@@ -2,8 +2,8 @@ rm(list = ls())
 graphics.off()
 
 # remove before final publication
-file.copy("~/Google Drive/CommonResources/refs.bib",
-          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+#file.copy("~/Google Drive/CommonResources/refs.bib",
+#          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {

### 2017-11-14 ms/ms.Rnw
finalizng ms

@@
 \\
+\textit{Twitter handle:} @thegoodphyte \\
+\\
 \textit{Author for correspondence:} \\
@@
 
-\linenumbers
-\setstretch{\stretchy}
-%\pagenumbering{gobble}
+%\linenumbers
+%\setstretch{\stretchy}
+\pagenumbering{gobble}
 %\raggedright
@@
 
-Data and annotated source code to generate this manuscript are available on GitHub (https://github.com/cdmuir/britstom) and Dryad \citep{Muir_dryad}.
+Data and annotated source code to generate this manuscript are available on GitHub (https://github.com/cdmuir/britstom) and processed data files are archived on Data Dryad (https://doi.org/10.5061/dryad.5q5q6).
 
@@
 \bibliography{refs}
-\bibliographystyle{evolution}
+\bibliographystyle{newphyt}
+
+\clearpage
+
+%--------------------------------------------------
+% Supporting Figure Captions
+%--------------------------------------------------
+
+\section*{Supporting figure captions}
+
+Figure S1: Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface
+
+Figure S2: Raunki\ae r life form and plant habit broadly overlap.
+
+Figure S3: Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+
+Figure S4: Growth forms have different tolerances for sun and shade among British angiosperms.
+
+Figure S5: The effect of light on stomatal ratio depends on growth form.
 
@@
 Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data.
-Points have been jittered for visual clarity.}
+Points have been jittered for visual clarity.
+Key to symbols: n.s.
+= not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.}
 \label{fig:SR-lf}
@@
 \section*{Supporting Information}
+
+\begin{center}
+{\Huge Light and growth form interact to shape stomatal ratio among British angiosperms}
+
+{\large Christopher D. Muir}
+
+{\large Accepted on 10 November 2017}
+
+\end{center}
+
+\clearpage
 
@@
 Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data.
-Points have been jittered for visual clarity.}
+Points have been jittered for visual clarity.
+Key to symbols: n.s.
+= not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.}
 \label{fig:SR-hf}

### 2017-11-14 r/03-plot-lifeform.R

@@ -20,7 +20,7 @@ lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyt
 names(lf) <- c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
 lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]
 
-pdf(str_c(path_figures, "/figureS_violin.pdf"), 8, 5.25)
+pdf(str_c(path_figures, "/figureS_violin.pdf"), 8, 5.25, useDingbats = FALSE)
 par(mai = c(1.75, 1.5, 0.25, 0.25), cex.lab = 1.5, las = 1, cex = 1.5, cex.axis = 0.5)
 
 vioplot(subset(stomata$sr_propAd, stomata$lifeform == names(lf[1]), na.rm = TRUE),

### 2017-11-14 r/header.R

@@ -2,8 +2,8 @@ rm(list = ls())
 graphics.off()
 
 # remove before final publication
-#file.copy("~/Google Drive/CommonResources/refs.bib",
-#          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
+file.copy("~/Google Drive/CommonResources/refs.bib",
+          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
 
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {

### 2017-11-15 ms/ms.Rnw
added citation to ms
small change to ms

@@
 The size, density, and distribution of stomata on a leaf vary widely and impact the flux of CO$_2$ and water vapour \citep[recently reviewed in][]{Sack_Buckley_2016}, as well as susceptibility to foliar pathogens that infect through stomata \citep{Mckown_etal_2014, Melotto_etal_2017}.
-Hence, stomata have been especially useful in understanding plastic and evolutionary response to climate change and domestication \citep{Woodward_1987, Beerling_Royer_2011, Milla_etal_2013}.
+Hence, stomata have been especially useful in understanding plastic and evolutionary responses to climate change and domestication \citep{Woodward_1987, Beerling_Royer_2011, Milla_etal_2013}.
 

### 2017-11-15 ms/ms.Rnw
final biorxiv version

@@
 \\
-\textit{Twitter handle:} @thegoodphyte \\
-\\
-\textit{Author for correspondence:} \\
-\textit{Christopher D. Muir} \\
-\textit{Tel:} +17782284851 \\
-\textit{Email:} chrisdmuir@gmail.com \\
-University of British Columbia \\
-6270 University Blvd. \\
-Vancouver, BC, Canada \\
-V6T 1Z4 \\
-\\
-Short title: Shedding light on stomatal evolution\\
-\\
-Word count: \\
-Summary: 199\\
-Introduction: 1355\\
-Materials and Methods: 1698\\
-Results: 582\\
-Discussion: 1935\\
-Acknowledgement: 26\\
-4 Figures and 1 Table, 5 Supplemental Figures
+\begin{center}
+In press at \textit{New Phytologist} \\
+\end{center}
+\clearpage
+%\textit{Twitter handle:} @thegoodphyte \\
+%\\
+%\textit{Author for correspondence:} \\
+%\textit{Christopher D. Muir} \\
+%\textit{Tel:} +1 984-218-5309 \\
+%\textit{Email:} chrisdmuir@gmail.com \\
+%1413 N Mangum St\\
+%Durham, NC 27701\\
+%\\
+%Short title: Shedding light on stomatal evolution\\
+%\\
+%Word count: \\
+%Summary: 199\\
+%Introduction: 1355\\
+%Materials and Methods: 1698\\
+%Results: 582\\
+%Discussion: 1935\\
+%Acknowledgement: 26\\
+%4 Figures and 1 Table, 5 Supplemental Figures
 
@@
 %\setstretch{\stretchy}
-\pagenumbering{gobble}
+%\pagenumbering{gobble}
 %\raggedright
@@
 
-Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio
+adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio
+
+\clearpage
 
@@
 
-% REMOVED: I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.
-
 I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$.
@@
 I compared the actual to stationary variance across simulated datasets using a paired $t$-test.
-
-% REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models).
-Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction.
-I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC.
-The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.
 
@@
 The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}.
-It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface.
+Additional CO$_2$ should increase quantum yield under high light when chloroplastic CO$_2$ concentration strongly limits photosynthetic rate.
+The association between light tolerance and amphistomy is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface.
 Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous.
@@
 In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another.
-For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll.
+For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll \citep{Flexas_etal_2012}.
 Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations.
@@
 
-\section*{Supporting figure captions}
+%\section*{Supporting figure captions}
 
-Figure S1: Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface
+%Figure S1: Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface
 
-Figure S2: Raunki\ae r life form and plant habit broadly overlap.
+%Figure S2: Raunki\ae r life form and plant habit broadly overlap.
 
-Figure S3: Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
+%Figure S3: Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.
 
-Figure S4: Growth forms have different tolerances for sun and shade among British angiosperms.
+%Figure S4: Growth forms have different tolerances for sun and shade among British angiosperms.
 
-Figure S5: The effect of light on stomatal ratio depends on growth form.
+%Figure S5: The effect of light on stomatal ratio depends on growth form.
 
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
-%\centerline{\includegraphics[angle=270]{figures/figure_SR-lf.eps}} % Needed this to upload file to journal website
+%\centerline{\includegraphics[angle=270]{figures_eps/figure_SR-lf.eps}} % Needed this to upload file to journal website
 \caption{The effect of light on stomatal ratio depends on Raunki\ae r life form.
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
-%\centerline{\includegraphics[angle=270]{figures/figure_SR-hf.eps}} % Needed this to upload file to journal website
+%\centerline{\includegraphics[angle=270]{figures_eps/figure_SR-hf.eps}} % Needed this to upload file to journal website
 \caption{The effect of light on stomatal ratio depends on growth form.

### 2017-11-25 r/07-plot-phylogeny.R
erased extraneous comments

@@ -107,14 +107,14 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
   
   r <- grconvertX(5.25, "in", "user")
   x <- cos(theta) * r
-  y <- sin(theta) * r #+ grconvertY(0.25, "ndc", "user")
+  y <- sin(theta) * r
   points(x, y, type = "l", col = "grey")
   text(cos(theta_axis) * r, sin(theta_axis) * r, 
        labels = "1", srt = 360 * (pi - theta_axis / (2 * pi)))
   
   r <- grconvertX(5, "in", "user")
   x <- cos(theta) * r
-  y <- sin(theta) * r #+ grconvertY(0.25, "ndc", "user")
+  y <- sin(theta) * r
   points(x, y, type = "l", col = "grey")
   text(cos(theta_axis) * r, sin(theta_axis) * r, 
        labels = "0.5", srt = 360 * (pi - theta_axis / (2 * pi)))
@@ -124,7 +124,7 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
 
   r <- grconvertX(4.75, "in", "user")
   x <- cos(theta) * r
-  y <- sin(theta) * r #+ grconvertY(0.25, "ndc", "user")
+  y <- sin(theta) * r
   points(x, y, type = "l", col = "grey")
   text(cos(theta_axis) * r, sin(theta_axis) * r,
        labels = "0", srt = 360 * (pi - theta_lab / (2 * pi)))
@@ -138,7 +138,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
     Y %<>% c(sin(seq(theta[i], theta[i + 1], length.out = 1e2)) * d[i])
   }
 
-  #Y %<>% add(grconvertY(0.25, "ndc", "user"))
   points(X, Y, type = "l", lwd = 3)
 
 dev.off()
@@ -246,7 +245,6 @@ gp <- plot(phy, type = "fan", show.tip.label = FALSE,
     Y %<>% c(sin(seq(theta[i], theta[i + 1], length.out = 1e2)) * d[i])
   }
   
-  #Y %<>% add(grconvertY(0.25, "ndc", "user"))
   points(X, Y, type = "l", lwd = 3)
   
 dev.off()

### 2017-11-25 r/09-analyse-stomatal-ratio.R

@@ -43,7 +43,6 @@ fitSR_lf_pl <- read_rds(str_c(path_objects, "/fitSR_lf_pl.rds"))
 # Figure
 
 pdf(str_c(path_figures, "/figure_SR-lf.pdf"), 4, 7, useDingbats = FALSE)
-#postscript(str_c(path_figures, "/figure_SR-lf.eps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {
@@ -106,25 +105,6 @@ mtext(2, text = expression(
 
 dev.off()
 
-# IN FLUX: compare slopes
-
-lf <- data.frame(lifeform = names(lf))
-lf %<>% 
-  tidyr::expand(lifeform, lifeform) %>%
-  remove_diag()
-lf %<>% mutate(mean = numeric(nrow(.)),
-               median = numeric(nrow(.)),
-               lower = numeric(nrow(.)),
-               upper = numeric(nrow(.)))
-
-for (i in 1:nrow(lf)) {
-  
-  s1 <- fitSR_lf_pl$bootstrap[, sprintf('lifeform%s', lf$lifeform[i])]
-  s2 <- fitSR_lf_pl$bootstrap[, sprintf('lifeform%s', lf$lifeform1[i])]
-  lf[i, 3:6] <- compare_slopes(s1, s2)
-  
-}
-
 ##### Habit and Ellenberg light indicator values versus sr_even -----
 
 hf <- c("tree", "shrub", "perennial", "biennial", "annual")
@@ -151,7 +131,6 @@ fitSR_hf_pl <- read_rds(str_c(path_objects, "/fitSR_hf_pl.rds"))
 # Figure
 
 pdf(str_c(path_figures, "/figure_SR-hf.pdf"), 4, 7, useDingbats = FALSE)
-#postscript(str_c(path_figures, "/figure_SR-hf.eps"), width = 4, height = 7) # need for submitting to journal website
 par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))
 
 for (i in 5:1) {

### 2017-11-25 r/header.R

@@ -1,10 +1,6 @@
 rm(list = ls())
 graphics.off()
 
-# remove before final publication
-file.copy("~/Google Drive/CommonResources/refs.bib",
-          "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)
-
 compile_pathd8 <- FALSE # changed to TRUE if you need to compile PATHd8
 if (compile_pathd8) {
   system("cd PATHd8; cc PATHd8.c -O3 -lm -o PATHd8")
